---
title: Tutorial on `model.tables()` Function in R
date: "1/6/2025"
date-modified: last-modified
date-format: long
---

# Tutorial on `model.tables()` Function in R

## Introduction

### Overview

The `model.tables()` function in the `stats` package generates summary tables for models fitted using `aov()`. It helps extract and interpret information such as effects or means for terms in a model. This function is particularly useful in analyzing complex ANOVA designs.

Key components:

- **Package**: `stats`
- **Function**: `model.tables()`

### Use Cases

- Summarizing model terms in ANOVA fits.
- Computing and displaying effects or means of factors.
- Adding standard errors for results, when required.

### Benefits

- Provides detailed summaries of complex ANOVA models.
- Facilitates interpretation of factor effects and mean responses.
- Seamless integration with other ANOVA tools in R.


## Getting Started

### Basic Usage

Here is a basic example using the `npk` dataset:

```{r}
data(npk)

# Fit an ANOVA model
npk.aov <- aov(yield ~ block + N * P * K, data = npk)

# Generate a table of means
model.tables(npk.aov, type = "means")
```

This generates a table summarizing mean responses for each term.

## Key Functions and Features

### Description

The `model.tables()` function computes summary tables of effects or means for terms in an `aov` model.

### Syntax

```r
model.tables(x, type = "effects", se = FALSE, cterms, ...)
```

#### Arguments

- **x**: An `aov` model object.
- **type**: Type of table to compute: `"effects"` (default) or `"means"`.
- **se**: Logical indicating whether standard errors should be included.
- **cterms**: A character vector specifying terms for which tables are computed (default: all terms).
- **…**: Additional arguments passed to methods.

### Example

Compute effects for all terms:

```{r}
model.tables(npk.aov, type = "effects")
```

## In-Depth Examples

### Title: Summarizing Mean Responses in a Three-Factor Design

#### Objective

Summarize the mean responses for combinations of factors in the `npk` dataset.

#### Steps

1. **Load the data**:

```{r}
data(npk)
```

2. **Fit the ANOVA model**:

```{r}
npk.aov <- aov(yield ~ block + N * P * K, data = npk)
```

3. **Generate mean tables**:

```{r}
model.tables(npk.aov, type = "means", se = TRUE)
```

4. **Interpret output**:

	- Each term's mean response is displayed.
	- Standard errors are included.

### Title: Analyzing Effects in an ANOVA Model

#### Objective

Examine factor effects in the `npk` dataset.

#### Steps

1. **Define the model**:

```{r}
npk.aov <- aov(yield ~ block + N + P + K, data = npk)
```

2. **Compute effects**:

```{r}
model.tables(npk.aov, type = "effects", se = TRUE)
```

3. **Output explanation**:

	- Lists coefficients for each term.
	- Includes standard errors, helping gauge variability.

## Advanced Features

### Customization

Focus on specific terms using `cterms`:

```{r}
model.tables(npk.aov, type = "means", cterms = c("N", "P"))
```

### Integration with Other Tools

Combine with `TukeyHSD()` for pairwise comparisons:

```{r}
TukeyHSD(npk.aov)
model.tables(npk.aov, type = "effects")
```

## Troubleshooting and FAQs

### Common Issues

1. **Error: "Weighted aov fits are not supported"**: Ensure the `aov` model is unweighted.
2. **Missing terms in output**: Use `cterms` to specify required terms explicitly.

### FAQs

- **What types of models are supported?** Only `aov` objects and certain `aovlist` objects.
- **Can it handle nested designs?** Yes, with appropriate formula specification.

## Best Practices

- Use `type = "effects"` to understand the contributions of each term.
- Always include `se = TRUE` for better interpretability.
- Validate model assumptions before interpreting tables.

## Conclusion

### Summary

The `model.tables()` function is a versatile tool for summarizing ANOVA model results, offering insights into factor effects and mean responses. Its ability to compute standard errors enhances its utility for interpreting results.

### Next Steps

- Explore `TukeyHSD()` for post-hoc comparisons.
- Combine with custom data preprocessing for more advanced analyses.

## References and Resources

- [R Documentation for `model.tables()`](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/model.tables.html)
- [ANOVA Techniques in R](https://www.statmethods.net/stats/anova.html)
- [Comprehensive Guide to R ANOVA Models](https://cran.r-project.org/doc/manuals/R-intro.pdf)