---
title: "Tutorial on `plot.design()` in R"
date: "1/6/2025"
date-modified: last-modified
date-format: long
---

# Comprehensive Tutorial on `plot.design()` in R

## Introduction

### Overview

The `plot.design()` function in R, part of the base `graphics` package, is used to visualize univariate effects of one or more factors in experimental designs. It is commonly applied in the context of `aov()` analyses and supports understanding of how factors affect a numeric response variable.

Key components:

- **Package**: `graphics`
- **Function**: `plot.design()`

### Use Cases

- Exploratory data analysis in experimental design.
- Comparing the effects of factors on response variables.
- Diagnosing potential outliers and interaction effects.

### Benefits

- Facilitates quick, visual understanding of main effects.
- Handles multiple factors simultaneously.
- Compatible with formulas and data frames for flexibility.

## Getting Started

### Basic Usage

The simplest use involves plotting the default dataset `warpbreaks`:

```{r}
plot.design(warpbreaks)
```

This creates a series of vertical lines where each factor's levels are plotted against a response variable.

## Key Functions and Features

### Description

`plot.design()` visualizes the effect of factors on a numeric response variable.

### Syntax

```r
plot.design(x, y = NULL, fun = mean, data = NULL, ...,
            ylim = NULL, xlab = "Factors", ylab = NULL,
            main = NULL, ask = NULL, xaxt = par("xaxt"),
            axes = TRUE, xtick = FALSE)
```

#### Arguments

- **x**: A data frame, formula, or terms object.
- **y**: The response variable (optional if included in `x`).
- **fun**: Function applied to subsets (default is `mean`).
- **data**: Data frame for variables referenced in `x` if formula-like.
- **ylim**: Range of y values.
- **xlab, ylab**: Axis labels.
- **main**: Main plot title.
- **axes, xtick**: Controls for axes and tick marks.

### Example

Basic plot for a formula and dataset:

```{r}
plot.design(breaks ~ wool + tension, data = warpbreaks)
```

This shows the effects of `wool` and `tension` on the response variable `breaks`.

## In-Depth Examples

### Title: Effects of Two Factors on Warp Breaks

#### Objective

Visualize and compare the effects of `wool` and `tension` on `breaks`.

#### Steps

1. **Prepare the data**:

```{r}
data("warpbreaks")
```

2. **Generate the plot**:

```{r}
plot.design(breaks ~ wool + tension, data = warpbreaks,
            main = "Effects of Wool and Tension on Breaks",
            col = c("red", "blue"))
```

#### Output

The plot displays vertical lines for each factor's levels and their impact on `breaks`, with overall mean as a horizontal reference.

## Advanced Features

### Customization

Adjust aesthetics and parameters:

```{r}
plot.design(breaks ~ wool + tension, data = warpbreaks,
            col = c("green", "purple"), pch = 16, ylim = c(20, 70),
            main = "Customized Plot")
```

### Integration with Other Packages

Combine `plot.design()` with `dplyr` for preprocessing:

```{r}
library(dplyr)
warpbreaks |> 
  filter(tension != "M") |> 
  plot.design(breaks ~ wool + tension, data = _,
              fun = median)
```

## Troubleshooting and FAQs

### Common Issues

1. **Error: "No numeric variables"**: Ensure the response variable is numeric.

```{r}
warpbreaks$breaks <- as.numeric(warpbreaks$breaks)
```

2. **Plot not displaying**: Use `dev.off()` to reset plotting device.

### FAQs

- **Can it handle interaction terms?** No, use `interaction.plot()` for interactions.

## Best Practices

- Always verify numeric response variables.
- Use `fun` parameter to explore summary statistics like `median`.
- Combine with diagnostic plots for a thorough analysis.

## Conclusion

### Summary

`plot.design()` is a versatile tool for visualizing factor effects in experimental data. It simplifies the analysis of variance by providing clear, interpretable plots.

### Next Steps

Explore:

- `interaction.plot()` for interaction effects.
- Advanced visualization packages like `ggplot2`.

## References and Resources

- [R Documentation for `plot.design`](https://www.rdocumentation.org/packages/graphics/versions/3.6.2/topics/plot.design)
- [Statistical Models in S](https://www.springer.com/gp/book/9780412830402)

