---
title: "Sub-setting Data in R"
date: "1/6/2025"
date-modified: last-modified
date-format: long
---
## Introduction

Sub-setting data in R is a fundamental operation for data manipulation and analysis. It involves selecting specific rows, columns, or elements of a data structure based on conditions or indices. This tutorial explores sub-setting techniques in R using **Base R** and **`tidyverse`** packages. It will provide comprehensive examples, tips, and nuances to master this essential skill.

Packages and Functions Covered:

- **Base R Functions**: `[]`, `subset()`
- **Tidyverse Functions**: `dplyr::filter()`, `dplyr::select()`, `dplyr::slice()`, `dplyr::pull()`

Use Cases:

- Extracting subsets of data for focused analysis.
- Filtering data for visualization or modeling.
- Reducing dataset size for computation efficiency.
- Handling missing or specific data points.

Benefits:

- **Base R**: Lightweight and does not require additional installations.
- **Tidyverse**: Provides clean, readable syntax and integrates well with pipelines (`|>`).

## Getting Started

Loading the Package:

```{r}
pacman::p_load(dplyr, tidyr) # Loads or installs the dplyr and tidyr packages
```

Basic Usage: Select rows where `mpg > 20` using both approaches:

**Base R**:

```{r}
data(mtcars)
mtcars[mtcars$mpg > 20, ]
```

**Tidyverse**:

```{r}
mtcars |>
    dplyr::filter(mpg > 20)
```

## Key Functions and Features

### Base R: Sub-setting with `[]`

**Description**:
The simplest and most versatile method for sub-setting in R. Use `[]` to extract rows, columns, or individual elements from data frames or matrices.

**Syntax**:

```r
data[rows, columns]
```

**Examples**:

1. Select rows where `mpg > 20`:

```{r}
mtcars[mtcars$mpg > 20, ]
```

2. Select specific columns:

```{r}
mtcars[, c("mpg", "cyl")]
```

3. Combine row and column conditions:

```{r}
mtcars[mtcars$mpg > 20, c("mpg", "cyl")]
```

**Nuances**:

- Omitting rows or columns selects all:

```{r}
mtcars[mtcars$mpg > 20, ] # All columns selected
mtcars[, c("mpg", "cyl")] # All rows selected
```

- Negative indexing excludes rows or columns:

```{r}
mtcars[-c(1, 2), ] # Exclude first two rows
```

### Base R: `subset()`

**Description**:
Provides a more human-readable syntax for sub-setting.

**Syntax**:

```r
subset(data, subset, select)
```

**Examples**:

1. Subset rows where `mpg > 20` and select `mpg` and `cyl` columns:

```{r}
subset(mtcars, mpg > 20, select = c(mpg, cyl))
```

2. Subset using multiple conditions:

```{r}
subset(mtcars, mpg > 20 & cyl == 6, select = c(mpg, cyl))
```

**Nuances**:

- Non-standard evaluation allows simpler syntax but can lead to ambiguities. Use column names explicitly when necessary.

### Tidyverse: `dplyr::filter()`

**Description**:
Filters rows based on logical conditions.

**Syntax**:

```r
dplyr::filter(data, condition)
```

**Examples**:

1. Filter rows with `mpg > 20`:

```{r}
mtcars |>
    dplyr::filter(mpg > 20)
```

2. Filter with multiple conditions:

```{r}
mtcars |>
    dplyr::filter(mpg > 20 & cyl == 6)
```

3. Use dynamic variables in filtering:

```{r}
threshold <- 20
mtcars |>
    dplyr::filter(mpg > threshold)
```

**Nuances**:

- Use `|` for "or" conditions:

```{r}
mtcars |>
    dplyr::filter(mpg > 20 | cyl == 6)
```

- Filter with `is.na()` to handle missing values:

```{r}
mtcars |>
    dplyr::filter(!is.na(mpg))
```

### Tidyverse: `dplyr::select()`

**Description**:
Selects specific columns from a data frame.

**Syntax**:

```r
dplyr::select(data, columns)
```

**Examples**:

1. Select specific columns:

```{r}
mtcars |>
    dplyr::select(mpg, cyl)
```

2. Use helpers for dynamic selection:

```{r}
mtcars |>
    dplyr::select(starts_with("d"))
```

**Nuances**:

- Combine helpers for advanced selection:

```{r}
mtcars |>
    dplyr::select(starts_with("d"), ends_with("t"))
```

### Tidyverse: `dplyr::slice()`

**Description**:
Selects rows by position.

**Syntax**:

```r
dplyr::slice(data, row_numbers)
```

**Examples**:

1. Select the first five rows:

```{r}
mtcars |>
    dplyr::slice(1:5)
```

2. Random sampling:

```{r}
mtcars |>
    dplyr::slice_sample(n = 5)
```

3. Select rows by percentages:

```{r}
mtcars |>
    dplyr::slice_sample(prop = 0.1) # 10% of rows
```

### Tidyverse: `dplyr::pull()`

**Description**:
Extracts a single column as a vector.

**Syntax**:

```r
dplyr::pull(data, column)
```

**Examples**:

1. Pull the `mpg` column:

```{r}
mtcars |>
    dplyr::pull(mpg)
```

2. Combine with filtering:

```{r}
mtcars |>
    dplyr::filter(cyl == 6) |>
    dplyr::pull(mpg)
```

**Nuances**:

- Useful for extracting data for external functions:

```{r}
mean(mtcars |> dplyr::pull(mpg))
```

## In-Depth Examples

### Example 1: Dynamic Filtering with External Variables

Objective: Filter rows based on user-defined thresholds for `mpg` and `hp`.

Steps:

1. Define thresholds:

```{r}
mpg_threshold <- 20
hp_threshold <- 100
```

2. Base R:

```{r}
mtcars[mtcars$mpg > mpg_threshold & mtcars$hp > hp_threshold, ]
```

3. Tidyverse:

```{r}
mtcars |>
    dplyr::filter(mpg > mpg_threshold & hp > hp_threshold)
```

## Advanced Features

Customization:

- Create reusable functions for sub-setting:

```{r}
filter_cars <- function(data, mpg_threshold, cyl_threshold) {
    data |>
        dplyr::filter(mpg > mpg_threshold, cyl == cyl_threshold)
}

filter_cars(mtcars, 20, 6)
```

Integration:

- Combine `tidyr` for reshaping with sub-setting:

```{r}
mtcars |>
    tidyr::pivot_longer(cols = everything()) |>
    dplyr::filter(value > 20)
```

## Troubleshooting and FAQs

- **Issue**: "undefined columns selected"
	Fix: Double-check column names.

- **Issue**: Sub-setting returns no rows.
	Fix: Ensure logical conditions are met.

FAQs:

- How to sub-set with regex?
	Use `grepl()` in Base R or `matches()` in `dplyr::select()`:

```{r}
mtcars |>
    dplyr::select(matches("^m"))
```

## Best Practices

- Preview data before sub-setting using `head()` or `glimpse()`.
- Chain operations for clarity in `tidyverse`.
- Avoid hard-coding indices; use dynamic variables.

## Conclusion

Recap: Sub-setting is a powerful tool for extracting specific data. Base R provides flexibility, while `tidyverse` offers an elegant syntax for modern workflows.
