# Common Errors in Mixed Models: Random Interaction Terms

This document explains a common error that occurs when trying to use random interaction terms in mixed models.

## The Error

When fitting a mixed model with random interaction terms using `lmer()`, you might encounter this error:

```r
Error: number of levels of each grouping factor must be < number of observations
(problems: FactorA:FactorB)
```

## What Causes This Error?

This error occurs in crossed random effects models when:

1. You have exactly one observation per cell (e.g., one observation per A:B combination)
2. The interaction term (A:B) has as many levels as there are observations
3. This creates a saturated model where the interaction perfectly predicts each observation

## Example That Produces the Error

```{r}
# Create a crossed design with 1 obs per cell
nA <- 5
nB <- 6
problem_data <- expand.grid(
    FactorA = factor(1:nA),
    FactorB = factor(1:nB)
)
problem_data$y <- rnorm(nrow(problem_data))

# This will fail with the error:
lmer(y ~ (1 | FactorA) + (1 | FactorB) + (1 | FactorA:FactorB), data = problem_data)
```

## Why This Is a Problem

When you have exactly one observation per cell:
- The A:B interaction term can perfectly predict every observation
- There's no residual variance left to estimate
- The model is overparameterized and cannot be estimated

## Solutions

1. **Multiple observations per cell**: Ensure you have at least 2 observations for each Factor A × Factor B combination.

```{r}
# Repeat each A:B combination at least twice
good_data <- problem_data[rep(1:nrow(problem_data), each = 2), ]
good_data$y <- rnorm(nrow(good_data))

# This will work:
lmer(y ~ (1|FactorA) + (1|FactorB) + (1|FactorA:FactorB), data = good_data)
```

2. **Incomplete design**: Use a design where not every A:B combination exists, so the number of cells is less than the number of observations.

3. **Omit the interaction**: If you truly have only one observation per cell, you cannot include the interaction term as random. Use only the main random effects:

```{r}
# This will work (but doesn't model cell-specific variation):
lmer(y ~ (1|FactorA) + (1|FactorB), data = problem_data)
```

4. **Use a different correlation structure**: Consider other ways to model the relationship, such as random slopes instead of interaction terms.

## Understanding Random Interaction Terms

- A random interaction term (1|A:B) models excess variation specific to combinations of factors
- It captures non-additive effects where some combinations deviate from what would be predicted by adding the main effects
- In experimental design, this is sometimes called the "cell-specific" component of variance

## Technical Requirements

For a model with p random effects terms, the data must satisfy:
1. For each random effects term, the number of levels must be less than n (total observations)
2. The total number of observations must exceed the sum of ranks of the random effects design matrices

## Application in Research

Random interaction terms are important in many research contexts:
- Rater × item designs in psychometrics
- Stimulus × subject designs in psychology
- Genotype × environment studies in agriculture
- Many other crossed experimental designs

By properly designing studies with sufficient replication within cells, you can properly estimate these important variance components.