---
title: "Week 5: Confounding and Identification"
---

## R Packages

```{r}
#| label: load-packages
pacman::p_load(
    "dagitty",
    "ggplot2",
    "ggdag",
    "tidyverse",
    "ggpubr"
)

ggplot2::theme_set(ggpubr::theme_pubr())
```

## Activate Prior Knowledge & Set Context

Imagine a study investigating whether a new drug, let's call it "Drug X," reduces the risk of heart attack. We compare the heart attack rates between patients who took Drug X and those who didn't. The analysis shows that patients who took Drug X actually had a *higher* rate of heart attacks. Should we conclude that Drug X is harmful?

Not so fast. What if doctors were more likely to prescribe Drug X to patients who were already at a higher risk of heart disease—perhaps because they had a history of high blood pressure or diabetes? In this case, the apparent association between Drug X and heart attacks is not a causal effect, but is instead **confounded** by the underlying patient health status.

This week, we'll build on our understanding of potential outcomes from Week 2 and introduce the formal tools to address this challenge. We'll connect the potential outcomes framework to a graphical language—Directed Acyclic Graphs (DAGs)—that helps us visualize these confounding structures and provides a clear, systematic way to decide which variables we need to adjust for to get an unbiased causal estimate. The goal is to move from a biased **association** to a valid **causal effect** by achieving **conditional exchangeability**.

## The Link Between Potential Outcomes and DAGs

The **Potential Outcomes Framework** and **DAGs** are two different but complementary languages for thinking about causal inference.

- **Potential Outcomes**: This framework is focused on the *outcomes* that would have occurred under different treatment assignments. The core concept is **exchangeability**, which states that the distribution of potential outcomes is the same for the treated and untreated groups. In observational studies, we aim for **conditional exchangeability**, $Y^{a} \indep A \mid X$, by adjusting for a set of covariates $X$. The challenge is knowing which $X$ to choose.
- **Directed Acyclic Graphs (DAGs)**: This framework provides a visual representation of our causal assumptions. The arrows represent our beliefs about causal relationships. The key insight is that **confounding** arises from specific graphical structures, namely **backdoor paths**. The DAG allows us to systematically identify which variables ($X$) need to be adjusted for to "block" these paths and satisfy the conditional exchangeability assumption.

In essence, the DAG helps us *identify* the set of variables that we must condition on to make the potential outcomes framework's core assumption—conditional exchangeability—plausible.

## Exchangeability and the Backdoor Criterion

In Week 2, we learned that a causal effect for a binary treatment $A$ is the difference in potential outcomes, $\E{Y^1} - \E{Y^0}$. The fundamental problem of causal inference is that we can never observe both potential outcomes for the same individual. Our goal is to use observed data to estimate these unobserved potential outcome means.

This is only possible if the treatment groups are **exchangeable**. That is, the group that received the treatment ($A=1$) and the group that did not ($A=0$) are, on average, identical in their distribution of potential outcomes.

- **Unconditional Exchangeability**: This is the "gold standard" and is achieved in a properly conducted **Randomized Controlled Trial (RCT)**. By randomly assigning treatment, we ensure that the groups are balanced on all baseline characteristics, both measured and unmeasured. This means the potential outcomes are independent of the treatment assignment:

$$
Y^a \indep A \quad \text{for } a=0,1
$$

- **Conditional Exchangeability** (also known as **Ignorability** or **No Unmeasured Confounding**): In observational studies, we can't randomize. Instead, we assume that the treatment groups are exchangeable *within strata* of observed covariates $X$. If we adjust for all relevant confounding variables $X$, then within each stratum defined by $X$, the treatment assignment is "as good as random." This can be written as:

$$
Y^a \indep A \mid X \quad \text{for } a=0,1
$$

The **backdoor criterion** is a graphical rule for determining which variables to adjust for to achieve conditional exchangeability. It provides a formal, visual method to identify a sufficient set of variables to block all sources of confounding.

A **backdoor path** from the treatment ($A$) to the outcome ($Y$) is any path that begins with an arrow pointing *into* the treatment node. These paths represent spurious associations that are not due to the causal effect of $A$ on $Y$.

A set of covariates $X$ is said to satisfy the **backdoor criterion** if:

1. **It blocks all backdoor paths** from $A$ to $Y$.
2. **It does not contain any colliders** that are on a path from $A$ to $Y$ and have descendants that are also colliders.

If we adjust for a set of variables $X$ that satisfies the backdoor criterion, we have effectively "closed" all the confounding pathways, and the remaining association between $A$ and $Y$ can be causally interpreted.

## Adjustment Sets

- **Sufficient Adjustment Set**: Any set of variables $X$ that satisfies the backdoor criterion. Adjusting for a sufficient set will produce an unbiased estimate of the causal effect.
- **Minimal Sufficient Adjustment Set**: A sufficient set that does not contain any redundant variables. In other words, if you remove any variable from a minimal sufficient set, the remaining set is no longer sufficient. It's often desirable to use a minimal sufficient set to improve statistical efficiency.

## Residual Confounding

Even when we are careful, we can still have **residual confounding**. This bias remains when:

- A confounder is **unmeasured** and thus cannot be included in our adjustment set.
- A confounder is measured with **error**, so our adjustment is imperfect.
- The confounder is **time-varying** and the causal structure is more complex than a simple cross-sectional DAG. We will discuss this more in later weeks.

## Intuitive Understanding

At its core, **residual confounding** is the bias that remains after you've tried to control for confounding variables. Think of it as the confounding you *couldn't* get rid of. This happens for a few key reasons, which you've outlined perfectly.

Let's use an analogy. Imagine you're trying to measure the true height of a person who is wearing a hat and standing on a slanted floor. The hat and the slant are confounders for your measurement.

- **Controlling for confounding** is like asking the person to take off the hat (adjusting for a measured confounder) and using a level to account for the floor's slant.
- **Residual confounding** is the error that remains because:
    1. You didn't realize they are wearing thick-soled shoes (**unmeasured confounder**).
    2. Your level isn't perfectly calibrated, so your adjustment for the slant is slightly off (**measurement error**).

### 1. Unmeasured Confounders

This is the most common and intuitive source of residual confounding. It occurs when there is a variable that meets the criteria for a confounder, but it was not measured in your dataset. Therefore, you cannot include it in your statistical model to adjust for its effect.

- **How it Works**: An unmeasured variable, let's call it $U$, creates a "backdoor path" between your treatment ($A$) and outcome ($Y$). Since you can't measure $U$, you cannot condition on it to block this path. The association you estimate between $A$ and $Y$ is a mix of the true causal effect and the spurious association created by $U$.
- **Example**: Suppose we are studying the effect of regular exercise ($A$) on longevity ($Y$). We adjust for age and socioeconomic status ($X$). However, we haven't measured "health consciousness" ($U$). People who are more health-conscious are more likely to exercise ($U \rightarrow A$) and are also more likely to engage in other healthy behaviors (like better diet, not smoking) that increase longevity ($U \rightarrow Y$). Because we didn't adjust for health consciousness, our estimate of the effect of exercise will be biased upwards, attributing longevity benefits to exercise that are actually due to this underlying trait.

### 2. Confounders Measured with Error

This scenario is more subtle. You correctly identify a confounder, you measure it, and you adjust for it. However, your measurement is imperfect.

- **How it Works**: When you adjust for a variable measured with error, you are only *partially* closing the backdoor path it creates. The adjustment is incomplete because the variable you're putting in the model isn't the true confounder; it's a noisy proxy for it. Some of the confounding influence "leaks" past your imperfect adjustment.
- **Example**: Let's study the effect of a specific air pollutant ($A$) on childhood asthma ($Y$). We know that family income ($X$) is a confounder, as it influences both neighborhood location (and thus pollution exposure) and access to healthcare. However, we measure income ($X^*$) by asking parents to self-report it, which can be inaccurate due to recall issues or social desirability bias. This self-reported income, $X^*$, is a proxy for true income, $X$. When we adjust for $X^*$, we reduce the confounding from income, but we don't eliminate it. The difference between the effect of *true* income and the effect of *reported* income remains as residual confounding.

## Mathematical Foundations

Let's formalize how these two scenarios introduce bias. Our goal is to estimate the Average Treatment Effect (ATE), $\E{Y^1 - Y^0}$. In an ideal observational study, we assume conditional exchangeability, $Y^a \indep A \mid X$, which allows us to identify the ATE by adjusting for a sufficient set of confounders, $X$. The problem is when our set of covariates is not sufficient.

## Bias from an Unmeasured Confounder

Let's assume the true data-generating process for an outcome $Y$ is a linear model:

$$
Y = \beta_0 + \beta_1 A + \beta_2 X + \beta_3 U + \varepsilon
$$

Here, $A$ is the treatment, $X$ is a *measured* confounder, and $U$ is an *unmeasured* confounder. By definition, $U$ is correlated with both $A$ and $Y$. Because we cannot observe $U$, the model we actually estimate is:

$$
Y = \alpha_0 + \alpha_1 A + \alpha_2 X + \varepsilon^*
$$

We want to know if our estimate, $\hat{\alpha}_1$, is a good estimate of the true causal effect, $\beta_1$. According to the theory of omitted variable bias (a foundational concept in econometrics), the expected value of our estimator $\hat{\alpha}_1$ is:

$$
\E{\hat{\alpha}_1} = \beta_1 + \beta_3 \cdot \delta_{UA}
$$

- **Derivation**: This classic result (see Wooldridge, 2010, *Econometric Analysis of Cross Section and Panel Data*, Chapter 3) arises from the fact that the OLS estimator for $\alpha_1$ is trying to account for the variation in $Y$ associated with $A$ after partialling out $X$. Since $U$ is missing, its effect gets improperly attributed to $A$ and $X$. The term $\delta_{UA}$ represents the association between the unmeasured confounder $U$ and the treatment $A$ after adjusting for $X$. It comes from an auxiliary regression of the omitted variable on the included ones: $U = \delta_0 + \delta_{UA}A + \delta_{UX}X + \text{residual}$.
- **Interpretation**: The bias, $\beta_3 \cdot \delta_{UA}$, is the **residual confounding**. It is the product of two quantities:
  1. The effect of the unmeasured confounder on the outcome ($\beta_3$).
  2. The partial correlation between the treatment and the unmeasured confounder, conditional on the measured confounders ($\delta_{UA}$).
  For residual confounding to exist, both terms must be non-zero, which is true by the definition of $U$ being a confounder.

## Bias from Measurement Error in a Confounder

This is slightly more complex. Let the true model be:

$$
Y = \beta_0 + \beta_1 A + \beta_2 X + \varepsilon
$$

where $X$ is the true, unobserved confounder. Instead, we observe $X^*$, which is a noisy version of $X$. Under the classical measurement error model, we assume:

$$
X^* = X + e
$$

where the error $e$ is independent of $X$, $A$, and $\varepsilon$, with $\E{e}=0$. We run the regression using the proxy:

$$
Y = \alpha_0 + \alpha_1 A + \alpha_2 X^* + \varepsilon^*
$$

The key issue here is that by controlling for $X^*$, we do not make $A$ and $Y^a$ independent. That is, $Y^a \nindep A \mid X^*$ even if $Y^a \indep A \mid X$. The adjustment is incomplete.

- **Derivation**: We can substitute $X = X^* - e$ into the true model:

$$
Y = \beta_0 + \beta_1 A + \beta_2 (X^* - e) + \varepsilon = \beta_0 + \beta_1 A + \beta_2 X^* + (\varepsilon - \beta_2 e)
$$

  The new error term is $\varepsilon' = \varepsilon - \beta_2 e$. In this regression, the regressor $X^*$ is correlated with the error term $\varepsilon'$ because both depend on $e$. This violates a core assumption of OLS (that regressors and errors are uncorrelated) and leads to biased estimates for *all* coefficients, including $\alpha_1$.

  The resulting bias in $\hat{\alpha}_1$ is complex, but its direction and magnitude depend on the correlations between $A$, $X$, and $Y$, and the amount of noise in $X^*$. As shown in many texts (e.g., VanderWeele, 2015, *Explanation in Causal Inference*, Chapter 9), the bias in $\hat{\alpha}_1$ will generally be an attenuation of the confounding bias. You've removed some of the bias due to $X$, but not all of it.
- **Interpretation**: The difference between the estimate $\hat{\alpha}_1$ and the true effect $\beta_1$ is the **residual confounding** due to measurement error. The less reliable the measurement of the confounder (i.e., the larger the variance of the error $e$ relative to the variance of the true confounder $X$), the greater the residual confounding. If the measurement is pure noise, adjusting for $X^*$ does nothing, and the confounding is the same as if you hadn't adjusted for it at all.

This topic highlights a fundamental challenge in observational studies. Since we can rarely rule out residual confounding, it's often crucial to assess how our conclusions might change if such confounding were present.

## Identification Process

A visual representation of the identification process:

```tikz
\begin{tikzpicture}[
    node distance=2cm,
    block/.style={rectangle, draw, text width=5cm, text centered, rounded corners, minimum height=1.5cm},
    decision/.style={diamond, draw, text width=4cm, text centered, rounded corners, minimum height=1.5cm, aspect=2},
    line/.style={draw, -latex'}
]
\node (A) [block] {Causal Question};
\node (B) [decision, right=2cm of A] {Assume Causal Structure (DAG)};
\node (C) [block, below=1cm of B] {Identify All Backdoor Paths from $A$ to $Y$};
\node (D) [block, below=1cm of C] {Select Adjustment Set $X$ to block paths};
\node (E) [decision, below=1cm of D] {Check if $X$ satisfies Backdoor Criterion};
\node (F) [block, below=1cm of E] {Estimate Causal Effect};
\node (G) [block, left=2cm of E] {Re-evaluate DAG or Adjustment Set};

\path [line] (A) -- (B);
\path [line] (B) -- (C);
\path [line] (C) -- (D);
\path [line] (D) -- (E);
\path [line] (E.south) -- node [right, xshift=1cm] {Yes} (F.north);
\path [line, dashed] (E.west) -- node [left, yshift=-1.5cm] {No} (G.east);
\path [line] (G) |- (B.west);
\end{tikzpicture}
```

## Illustrate of Concepts with Examples

Let's consider a study on the effect of a new blood pressure medication ($A$) on heart disease mortality ($Y$). We suspect that age ($U$) is a confounder.

The causal structure is: $A \leftarrow U \rightarrow Y$.

We collect data on 100 individuals and find the following crude proportions of mortality:

- Among those who took the medication ($A=1$): 50% mortality
- Among those who did not ($A=0$): 25% mortality

This crude association suggests the medication is harmful. However, we suspect confounding by age. So, we stratify the data by age group (Young, Old).

| Group             | Treatment ($A$) | Mortality ($Y$) | Total |
| ----------------- | --------------- | --------------- | ----- |
| **Young ($U=0$)** | $A=1$           | 10              | 40    |
|                   | $A=0$           | 5               | 20    |
| **Old ($U=1$)**   | $A=1$           | 40              | 60    |
|                   | $A=0$           | 20              | 80    |

: Mortality by Age and Treatment {#tbl-mortality-age-treatment}

Let's calculate the conditional effects within each stratum:

- **Young Group ($U=0$)**:
    - Mortality with treatment: $10/40 = 25\%$
    - Mortality without treatment: $5/20 = 25\%$
    - Causal effect in young: $25\% - 25\% = 0\%$
- **Old Group ($U=1$)**:
    - Mortality with treatment: $40/60 = 66.7\%$
    - Mortality without treatment: $20/80 = 25\%$
    - Causal effect in old: $66.7\% - 25\% = 41.7\%$

The crude association was heavily biased. We see a large causal effect in the old group, but no effect in the young group. To get an overall estimate, we need to standardize or pool these results. For example, using standardization to estimate the ATE:

$$
\E{Y^1} = \sum_z \mathbb{P}(Y=1 \mid A=1, U=u) \mathbb{P}(U=u)
$$
$$
\E{Y^0} = \sum_z \mathbb{P}(Y=1 \mid A=0, U=u) \mathbb{P}(U=u)
$$

Total young: $40+20 = 60$. Total old: $60+80=140$.

The marginal distribution of $U$ is $\mathbb{P}(U=0) = 60/200 = 0.3$ and $\mathbb{P}(U=1) = 140/200 = 0.7$.

$$
\E{Y^1} = (0.25 \times 0.3) + (0.667 \times 0.7) \approx 0.075 + 0.467 = 0.542
$$

$$
\E{Y^0} = (0.25 \times 0.3) + (0.25 \times 0.7) = 0.25
$$

The ATE is $\E{Y^1} - \E{Y^0} \approx 0.542 - 0.25 = 0.292$. The treatment causes, on average, a 29.2% increase in mortality, which is a very different conclusion from the crude 25% increase. The adjustment for age was critical.

## Apply with Code

Let's simulate this scenario in R and use regression to adjust for the confounder. We'll use the `dagitty` package to visualize the DAG and confirm our adjustment set.

```{r}
#| label: fig-simulate-data
#| tbl-cap: "Simulated Data"
# 1. Visualize the DAG and Identify the Adjustment Set
# Causal Structure: A <- U -> Y
dag <- dagitty(
    "dag {
    A [exposure];
    Y [outcome];
    U [confounder];
    U -> A;
    U -> Y;
    A -> Y;
}"
)
# plot(dag)
ggdag::ggdag(dag) + ggtitle("DAG: A <- U -> Y")
```

```{r}
# Find a Sufficient Adjustment Set Using the Backdoor Criterion
# The Backdoor Path is A <- U -> Y.
# We Need to Block This Path. U is the only Non-collider on This Path.
# So, U is a Minimal Sufficient Adjustment Set.
adjustment_sets <- adjustmentSets(
    dag,
    exposure = "A",
    outcome = "Y",
    effect = "total"
)
print(adjustment_sets)
# The Output Should Confirm that { U } is a Valid Adjustment Set.

# 2. Simulate Data Based on This Causal Structure
set.seed(579)
n <- 200
U <- rbinom(n, 1, 0.7) # Simulate Age (0=Young, 1=Old)
A <- rbinom(n, 1, prob = case_when(U == 0 ~ 0.4, U == 1 ~ 0.6)) # Treatment depends on age
Y <- rbinom(
    n,
    1,
    prob = case_when(
        A == 0 & U == 0 ~ 0.25,
        A == 1 & U == 0 ~ 0.25,
        A == 0 & U == 1 ~ 0.25,
        A == 1 & U == 1 ~ 0.667
    )
) # Outcome depends on A and U
data <- data.frame(A, Y, U)

# 3. Perform the Crude (unadjusted) Analysis
crude_model <- glm(Y ~ A, data = data, family = binomial(link = "logit"))
summary(crude_model)
# The Coefficient for A Will Be Biased, Reflecting the Confounded Association.
# The `exp(coef(crude_model)["A"])` Will Be the Crude Odds Ratio.

# 4. Perform the Adjusted Analysis by including the Confounder U
adjusted_model <- glm(Y ~ A + U, data = data, family = binomial(link = "logit"))
summary(adjusted_model)
# The Coefficient for A in This Model is an Estimate of the Conditional Causal Effect.
# `exp(coef(adjusted_model)["A"])` Will Be the Conditional Odds Ratio.
# Note: The Coefficients for A and U here Represent the Conditional Log-odds Ratios.
# To Get the Marginal ATE, We Would Need to Use Standardization Methods like G-computation,
# Which We Will Cover in a Future Lecture. For Now, the Key is that This Regression
# Model Successfully Blocks the Backdoor Path and Provides an Unbiased Estimate
# Of the *conditional* Causal Effect.
```

In the code above, the `dagitty` output confirms that we must adjust for `U` to get an unbiased estimate. The crude model, which ignores `U`, will give a biased coefficient for `A`. The adjusted model, by including `U`, correctly isolates the causal effect of `A` from the confounding influence of `U`.

## Promote Retrieval & Metacognition

## Knowledge Check

1. **Question**: In the context of causal inference, what is the key difference between unconditional and conditional exchangeability?
  - **Answer**: **Unconditional exchangeability** means the treatment groups are balanced on all confounders, both measured and unmeasured. This is the ideal state achieved through proper randomization in an RCT. **Conditional exchangeability** is an assumption in observational studies that the treatment groups are exchangeable *within strata* of a specific set of measured covariates. We achieve this by adjusting for those covariates.
2. **Question**: Describe a backdoor path from an exposure to an outcome using a simple DAG. What is the purpose of adjusting for a variable on this path?
  - **Answer**: A backdoor path is a non-causal path from the exposure ($A$) to the outcome ($Y$) that begins with an arrow pointing *into* $A$. For example, in the DAG $A \leftarrow U \rightarrow Y$, the path is $A \leftarrow U \rightarrow Y$. Adjusting for a variable like $U$ on this path "blocks" or "closes" it, removing the confounding association and isolating the causal effect of $A$ on $Y$.
3. **Question**: Consider a DAG with nodes $A \rightarrow M \rightarrow Y$ and $A \leftarrow C \rightarrow Y$. What are the minimal sufficient adjustment sets for the causal effect of $A$ on $Y$? What if we also had the path $A \leftarrow C \rightarrow M \rightarrow Y$?
  - **Answer**: For the first DAG ($A \leftarrow C \rightarrow Y$ and $A \rightarrow M \rightarrow Y$), the only backdoor path is $A \leftarrow C \rightarrow Y$. $M$ is a mediator and should not be adjusted for. Thus, the minimal sufficient adjustment set is just ${C}$. In the second scenario, $A \leftarrow C \rightarrow M \rightarrow Y$, there is a second backdoor path, $A \leftarrow C \rightarrow M \rightarrow Y$. Adjusting for $C$ blocks both of these, so ${C}$ remains the minimal sufficient adjustment set.
4. **Question**: You conduct a study on the effect of a new medication on patient recovery time. You adjust for age and gender, but you know that patient genetics are a strong predictor of both medication response and recovery time. If you do not have data on genetics, what type of bias will your results suffer from?
  - **Answer**: Your results will suffer from **residual confounding** due to the unmeasured variable, genetics. Even though you adjusted for other known confounders (age and gender), the bias from the unmeasured genetic confounder will remain in your effect estimate.

## Reflective Prompt

What was the most confusing part of the relationship between DAGs and the potential outcomes framework this week? How might you explain the backdoor criterion to a classmate to solidify your understanding? For instance, what analogy could you use to explain a "backdoor path" and how "blocking" it with an adjustment set works?