---
title: "Week 3: Confounding & Identification"
---

## 🧭 Introduction: The Problem of Confounding

Welcome to Week 3 of STAT 579! This week, we will tackle the central challenge of causal inference in observational studies: **confounding**. The goal is to move from simply observing an association to confidently making a causal claim. We'll build on our understanding of potential outcomes from Week 2 and introduce the formal tools to address this challenge. We'll connect the potential outcomes framework to a graphical language—Directed Acyclic Graphs (DAGs)—that helps us visualize these confounding structures and provides a clear, systematic way to decide which variables we need to adjust for to get an unbiased causal estimate. The ultimate goal is to achieve **conditional exchangeability**.

## Confounding: A Detailed Explanation

Confounding is a specific type of bias that arises when a third variable, the **confounder**, is a common cause of both the treatment and the outcome. This creates a non-causal association between the treatment and the outcome that can be mistaken for a true causal effect.

::: callout-tip
## 🧠 Learning Tip: Two Frameworks, One Problem

The concept of confounding can be understood through two lenses:

1.  **Potential Outcomes Framework**: Confounding exists when the treatment assignment ($A$) and the potential outcomes ($Y^a$) are not independent. That is, $\nindep{A}{Y^a}$. This means that the treated and untreated groups are fundamentally different with respect to their outcomes, even if no one received the treatment. This difference is due to the confounder's influence.
2.  **Graphical Framework (DAGs)**: Confounding is visually represented by a **backdoor path**. A backdoor path is a sequence of arrows that connects the treatment ($A$) to the outcome ($Y$) and begins with an arrow pointing into the treatment node. The variable $Z$ is a confounder.
:::

## The Graphical Perspective: The Backdoor Path

Let's consider a classic confounding scenario where age ($Z$) confounds the relationship between a medication ($A$) and an outcome ($Y$).

``` tikz
\begin{tikzpicture}[
   node distance=2cm,
   block/.style={rectangle, draw, text width=2cm, text centered, rounded corners},
   line/.style={draw, -latex'}
]
   % Nodes
   \node (Z) [block] {Confounder (Z)};
   \node (A) [block, below left=1.5cm of Z] {Treatment (A)};
   \node (Y) [block, below right=1.5cm of Z] {Outcome (Y)};
   
   % Arrows
   \path [line] (A) -- (Y);
   \path [line] (Z) -- (A);
   \path [line] (Z) -- (Y);
   
   % Labels
   \node [above, align=center] at ($(A)!0.5!(Y) + (0,1.5)$) {Backdoor Path};
   \node [above, align=center] at ($(A)!0.5!(Y) + (0,0.5)$) {Causal Effect};
\end{tikzpicture}
```

In this DAG, the direct causal effect is represented by the solid arrow $A \to Y$. However, there is also a non-causal "backdoor" path: $A \leftarrow Z \rightarrow Y$. This path generates a spurious association because a change in $A$ is correlated with a change in $Z$, which in turn causes a change in $Y$. Our goal is to **block** this backdoor path.

## Exchangeability

In the potential outcomes framework, we can write the following statement: 
$$
A \indep Y^a
$$

where $Y^a$ is the potential outcome. It represents the outcome ($Y$) that would have been observed for an individual if their treatment level ($A$) had been set to a specific value $a$. $A$ is the actual observed treatment that an individual actually received. The notation $\indep$ means "is statistically independent of".

The entire statement $Y^a \indep A$ means that the potential outcome under a fixed treatment $a$ is independent of the treatment that was actually assigned. In other words, the treatment assignment process is not related to what the outcome would have been under that treatment.

### Which one is fixed?

In the term $Y^a$, the superscript $a$ is the fixed value. It represents a specific, hypothetical level of treatment. $A$ (without a superscript) is a random variable representing the observed treatment group.

In simpler terms: Imagine $a=1$ (treated) and $A$ can be 1 (treated) or 0 (not treated). The formula $Y^1 \indep A$ assumes that the outcome we would see if everyone were treated ($Y^1$) is independent of who actually got the treatment ($A$). This means the treated and untreated groups are comparable (exchangeable) with respect to this potential outcome.

## The Goal: Achieving Exchangeability

To estimate a causal effect, we need the treatment and control groups to be **exchangeable**. This means that the distribution of potential outcomes is the same for the treated and untreated groups. This assumption is also called **no unmeasured confounding**.

-   **Unconditional Exchangeability**: In an RCT, randomization ensures that treatment assignment ($A$) is independent of the potential outcomes ($Y^a$). We can write this formally as:

$$
Y^a \indep A
$$

**In plain English:** "The potential outcomes if a person received treatment $a$ are statistically independent of whether or not they actually received treatment." This formula means that the distribution of potential outcomes is the same for the treated and untreated groups. The observed difference in outcomes is a direct reflection of the causal effect.

-   **Conditional Exchangeability**: In observational studies, we must assume that the treatment and control groups are exchangeable *within strata* of a set of measured covariates, $\vect{Z}$. This is the assumption of no unmeasured confounding. We write this as:

$$
Y^a \indep A \mid \vect{Z}
$$

**In plain English:** "The potential outcomes if a person received treatment $a$ are statistically independent of whether or not they actually received treatment, *once we account for the variables in Z*." This formula means that once we "condition on" or "adjust for" the confounders in $\vect{Z}$, the treatment assignment is effectively random. The average potential outcome is the same for the treated and untreated within each level of $\vect{Z}$.

## The Backdoor Criterion: The Graphical Rule for Identification

The **backdoor criterion** is a formal rule for choosing a sufficient set of variables to adjust for. If we adjust for a set of variables $\vect{Z}$ that satisfies this criterion, we have effectively blocked all confounding paths and can assume conditional exchangeability.

## The Criterion Stated Formally

A set of variables $\vect{Z}$ satisfies the backdoor criterion relative to an exposure $A$ and an outcome $Y$ if:

1.  **No node in** $\vect{Z}$ is a descendant of $A$. This prevents us from adjusting for mediators, which would block the causal effect.
2.  $\vect{Z}$ **blocks every path** between $A$ and $Y$ that has an arrow pointing into $A$.

A path is blocked if it contains either:

-   A **non-collider** that is in $\vect{Z}$.
-   A **collider** that is *not* in $\vect{Z}$ and has no descendants in $\vect{Z}$.

To block the backdoor path $A \leftarrow Z \rightarrow Y$, we must adjust for $Z$. The variable $Z$ is a non-collider on this path. By including $Z$ in our analysis, we satisfy the backdoor criterion and can obtain an unbiased estimate of the causal effect.

Let's illustrate with a slightly more complex DAG where $Z_1$ and $Z_2$ are confounders and $Z_3$ is a collider.

``` tikz
\begin{tikzpicture}[
    node distance=2.5cm,
    block/.style={rectangle, draw, text width=2cm, text centered, rounded corners, minimum height=1cm},
    line/.style={draw, -latex'}
]

\node (Z1) [block] {Z1};
\node (Z2) [block, right=2cm of Z1] {Z2};
\node (A) [block, below left=1.5cm of Z1] {Treatment (A)};
\node (Y) [block, below right=1.5cm of Z2] {Outcome (Y)};
\node (Z3) [block, below=1cm of Z2] {Z3};

% Causal arrows
\path [line] (A) -- (Y);
\path [line] (Z1) -- (A);
\path [line] (Z1) -- (Y);
\path [line] (Z2) -- (Y);

% Collider path
\path [line] (A) -- (Z3);
\path [line] (Y) -- (Z3);

\end{tikzpicture}
```

In this DAG, the paths from A to Y are:

-   $A \to Y$ (the causal effect we want to estimate)
-   $A \leftarrow Z_1 \to Y$ (backdoor path)
-   $A \leftarrow Z_2 \to Y$ (backdoor path)
-   $A \to Z_3 \leftarrow Y$ (collider path)

To satisfy the backdoor criterion, we must block all backdoor paths. We do this by adjusting for $Z_1$ and $Z_2$. The path through the collider, $A \to Z_3 \leftarrow Y$, is already blocked because we are **not** adjusting for $Z_3$. The sufficient adjustment set is ${Z_1, Z_2}$.

## Illustrate with a Concrete Example

We are studying the effect of a new blood pressure medication ($A$) on heart disease mortality ($Y$). We suspect that `age` ($Z$) is a confounder. The DAG is: $A \leftarrow Z \rightarrow Y$. The backdoor path is $A \leftarrow Z \rightarrow Y$. The minimal sufficient adjustment set is ${Z}$.

## Step-by-Step Calculation

We have the following data from 200 patients:

|       Group       | Treatment ($A$) | Mortality ($Y$) | Total |
| :---------------: | :-------------: | :-------------: | :---: |
| **Young (**$Z=0$) |      $A=1$      |       10        |  40   |
|                   |      $A=0$      |        5        |  20   |
|  **Old (**$Z=1$)  |      $A=1$      |       40        |  60   |
|                   |      $A=0$      |       20        |  80   |

1.  **Crude Association**: The unadjusted association between treatment and mortality.
    -   Total treated ($A=1$): $40+60 = 100$. Deaths: $10+40=50$. Crude mortality: $50/100 = 50%$.
    -   Total untreated ($A=0$): $20+80=100$. Deaths: $5+20=25$. Crude mortality: $25/100 = 25%$.
    -   Crude risk difference: $50% - 25% = 25%$. This is a biased estimate.
2.  **Conditional Effects**: We calculate the risk difference within each stratum of the confounder $Z$.
    -   **Young Group (**$Z=0$): Risk difference is $25% - 25% = 0%$.
    -   **Old Group (**$Z=1$): Risk difference is $66.7% - 25% \approx 41.7%$.
3.  **Adjusted Estimate (Standardization)**: To get the Average Treatment Effect (ATE), we average the conditional effects across the marginal distribution of the confounder $Z$. The formula is:

$$
\mathbb{E}[Y^a] = \sum_{\vect{z}} \mathbb{E}[Y \mid A=a, \vect{Z}=\vect{z}] \mathbb{P}(\vect{Z}=\vect{z})
$$

**What this formula means**: "The average outcome if everyone in the population received treatment $a$ is the sum of the average outcome of those who received treatment $a$ *within each confounder stratum*, weighted by the proportion of the population that falls into that stratum."

-   First, we calculate the probabilities for our confounder, $Z$:
    -   $\mathbb{P}(Z=0) = 60/200 = 0.3$.
    -   $\mathbb{P}(Z=1) = 140/200 = 0.7$.
-   Now, let's plug in the numbers to find the ATE:
    -   $\text{ATE} = (\text{conditional effect for } Z=0) \times \mathbb{P}(Z=0) + (\text{conditional effect for } Z=1) \times \mathbb{P}(Z=1)$
    -   $\text{ATE} = (0% \times 0.3) + (41.7% \times 0.7) = 0 + 29.2% = 29.2%$.
-   The adjusted causal effect is a 29.2% increase in mortality. This is the unbiased estimate of the causal effect.

## Apply with Code

Let's simulate this scenario in R and use regression to adjust for the confounder.

```{r}
# Load Necessary Libraries
pacman::p_load(dagitty, dplyr, ggplot2, ggdag, tidyverse)

# 1. Visualize the DAG and Identify the Adjustment Set
dag <- dagitty(
  "dag {
    A [exposure, pos=\"0,0\"];
    Y [outcome, pos=\"1,0\"];
    Z [confounder, pos=\"0.5,1\"];
    Z -> A;
    Z -> Y;
    A -> Y;
}"
)
ggdag(dag) + theme_dag()
```


```{r}
#| label: dag-confounding
adjustment_sets <- adjustmentSets(
  dag,
  exposure = "A",
  outcome = "Y",
  effect = "total"
)
print(adjustment_sets)
# The Output Should Confirm that { Z } is a Valid Adjustment Set.
```

```{r}
#| label: tbl-raw-confounding-simulation
#| tbl-cap: "Simulated Data Reflecting Confounding by Age"
# 2. Simulate Data
set.seed(579) # For reproducibility
n <- 200
Z <- rbinom(n, 1, 0.7) # Age (0=young, 1=old)
A <- rbinom(n, 1, prob = case_when(Z == 0 ~ 0.4, Z == 1 ~ 0.6))
Y <- rbinom(
  n,
  1,
  prob = case_when(
    A == 0 & Z == 0 ~ 0.25,
    A == 1 & Z == 0 ~ 0.25,
    A == 0 & Z == 1 ~ 0.25,
    A == 1 & Z == 1 ~ 0.667
  )
)
data <- data.frame(A, Y, Z)
head(data)
```

```{r}
#| label: tbl-fit-model-confounding-simulation
#| tbl-cap: "Crude (Unadjusted) Logistic Regression"
# 3. Perform the Crude (unadjusted) Analysis
crude_model <- glm(Y ~ A, data = data, family = binomial(link = "logit"))
summary(crude_model)
cat("Crude odds ratio for A:", exp(coef(crude_model)["A"]), "\n")
```

```{r}
#| label: tbl-fit-model-adjusted-confounding-simulation
#| tbl-cap: "Adjusted Logistic Regression Controlling for Confounder Z"
# 4. Perform the Adjusted Analysis
adjusted_model <- glm(Y ~ A + Z, data = data, family = binomial(link = "logit"))
summary(adjusted_model)
cat(
  "Adjusted odds ratio for A (within strata of Z):",
  exp(coef(adjusted_model)["A"]),
  "\n"
)
# The Coefficient for A in This Model Provides an Unbiased Estimate of the *conditional* Causal Effect.
```

## Knowledge Check

1.  **Question**: In the context of causal inference, what is the key difference between unconditional and conditional exchangeability?
    -   **Answer**: **Unconditional exchangeability** is the ideal state in an RCT where treatment groups are balanced on all confounders. **Conditional exchangeability** is an assumption in observational studies that treatment groups are exchangeable *within strata* of a specific set of measured covariates, allowing for causal inference by adjustment.
2.  **Question**: Describe a backdoor path from an exposure to an outcome using a simple DAG. What is the purpose of adjusting for a variable on this path?
    -   **Answer**: A backdoor path is a non-causal path from the exposure ($A$) to the outcome ($Y$) that begins with an arrow pointing *into* $A$. For example, in the DAG $A \leftarrow Z \rightarrow Y$, the path is $A \leftarrow Z \rightarrow Y$. Adjusting for a variable like $Z$ on this path "blocks" or "closes" it, removing the confounding association and isolating the causal effect of $A$ on $Y$.
3.  **Question**: A study on the effect of exercise ($A$) on blood pressure ($Y$) adjusts for both diet ($Z$) and a genetic marker ($C$). The DAG shows $A \leftarrow Z \rightarrow Y$, $A \to C \leftarrow Y$. Based on this DAG, is adjusting for $C$ a good idea? Why or why not?
    -   **Answer**: No, adjusting for $C$ is a bad idea. In this DAG, $C$ is a **collider** on the path between $A$ and $Y$. Conditioning on a collider opens a non-causal path, which can induce a spurious association and bias your estimate. You should only adjust for the confounder $Z$.

## Reflective Prompt

Consider the difference between confounding and selection bias. Can you use the graphical framework (DAGs) to describe a scenario where you have only confounding, a scenario with only selection bias, and a scenario with both? What variables would you need to adjust for in each case to get an unbiased estimate?
