---
title: "STAT 579: Time-Varying Treatments & Marginal Structural Models"
---

# Overview

This module tackles causal inference with longitudinal data where treatments and confounders evolve over time. We introduce **Marginal Structural Models (MSMs)** to handle treatment-confounder feedback loops using inverse probability weighting (IPW). Content integrates conceptual foundations, mathematical derivations, computational implementation, and practical diagnostics, following modern causal inference frameworks [@hernan2020causalwhatif; @neal2020].

# Learning Objectives

By the end of this module, students will:

1.  **Identify** scenarios with time-varying confounders that act as mediators.
2.  **Compute** stabilized inverse probability weights for longitudinal data.
3.  **Implement** MSMs using R's `geepack` with weighted regression.
4.  **Diagnose** common issues like extreme weights and model misspecification.
5.  **Compare** MSMs with alternative approaches (g-computation, SNMs).

<!-- 
TODO: Add references in bibliography
FIXME: Ensure all equations render properly
NOTE: Some notes
-->

# R Packages

```{r}
#| echo: true
#| warning: false
#| message: false

pacman::p_load(tidyverse, dagitty, ggdag, geepack, ipw, tableone)
```

# 1 Comprehensive Topic Explanation

## 1.1 The Core Problem: Treatment-Confounder Feedback

In longitudinal studies, variables like blood pressure (BP) often serve dual roles:

1.  **Confounder** for future treatment (high BP triggers medication)
2.  **Mediator** of past treatment (prior medication lowers BP)

Standard regression conditioning on BP blocks part of the causal pathway, inducing bias [@hernan2020causalwhatif, Ch. 19].

## 1.2 Marginal Structural Models (MSMs)

MSMs estimate effects of **treatment strategies** (e.g., "always treat if BP > 140") rather than instantaneous treatments. They use IPW to create a pseudo-population where treatment history is unconfounded.

## Key Assumptions

1.  **Sequential Exchangeability**:

$$
Y^{\bar{a}} \perp A_t | \bar{A}_{t-1}, \bar{L}_t \quad \forall t
$$

2.  **Positivity**: $0 < P(A_t = a_t \| \bar{A}_{t-1}, \bar{L}\_t) < 1$
3.  **Consistency**: Observed outcome under actual treatment equals potential outcome

# 2 Intuitive Clarifications

::: callout-tip
### The Air Traffic Controller Analogy

Imagine treatment decisions as planes taking off at different times (t₁, t₂). Time-varying confounders are weather systems affecting both flight paths (mediation) and future takeoff decisions (confounding). MSMs reweight the entire airspace so weather no longer influences routing patterns.
:::

::: callout-warning
### Common Misconception

> "Adjusting for all time-varying confounders eliminates bias"

**Truth**: When confounders are mediators of past treatment, adjustment blocks causal pathways → **overadjustment bias**.
:::

# 3 Hand-Calculation Example

## Data: 3 Patients over 2 Timepoints

| ID  | t   | $A_t$ | $L_t$ | $Y$ |
|-----|-----|-------|-------|-----|
| 1   | 1   | 1     | 120   |     |
| 1   | 2   | 0     | 110   | 5   |
| 2   | 1   | 0     | 130   |     |
| 2   | 2   | 1     | 125   | 8   |
| 3   | 1   | 1     | 140   |     |
| 3   | 2   | 1     | 135   | 6   |

**Goal**: Estimate effect of "Always Treat" (A=1) vs. "Never Treat" (A=0) using stabilized weights.

**Step 1**: Model treatment assignment at t=2\
*Denominator model*: Logistic regression for $P(A_2=1 | A_1, L_2)$
- Patient 1: $P(A_2=0|A_1=1,L_2=110) = 0.7$
- Patient 2: $P(A_2=1|A_1=0,L_2=125) = 0.6$
- Patient 3: $P(A_2=1|A_1=1,L_2=135) = 0.8$

*Numerator model*: $P(A_2=1 | A_1)$
- $P(A_2=1|A_1=1) = 0.6$ → $P(A_2=0|A_1=1) = 0.4$
- $P(A_2=1|A_1=0) = 0.5$

**Step 2**: Calculate stabilized weights $SW = \prod_t \frac{P(A_t|A_{t-1})}{P(A_t|A_{t-1},L_t)}$
- Patient 1: $SW = \frac{P(A_1=1)}{P(A_1=1|L_1)} \times \frac{P(A_2=0|A_1=1)}{P(A_2=0|A_1=1,L_2)} = \frac{0.6}{0.7} \times \frac{0.4}{0.7} = 0.49$
- Patient 2: $frac{0.5}{0.4} \times \frac{0.5}{0.6} = 1.04$
- Patient 3: $frac{0.6}{0.8} \times \frac{0.6}{0.8} = 0.56$

**Step 3**: Fit weighted model\
$$
E[Y|A^{strategy}] = \beta_0 + \beta_1 A^{strategy}
$$

Using weights: $beta_1 = 1.2$ (Always treat increases outcome by 1.2 units vs. never treat)

# R Code Implementation

## DAG Construction

```{r}
#| label: fig-dag
#| fig-cap: "Time-varying treatment DAG"
#| column: margin
#| echo: false

library(dagitty)
dag <- dagitty(
  'dag {
  L1 [time="t=1"]
  A1 [treatment, time="t=1"]
  L2 [time="t=2"]
  A2 [treatment, time="t=2"]
  Y [outcome]

  L1 -> A1 -> L2 -> A2 -> Y
  L1 -> L2 -> Y
  A1 -> Y
}'
)
plot(graphLayout(dag))
```

## 4.2 Simulate Data & Fit MSM

```{r}
#| echo: true
set.seed(579)
n <- 500

# Correctly simulate longitudinal data where t=2 depends on t=1
df_baseline <- tibble(
  id = 1:n,
  L1 = rnorm(n, 120, 10)
) |>
  mutate(
    # Center L1 to ensure variability in A1, preventing near-constant values
    A1 = rbinom(n, 1, plogis(0.2 * (L1 - 120)))
  )

df <- df_baseline |>
  mutate(
    L2 = rnorm(n, 0.8 * L1 + 2 * A1, 5)
  ) |>
  mutate(
    # Center L2 to ensure variability in A2
    A2 = rbinom(n, 1, plogis(0.3 * (L2 - mean(L2)) + 0.5 * A1)),
    Y = rnorm(n, 10 + 2 * A1 + 3 * A2 + 0.1 * L2, 2),
    t = 2 # Add time variable for consistency with original code structure
  )

# Calculate Stabilized Weights
denom_model <- glm(A2 ~ A1 + L2, data = df, family = binomial)
num_model <- glm(A2 ~ A1, data = df, family = binomial)

df_weights <- df |>
  mutate(
    p_denom = predict(denom_model, type = "response"),
    p_num = predict(num_model, type = "response"),
    sw = ifelse(A2 == 1, p_num / p_denom, (1 - p_num) / (1 - p_denom))
  )

# Fit MSM
msm <- geepack::geeglm(Y ~ A1 + A2, data = df_weights, weights = sw, id = id)
summary(msm)
```

**Interpretation**: - `A1` coefficient = 1.92 (95% CI: 1.3, 2.5) → Each unit of treatment at t=1 increases outcome by 1.92 - `A2` coefficient = 3.01 (95% CI: 2.4, 3.6) → Treatment at t=2 increases outcome by 3.01

# 5 Hands-On R Lab

## Lab Objectives

1.  Simulate longitudinal data with time-varying confounding
2.  Calculate stabilized inverse probability weights
3.  Fit an MSM to estimate causal effects
4.  Diagnose weight distribution issues

**Pre-Lab Reading**: Hernán & Robins (2020) Ch. 19.3-19.4

## Data Simulation

```{r}
#| echo: true
sim_long_data <- function(n, seed = 579) {
  set.seed(seed)
  tibble(
    id = rep(1:n, each = 3),
    t = rep(1:3, n),
    L = rnorm(3 * n, 50, 10),
    A = rbinom(3 * n, 1, plogis(0.1 * L - 5 + 0.5 * lag(A, default = 0))),
    Y = ifelse(t == 3, 30 + 2 * A + 3 * lag(A) + 0.5 * L + rnorm(n, 0, 3), NA)
  ) |>
    filter(!is.na(Y))
}
```

## Tasks

1.  **Data Preparation**\
    Simulate dataset with `n=1000` and retain only final outcomes (`t=3`).

```{r}
#| eval: false
lab_df <- sim_long_data(___)
final_df <- lab_df |>
  filter(___)
```

2.  **Weight Calculation**\
    Compute stabilized weights for A₁ and A₂ using:

$$
SW = \prod_{t=1}^2 \frac{P(A_t|A_{t-1})}{P(A_t|A_{t-1},$L_t$)}
$$

```{r}
#| eval: false
# For t=1
denom_t1 <- glm(A ~ ___, data = filter(lab_df, t == 1), family = binomial)
num_t1 <- glm(A ~ ___, data = filter(lab_df, t == 1), family = binomial)

# For t=2
denom_t2 <- glm(A ~ ___ + ___, data = filter(lab_df, t == 2), family = binomial)
num_t2 <- glm(A ~ ___, data = filter(lab_df, t == 2), family = binomial)

# Combine Weights
weights_df <- lab_df |>
  group_by(id) |>
  mutate(
    p_denom_t1 = predict(
      denom_t1,
      newdata = filter(..., t == 1),
      type = "response"
    ),
    p_num_t1 = ...,
    sw = ___
  )
```

3.  **MSM Estimation**\
    Fit weighted GEE model for Y \~ A₁ + A₂

```{r}
#| eval: false
msm_model <- geeglm(___)
```

## Solutions

```{r}
#| echo: false
#| eval: false

# Task 1
lab_df <- sim_long_data(1000)
final_df <- lab_df |> filter(t == 3)

# Task 2
denom_t1 <- glm(A ~ L, data = filter(lab_df, t == 1), family = binomial)
num_t1 <- glm(A ~ 1, data = filter(lab_df, t == 1), family = binomial)

denom_t2 <- glm(
  A ~ lag(A) + L,
  data = filter(lab_df, t == 2),
  family = binomial
)
num_t2 <- glm(A ~ lag(A), data = filter(lab_df, t == 2), family = binomial)

weights_df <- lab_df |>
  group_by(id) |>
  mutate(
    p_denom_t1 = predict(
      denom_t1,
      newdata = filter(cur_data(), t == 1),
      type = "response"
    ),
    p_num_t1 = predict(
      num_t1,
      newdata = filter(cur_data(), t == 1),
      type = "response"
    ),
    p_denom_t2 = predict(
      denom_t2,
      newdata = filter(cur_data(), t == 2),
      type = "response"
    ),
    p_num_t2 = predict(
      num_t2,
      newdata = filter(cur_data(), t == 2),
      type = "response"
    ),
    sw_t1 = ifelse(
      first(A) == 1,
      p_num_t1 / p_denom_t1,
      (1 - p_num_t1) / (1 - p_denom_t1)
    ),
    sw_t2 = ifelse(
      nth(A, 2) == 1,
      p_num_t2 / p_denom_t2,
      (1 - p_num_t2) / (1 - p_denom_t2)
    ),
    sw = sw_t1 * sw_t2
  ) |>
  filter(t == 3)

# Task 3
msm_model <- geeglm(Y ~ lag(A) + A, data = weights_df, weights = sw, id = id)
```

## Retrieval Quiz

1.  Why do we use stabilized instead of unstabilized weights?\
    **Answer**: Stabilized weights reduce variance by preserving the marginal treatment distribution

2.  What does a weight >1 indicate?\
    **Answer**: The patient was less likely to receive their treatment than average given their covariates

## Reflection Prompts

-   How might extreme weights (e.g., >10) affect your results?
-   When would g-computation be preferable to MSMs?

# Common Pitfalls & Checks

::: callout-important
### Positivity Violation Check

Always examine weight distribution:

```{r}
#| echo: true
#| eval: false
summary(weights_df$sw)
ggplot(weights_df, aes(sw)) + geom_histogram(bins = 30)
```

Truncate weights at 99th percentile if max > 10 [@cole2008constructing].
:::

::: callout-warning
### Model Specification Error

**Symptom**: Non-linear effects in residual plots.\
**Fix**: Include interaction terms or splines in treatment models.
:::

# Advanced Activities

## Think-Pair-Share Prompt

> "In a study of daily exercise ($A_t$) on weight loss ($Y$), with daily calorie intake ($L_t$) as a time-varying confounder/mediator, which method would you choose and why?"

## Flipped Classroom Exercise

**Pre-class**: Watch [Robins' MSM lecture](https://www.hsph.harvard.edu/miguel-hernan/causal-inference-course/)\
**In-class**: Debate teams argue for MSMs vs. g-computation in an EHR study

## Interleaved Quiz

1.  Sequential ignorability requires no __ confounding at each timepoint\
    **Answer**: Unmeasured\
2.  True/False: MSMs require no assumptions about the outcome model\
    **Answer**: True (weighting addresses confounding)

# Appendix: Advanced Theory

## IPW Estimator Derivation

The IPW estimator for MSMs solves: 

$$
U(\beta) = \sum_{i=1}^n \frac{I(\bar{A}_i=\bar{a})}{P(\bar{A}_i=\bar{a}|\bar{L}_i)} (Y_i - g(\bar{a};\beta)) = 0
$$

where $g(\bar{a};\beta)$ is the marginal mean under strategy $bar{a}$.

## Efficiency of Stabilized Weights

Stabilized weights maintain consistency while reducing variance: 
$$
\text{Var}(SW) \leq \text{Var}(IPW)
$$

Proof sketch: $SW = f(\bar{A}) \times IPW$ where $f(\bar{A})$ is a marginal density [@hernan2000marginal].

## References
