---
title: "Week 10: Time-Varying Treatments & Marginal Structural Models ⏳"
---

## Learning Objectives

By the end of this module, you will be able to:

- **Explain** treatment–confounder feedback and why standard adjustment fails (Hernán & Robins, *What If*, §19–20).
- **State and justify** the three core identification assumptions—sequential exchangeability, positivity, and consistency—and diagnose threats to each.
- **Derive and compute** stabilized inverse probability weights (IPWs) for multi-time treatments (Robins, Hernán & Brumback, 2000).
- **Estimate** parameters of a marginal structural model (MSM) with weighted GEE/GLM and **interpret** them as effects of sustained strategies.
- **Compare and contrast** MSM/IPW with the g-formula and structural nested models; articulate when each is advantageous (Neal, *Intro to CI*, Ch. 10–11).
- **Evaluate diagnostics** (weight distribution, truncation, covariate balance) and implement sensitivity checks.

## R Packages

```{r}
#| label: load-packages
#| message: false
#| warning: false
pacman::p_load(
  dagitty, # causal DAGs
  ggdag, # DAG plotting
  dplyr, # wrangling
  tidyr, # reshaping
  tibble, # tibbles
  purrr, # iteration
  ggplot2, # visualization
  geepack, # GEE for MSMs
  survey, # weighted GLMs
  WeightIt, # MSM weights
  cobalt, # balance diagnostics
  sandwich, # robust vcov
  lmtest # coeftest
)
```

## 1 Comprehensive Topic Explanation

### 1.1 The Longitudinal Setup

We observe for each individual $i$ across $k=0,\dots,K$:

- Time-varying covariates $\bar L_k=(L_0,\dots,L_k)$,
- Treatments $\bar A_k=(A_0,\dots,A_k)$,
- Outcome $Y$ at $K{+}1$.

**Treatment–confounder feedback:** $L_k$ both (i) confounds $A_k \to Y$ and (ii) lies on the pathway $A_{k-1}\to L_k \to Y$. Adjusting for $L_k$ in a standard outcome regression blocks a piece of the very effect we want (mediated path), inducing bias or sign reversal (Hernán & Robins §19).

### 1.2 Target Estimands & MSMs

We shift from “effect of $A_k$ given $L_k$” to the **marginal effect of a sustained strategy**, e.g., “treat whenever biomarker $> x$.” MSMs parameterize the **marginal** counterfactual mean under a regime:

$$
\mathbb{E}\{Y^{\bar a}\} = m(\bar a; \beta),
$$

e.g., a logit-linear MSM for two time points:

$$
\text{logit}\{\mathbb{E}(Y^{a_0,a_1})\}=\beta_0+\beta_1 a_0+\beta_2 a_1+\beta_3 a_0 a_1.
$$

Parameters $\beta$ are interpreted as **causal contrasts of sustained histories**.

### 1.3 Identification Assumptions (Hernán & Robins §19; Neal Ch. 10)

- **Sequential exchangeability:**
  $Y^{\bar a}\ \indep\ A_k \mid \bar L_k,\bar A_{k-1}$ for all $k$.
  Given observed history, treatment is as-if random.
- **Positivity:**
  $\Pr(A_k=a_k\mid \bar L_k=\bar \ell_k,\bar A_{k-1}=\bar a_{k-1})>0$ whenever $\Pr(\bar L_k=\bar \ell_k,\bar A_{k-1}=\bar a_{k-1})>0$.
- **Consistency:**
  If an individual follows $\bar A=\bar a$, then $Y=Y^{\bar a}$.

### 1.4 IPW Construction & Stabilized Weights

To break confounding by $\bar L_k$, define **stabilized weight** for individual $i$:

$$
SW_i=\prod_{k=0}^{K}\frac{\Pr(A_{ik}=a_{ik}\mid \bar A_{i,k-1})}
{\Pr(A_{ik}=a_{ik}\mid \bar A_{i,k-1},\bar L_{ik})}.
$$

Intuition: **upweight** histories that were unlikely to get the treatment they received *given* their covariates; **downweight** those that were likely. In the weighted pseudo-population, $\bar L_k \not\to A_k$ by design; MSM parameters can be fit by weighted regression/GEEs (Robins, Hernán & Brumback, 2000).

### 1.5 Estimation Strategies

- **IPTW of the marginal mean:** estimate $\mathbb{E}(Y^{\bar a})$ by weighting each unit by the inverse probability of their observed treatment path (restricted to those following $\bar a$ when estimating that regime).
- **Weighted GLM/GEE (common in practice):** regress $Y$ on functions of $\bar A$ per the MSM using $SW$ as weights; use robust (sandwich) SEs.
- **Variance:** robust sandwich; cluster on ID for repeated measures.
- **Censoring:** incorporate **censoring weights** by multiplying treatment weights with inverse probabilities of remaining uncensored (Hernán & Robins §12, §19).

### 1.6 MSMs Vs G-formula Vs SNMs (compare–contrast)

- **MSM + IPW:** Targets marginal regime effects; robust to outcome model misspecification but sensitive to treatment model/positivity & weight instability.
- **G-formula (parametric g-comp):** Models $Y \mid$ history and simulates under regimes. Sensitive to outcome model misspecification; no unstable weights but more modeling.
- **SNMs/g-estimation:** Alternative parameterization focusing on treatment blip effects; useful with time-varying treatments and confounding; often less familiar operationally.

::: {.callout-note}
## Decision Guide
 **When choose what?**
 - Strong domain knowledge to specify $Y\mid$ history → **g-formula**.
 - Cleaner treatment model & desire marginal interpretation → **MSM**.
 - Focus on dynamic blip functions & structural nesting → **SNM**.
:::

Citations: Hernán & Robins §12, §19–20; Neal Ch. 10–11; Robins (1986, 1997, 2000); VanderWeele Ch. 19.

## 2 Intuitive Clarifications of Challenging Ideas

::: {.callout-tip}

## Learning Tip

Think of **IPW** as “re-queuing the study” so that—conditional on past treatments—the *next* treatment looks like a coin flip, independent of the covariates that used to influence it.

:::

**Visual metaphor:**
- **Pipes and valves.** Past treatment opens/closes valves that change $L_k$ (flow rate). Those $L_k$ then steer future treatment. IPW temporarily “disconnects” the $L_k \to A_k$ pipe so we can see the net effect of opening the treatment valves over time.

**Common misconceptions (and fixes):**
- *“I adjusted for $L_k$ in an outcome regression, so I controlled confounding.”*
  **No**—you also block mediated effects of earlier $A$ through $L_k$, biasing the target marginal effect (Hernán & Robins §19).
- *“Large weights just mean more precision.”*
  **No**—they usually signal near-positivity violations; consider truncation, coarsening, or redefining the target regime/support.
- *“Stabilized weights always fix everything.”*
  They reduce variance, not bias from misspecified models or violations of assumptions.

**DAG for treatment–confounder feedback (drawn below) clarifies the double role of $L_1$.**

```{r}
#| label: fig-msm-dag-basic
#| fig-cap: "Minimal DAG with treatment–confounder feedback at two times"
#| column: margin
dag_basic <- dagitty(
  "
dag {
  L0; A0; L1; A1; Y
  L0 -> A0
  L0 -> L1
  A0 -> L1
  L1 -> A1
  A0 -> Y
  A1 -> Y
  L1 -> Y
}"
)
ggdag(dag_basic, layout = "nicely") + theme_dag()
```

## 3 Simple, Hand-Calculation Numerical Example(s)

### 3.1 Four-person Toy Example (IPW mechanics)

Two times ($k=0,1$), all binary. Data:

| ID  | L0  | A0  | L1  | A1  |  Y  |
| :-: | :-: | :-: | :-: | :-: | :-: |
|  1  |  0  |  0  |  0  |  0  |  0  |
|  2  |  0  |  1  |  1  |  1  |  1  |
|  3  |  1  |  0  |  1  |  0  |  0  |
|  4  |  1  |  1  |  0  |  1  |  1  |

**Step A (fit by hand):** empirical treatment probabilities

- $P(A_0{=}1\mid L_0{=}0)=1/2$, $P(A_0{=}1\mid L_0{=}1)=1/2$
- $P(A_1{=}1\mid A_0{=}1, L_1{=}0)=1$, $P(A_1{=}1\mid A_0{=}0,L_1{=}0)=0$,
  $P(A_1{=}1\mid A_0{=}1,L_1{=}1)=1$, $P(A_1{=}1\mid A_0{=}0,L_1{=}1)=0$ (from table)

**Numerators** (stabilized) use reduced models:
- $P(A_0{=}1)=1/2$; $P(A_1{=}1\mid A_0)$ equals 1 when $A_0{=}1$, and 0 when $A_0{=}0$ (in this tiny table).

**Step B (compute each $SW_i$):**
Example, ID 2 has $(A_0,A_1)=(1,1)$ with $(L_0,L_1)=(0,1)$:

$$
SW_2=\frac{P(A_0{=}1)}{P(A_0{=}1\mid L_0{=}0)}\cdot
      \frac{P(A_1{=}1\mid A_0{=}1)}{P(A_1{=}1\mid A_0{=}1, L_1{=}1)}
= \frac{1/2}{1/2}\cdot\frac{1}{1}=1.
$$

Do likewise for IDs 1,3,4 (here, all weights turn out 1 due to deterministic $A_1$ within strata). This *illustrates* stabilized weight definition even if trivial.

**Step C (weighted mean under regime “always treat”)**
To estimate $\mathbb{E}(Y^{1,1})$, keep only those with $(A_0,A_1)=(1,1)$ (IDs 2 and 4), and compute the **weighted** mean of $Y$: $(1\cdot1 + 1\cdot1)/(1+1)=1$.
For “never treat,” use $(0,0)$ (ID 1 only): mean $=0$ (ID 3 has $(0,0)$? No; it’s $(0,0)$ at $A$, yes—$Y=0$).
Estimated effect (risk difference) $\approx 1-0=1$ in this toy.

::: {.callout-important}

## 📌 Key Assumption

This hand example has **deterministic** $A_1$ given $(A_0, L_1)$, making positivity fragile. Real data rarely behave so cleanly—expect non-trivial weights and perform diagnostics.

:::

### 3.2 Scaffolding Extension (non-trivial numerics)

Modify the table so $A_1$ sometimes flips in $(A_0{=}1,L_1{=}0)$ with probability $0.7$ and in $(A_0{=}1,L_1{=}1)$ with $0.6$. Recompute denominators, numerators (e.g., $P(A_1{=}1\mid A_0{=}1)=0.65$), then each $SW_i$. Observe how **rare** patterns get **larger** weights.

## 4 R Code Examples & Interpretation

### 4.1 DAGs for Feedback and for Censoring (advanced)

```{r}
#| label: fig-msm-dag-censor
#| fig-cap: "MSM DAG with time-varying confounders and censoring"
#| column: margin
dag_censor <- dagitty(
  "
dag {
  L0; A0; L1; A1; C1; Y
  L0 -> A0
  L0 -> L1
  A0 -> L1
  L1 -> A1
  L1 -> C1
  A0 -> Y
  A1 -> Y
  L1 -> Y
  C1 -> Y
}"
)
ggdag(dag_censor, layout = "nicely") + theme_dag()
```

Interpretation: $C_1$ (censoring/loss to follow-up) depends on $L_1$; we need **censoring weights** in addition to treatment weights when $Y$ is missing for some.

### 4.2 Basic Simulation: Two time Points, Binary Outcome

```{r}
#| label: fig-sim-weights
#| fig-cap: "Distribution of stabilized weights (basic simulation)"
#| column: margin
set.seed(123)
n <- 4000
L0 <- rbinom(n, 1, 0.5)
A0 <- rbinom(n, 1, plogis(-0.5 + 1.0 * L0))
L1 <- rbinom(n, 1, plogis(-0.2 + 1.2 * A0 + 0.6 * L0))
A1 <- rbinom(n, 1, plogis(-0.8 + 1.6 * L1 + 0.3 * A0))
linY <- -0.3 + 0.9 * A0 + 1.1 * A1 + 0.5 * L1
Y <- rbinom(n, 1, plogis(linY))
dat <- tibble(id = 1:n, L0, A0, L1, A1, Y)

# Denominator Models (rich)
pA0_den <- glm(A0 ~ L0, family = binomial, data = dat)$fitted.values
pA1_den <- glm(A1 ~ A0 + L0 + L1, family = binomial, data = dat)$fitted.values

# Numerator Models (reduced)
pA0_num <- glm(A0 ~ 1, family = binomial, data = dat)$fitted.values
pA1_num <- glm(A1 ~ A0, family = binomial, data = dat)$fitted.values

dat <- dat |> mutate(sw = (pA0_num / pA0_den) * (pA1_num / pA1_den))

ggplot(dat, aes(sw)) +
  geom_histogram(bins = 50)
```

**Interpretation.** A tight weight distribution suggests OK overlap; heavy tails signal potential positivity issues.

**Optional truncation & normalization:**

```{r}
#| label: tbl-sim-weights-trunc
#| tbl-cap: "Simulated weights with truncation and normalization"
alpha <- 0.01
lower <- quantile(dat$sw, alpha)
upper <- quantile(dat$sw, 1 - alpha)
dat <- dat |>
  mutate(
    sw_trunc = pmin(pmax(sw, lower), upper),
    sw_norm = sw / mean(sw),
    swt_norm = sw_trunc / mean(sw_trunc)
  )
summary(dat$sw)
summary(dat$sw_trunc)
```

### 4.3 MSM Estimation (weighted GLM/GEE) and Naïve Comparison

```{r}
#| label: tbl-msm-weights
#| tbl-cap: "Naive and MSM Weighted GEE Estimates with Truncation"
# Naïve (condition on L1 incorrectly)
naive <- glm(Y ~ A0 + A1 + L1, family = binomial, data = dat)
summary(naive)

# MSM with Weighted GEE (independence Working correlation)
msm <- geeglm(
  Y ~ A0 + A1,
  id = id,
  weights = sw_trunc,
  data = dat,
  family = binomial,
  corstr = "independence"
)
summary(msm)
```

**Compare–contrast:** The naïve model adjusts for $L_1$ (a mediator of $A_0$), biasing $\hat\beta_{A0}$. The MSM targets **marginal** effects; coefficients approximate the causal log-odds effects of treatment at each time.

### 4.4 Diagnostics: Balance with `WeightIt` + `cobalt` (scaffolded)

```{r}
#| label: fig-msm-weightit
#| fig-cap: "WeightIt Stabilized Weights for MSM"
# WeightIt for MSM (two time points)
w_msm <- weightitMSM(
  list(A0 ~ L0, A1 ~ A0 + L0 + L1),
  data = dat,
  method = "ps",
  stabilize = TRUE
)
all.equal(as.numeric(w_msm$weights), dat$sw, tolerance = 1e-6) # similar

# Balance Diagnostics at Each K Using Cobalt
bal <- bal.tab(w_msm, m.threshold = .1, un = TRUE)
bal
love.plot(bal)
```

**Interpretation.** Covariate balance (standardized mean differences) should improve across treatment groups after weighting, **at the times they act as confounders**.

### 4.5 Advanced Simulation: Add Censoring & Censored Weights

```{r}
#| label: tbl-sim-weights-censored
#| tbl-cap: "Simulated weights with censoring"

set.seed(1)
dat2 <- dat |>
  mutate(C1 = rbinom(n(), 1, plogis(-2 + 1.3 * L1 - 0.6 * A1))) # 1 = censored
# Denominator and Numerator Models for Censoring (prob of *not* censored)
pC1_den <- glm(
  I(1 - C1) ~ A0 + A1 + L0 + L1,
  family = binomial,
  data = dat2
)$fitted.values
pC1_num <- glm(
  I(1 - C1) ~ A0 + A1,
  family = binomial,
  data = dat2
)$fitted.values
dat2 <- dat2 |>
  mutate(
    sc = (pC1_num / pC1_den),
    swc = sw * sc,
    swc_trunc = pmin(pmax(swc, lower), upper)
  )

# Fit MSM among Uncensored with Combined Weights
msm_c <- geeglm(
  Y ~ A0 + A1,
  id = id,
  weights = swc_trunc,
  data = filter(dat2, C1 == 0),
  family = binomial,
  corstr = "independence"
)
summary(msm_c)
```

**Note.** For time-to-event outcomes, use `survival::coxph` with weights (and robust variance).

## 5 Hands-On R Lab

**Purpose:** Practice computing stabilized weights, diagnosing overlap, and estimating MSMs.
**Pre-lab reading (flipped):** Hernán & Robins §19–20; Neal Ch. 10.
**Outcomes:** You will implement IPW for a 2-time MSM, compare with a naïve model, and reflect on assumptions and diagnostics.

### Lab Objectives

1. Simulate a dataset with treatment–confounder feedback.
2. Draw the causal DAG and mark confounder/mediator roles.
3. Compute stabilized weights (with and without truncation).
4. Fit naïve vs MSM models; interpret differences.
5. Assess weight distributions and covariate balance.
6. (Stretch) Add censoring weights and refit.

### Reading Warm-up (2–3 bullets)

- Why does adjusting for $L_1$ in $Y\sim A_0,A_1,L_1$ bias the effect of $A_0$? (*What If* §19)
- What does the **numerator** in stabilized weights accomplish?

### Task 1 — Simulate Data

```{r}
#| label: tbllab-sim-1
#| tbl-cap: "Simulated dataset for lab"
# Skeleton (fill in the … lines)
set.seed(10)
n <- 3000
L0 <- rbinom(n, 1, 0.5)
A0 <- rbinom(n, 1, plogis(-0.4 + 0.9 * L0))
L1 <- rbinom(n, 1, plogis(-0.1 + 1.1 * A0 + 0.5 * L0))
A1 <- rbinom(n, 1, plogis(-0.6 + 1.4 * L1 + 0.3 * A0))
Y <- rbinom(n, 1, plogis(-0.2 + 0.8 * A0 + 1.0 * A1 + 0.4 * L1))
lab <- tibble(id = 1:n, L0, A0, L1, A1, Y)
head(lab)
```

**Solution:**

```r
lab <- lab
```

### Task 2 — Draw the DAG

```{r}
#| label: fig-lab-dag
#| fig-cap: "Lab DAG: feedback at two times"
#| column: margin
dag_lab <- dagitty(
  "
dag {
  L0; A0; L1; A1; Y
  L0 -> A0
  L0 -> L1
  A0 -> L1
  L1 -> A1
  L1 -> Y
  A0 -> Y
  A1 -> Y
}"
)
ggdag(dag_lab, layout = "nicely") + theme_dag()
```

### Task 3 — Compute Stabilized Weights

```{r}
#| label: tbl-lab-sw
#| tbl-cap: "Lab stabilized weights"
# Denominators
pA0_d <- glm(A0 ~ L0, family = binomial, data = lab)$fitted.values
pA1_d <- glm(A1 ~ A0 + L0 + L1, family = binomial, data = lab)$fitted.values
# Numerators
pA0_n <- glm(A0 ~ 1, family = binomial, data = lab)$fitted.values
pA1_n <- glm(A1 ~ A0, family = binomial, data = lab)$fitted.values
lab <- lab |>
  mutate(sw = (pA0_n / pA0_d) * (pA1_n / pA1_d))
summary(lab$sw)
```

**Truncation (α = 1%):**

```{r}
#| label: tbl-lab-sw-trunc
#| tbl-cap: "Lab stabilized weights with truncation"
alpha <- 0.01
lwr <- quantile(lab$sw, alpha)
upr <- quantile(lab$sw, 1 - alpha)
lab <- lab |>
  mutate(sw_t = pmin(pmax(sw, lwr), upr))
head(lab)
```

### Task 4 — Fit Naïve Vs MSM

```{r}
naive_lab <- glm(Y ~ A0 + A1 + L1, family = binomial, data = lab)
msm_lab <- geeglm(
  Y ~ A0 + A1,
  id = id,
  weights = sw_t,
  data = lab,
  family = binomial,
  corstr = "independence"
)
summary(naive_lab)
summary(msm_lab)
```

**Interpretation prompts (write inline answers):**
- Compare $\hat\beta_{A0}$ across models. Which aligns with the data-generating truth?
- Why does including $L_1$ in the naïve model distort the $A_0$ coefficient?

### Task 5 — Diagnose Weights & Balance

```{r}
#| label: fig-lab-weights
#| fig-cap: "Lab: weight histograms"
p1 <- ggplot(lab, aes(sw)) +
  geom_histogram(bins = 50)
p2 <- ggplot(lab, aes(sw_t)) +
  geom_histogram(bins = 50)
p1
p2
```

```{r}
#| label: fg-lab-balance
#| fig-cap: "Lab: balance"
w_msm_lab <- weightitMSM(
  list(A0 ~ L0, A1 ~ A0 + L0 + L1),
  data = lab,
  method = "ps",
  corstr = "independence",
  stabilize = TRUE
)
bal_lab <- bal.tab(w_msm_lab, un = TRUE, m.threshold = .1)
bal_lab
love.plot(bal_lab)
```

### Task 6 — (Stretch) Add Censoring Weights

```{r}
set.seed(9)
lab2 <- lab |>
  mutate(C1 = rbinom(n(), 1, plogis(-2 + 1.0 * L1 - 0.5 * A1))) # 1=censored
pC1_d <- glm(
  I(1 - C1) ~ A0 + A1 + L0 + L1,
  family = binomial,
  data = lab2
)$fitted.values
pC1_n <- glm(I(1 - C1) ~ A0 + A1, family = binomial, data = lab2)$fitted.values
lab2 <- lab2 |>
  mutate(sc = pC1_n / pC1_d, swc = sw_t * sc)
msm_c_lab <- geeglm(
  Y ~ A0 + A1,
  id = id,
  weights = swc,
  data = filter(lab2, C1 == 0),
  family = binomial,
  corstr = "independence"
)
summary(msm_c_lab)
```

### Retrieval Quiz (answer in Complete sentences)

1. Define *sequential exchangeability* and provide a concrete example of when it is plausible.
2. Explain why **positivity** is a property of the data-generating process, not the analyst.
3. What role does the **numerator** of stabilized weights play?
4. Describe two diagnostics you would report to defend your MSM analysis.

### Reflection Prompts

- In your own study area, propose a sustained strategy (static or dynamic). Which covariates risk violating positivity, and how would you mitigate that risk?

## 6 Common Pitfalls & Checks

::: {.callout-warning}

## Common Pitfall

**Extreme or unstable weights** (heavy right tail) → inflated variance and sensitivity.
**Check:** histograms, quantiles; report $\max(sw)$ and $\mathrm{CV}(sw)$.
**Mitigate:** truncation (e.g., at 1st/99th), coarsen covariates, redefine estimand (narrow regime or support), or switch to g-formula if overlap is non-existent (Hernán & Robins §19–20).
:::

::: {.callout-warning}

## Common Pitfall

**Model misspecification** in treatment models → biased MSM even with perfect outcome model neutrality.
**Check:** include flexible terms (splines, interactions), consider Super Learner (if allowed), compare alternative specifications (Neal Ch. 10).
:::

::: {.callout-warning}

## Common Pitfall

**Adjusting for mediators** in the outcome model (e.g., adding $L_1$) undermines the marginal target.
**Check:** keep MSM outcome models *marginal in $\bar A$* and push confounding control into weights.
:::

::: {.callout-question}

## Check Your Understanding

Why might you **prefer** the g-formula over MSMs when weight tails remain heavy even after truncation?

:::

## 7 Advanced Teaching Activities & Interactivity

- **Think–Pair–Share:** Given a clinical p1olicy “treat if $L_k > c$,” decide if it’s static or dynamic; identify confounders and mediators at each time; sketch the DAG.
- **Problem-based case:** Simulate a cohort with near-deterministic treatment when $L_1$ is extreme. Attempt MSM, then redesign the target estimand to satisfy positivity (e.g., restrict to $L_1\in[0.2,0.8]$).
- **Spaced/interleaved quiz:** Alternate items asking for (i) weight formulas and (ii) g-formula simulation steps.
- **Self-explanation:** Students write a paragraph comparing the interpretation of $\beta_{A1}$ in MSM vs the coefficient of $A_1$ in a fully adjusted outcome regression.
- **Optional Shiny idea:** Live sliders for $A_1$ assignment strength and $L_1$ effect; show changes in weight tails and MSM bias.

## 8 Appendix: Advanced Theory & Derivations (optional)

### A.1 Identification Sketch (two times)

Under sequential exchangeability, positivity, and consistency,

$$
\mathbb{E}(Y^{a_0,a_1})
= \mathbb{E}\bigg[\frac{\mathbb{I}(A_0{=}a_0)\mathbb{I}(A_1{=}a_1)}
{\Pr(A_0{=}a_0\mid L_0)\Pr(A_1{=}a_1\mid A_0,L_0,L_1)}Y\bigg].
$$


Replacing denominators with estimated propensities yields IPTW. Stabilized weights multiply by numerators $\Pr(A_0{=}a_0)$ and $\Pr(A_1{=}a_1\mid A_0)$ to improve efficiency (Robins, Hernán & Brumback, 2000; Hernán & Robins §19).

### A.2 Efficiency and variance

Weighted estimating equations with independence working correlation yield consistent estimates of MSM parameters; robust sandwich variance accounts for misspecification of the working correlation (Liang & Zeger, 1986; see also Hernán & Robins §20).

### A.3 Dynamic Regimes

For a dynamic rule $d=(d_0,d_1)$ with $a_k=d_k(\bar L_k,\bar A_{k-1})$, replace indicators $\mathbb{I}(A_k{=}a_k)$ with $\mathbb{I}(A_k{=}d_k(\cdot))$ in the IPTW estimator; the MSM can include summaries of “following the rule” (Cunningham §10; VanderWeele §19).

### A.4 Numerical Illustration (mini)

Suppose $\Pr(A_0{=}1\mid L_0)=\text{logit}^{-1}(-0.5+L_0)$; draw 10 individuals; compute denominators and numerators, then $SW_i$. Summarize $\bar{SW}$, $\max(SW)$, and the truncated proportion. Observe how increasing the $L_1 \to A_1$ coefficient inflates $\max(SW)$.

## References (inline Citations in text)

- Hernán, M. A., & Robins, J. M. (2025). *What If*. (see §§12, 19–20).
- Neal, B. (2020). *Introduction to Causal Inference*. (Chs. 10–11).
- Robins, J. M., Hernán, M. A., & Brumback, B. (2000). Marginal structural models and causal inference in epidemiology.
- VanderWeele, T. (2015). *Explanation in Causal Inference* (Ch. 19).
- Cunningham, S. (2021). *Causal Inference: The Mixtape* (Ch. on panel/time-varying).

