---
title: "ggdag: A Tutorial for Causal DAGs in R"
execute: 
  echo: true
---

This tutorial follows the core logic of the `ggdag` package, showing how to define, analyze, and visualize causal models in R.

### The Core Idea: Tidy DAGs

The power of `ggdag` comes from its integration with the `tidyverse`. The package works by converting a `dagitty` graph object into a **tidy data frame** (a tibble). In this format, each row represents a single node, and columns contain information like the node's name, its coordinates for plotting, and its causal role (e.g., exposure or outcome).

This "tidy" structure is key because it allows you to use standard `ggplot2` functions to build and customize your DAG plots layer by layer.

Let's start by defining a simple confounding model and then use `tidy_dagitty()` to see its underlying data structure.

```{r}
# Load the necessary packages
if (!require("pacman")) install.packages("pacman")
pacman::p_load(ggdag, tidyverse)

# Define a DAG where Z confounds the X -> Y relationship
confounder_dag <- dagify(
    Y ~ X + Z,
    X ~ Z,
    exposure = "X",
    outcome = "Y"
)

# Convert the DAG to a tidy data frame
tidy_dag <- tidy_dagitty(confounder_dag)

# Inspect the tidy output
print(tidy_dag)
#> # A DAG with 3 nodes and 3 edges
#> #
#> # Exposure: X
#> # Outcome: Y
#> #
#> # A tibble: 3 × 8
#>   name      x     y direction to     xend  yend circular
#>   <chr> <dbl> <dbl> <fct>     <chr> <dbl> <dbl> <lgl>
#> 1 X      3.37  2.01 ->        Y      4.41  2.01 FALSE
#> 2 Y      4.41  2.01 <NA>      <NA>  NA    NA    FALSE
#> 3 Z      3.89  2.97 ->        X      3.37  2.01 FALSE
```

As you can see, the DAG is now represented as a simple table, making it easy for `ggplot2` to map variables like `x`, `y`, `xend`, and `yend` to plotting aesthetics.

-----

### Plotting DAGs with `ggdag()`

The primary plotting function, `ggdag()`, takes a DAG object and renders it. You can use themes like `theme_dag()` or `theme_dag_blank()` for a clean appearance.

```{r}
# Plot our simple confounding DAG
ggdag(confounder_dag) +
    theme_dag()
```

DAGs are particularly useful for understanding complex causal structures, like the "M-bias" graph, where two variables are ancestors of the exposure and outcome, respectively.

```{r}
# Define an M-bias DAG with unobserved variables U1 and U2
m_bias_dag <- dagify(
    Y ~ M2 + U1,
    X ~ M1,
    M1 ~ U2,
    M2 ~ U1,
    exposure = "X",
    outcome = "Y",
    latent = c("U1", "U2")
)

# Plot the M-bias DAG
ggdag(m_bias_dag) +
    theme_dag()
```

-----

### Identifying Causal Paths and Bias

The main analytical purpose of a DAG is to identify which variables you must **adjust for** to estimate a causal effect, free of confounding and other biases. This is a three-step process.

#### 1. Visualize All Paths with `ggdag_paths()`
```{r}
ggdag_paths(m_bias_dag) +
    theme_dag()
```

First, we identify all open paths between the exposure and outcome. A path is a route connecting two variables; it's "open" if it creates a statistical association. An open path that is not a causal path is a "backdoor" path that introduces bias.

```{r}
ggdag_paths(confounder_dag) +
    theme_dag()
```

The plot clearly shows the causal path of interest (`X → Y`) and the biasing backdoor path (`X ← Z → Y`). Our goal is to block this backdoor path.

#### 2. Find the Adjustment Set with `ggdag_adjustment_set()`

Next, we use `ggdag_adjustment_set()` to automatically find a **sufficient adjustment set**. This is a minimal set of variables that, if conditioned on (e.g., included as covariates in a regression), will block all backdoor paths.

```{r}
ggdag_adjustment_set(confounder_dag, shadow = TRUE) +
    theme_dag()
```

The plot correctly identifies `Z` as the variable we must adjust for. The `shadow = TRUE` argument is a nice way to make the adjustment set stand out.

#### 3. Verify the Adjustment with `ggdag_dseparated()`

Finally, how do we confirm our adjustment set works? We check for **d-separation**. Two variables are d-separated if there are no open paths between them. After we control for `Z`, `X` and `Y` should be d-separated (meaning all non-causal paths are blocked).

The `ggdag_dseparated()` function visualizes this, showing a dashed line if the variables are d-separated.

```{r}
# Test for d-separation while controlling for Z
ggdag_dseparated(confounder_dag, controlling_for = "Z") +
    theme_dag() +
    labs(title = "Controlling for Z blocks the backdoor path")
```

Success! The dashed line confirms that conditioning on `Z` closes the backdoor path, leaving only the causal effect of interest. This three-step workflow—`ggdag_paths`, `ggdag_adjustment_set`, and `ggdag_dseparated`—provides a rigorous and reproducible way to plan a causal analysis.

Now that you've seen the standard workflow, are you interested in how to apply this to a more complex scenario, like one involving mediation?