---
title: "Introduction to dagitty"
---

Below is a guide to the **R package `dagitty`** (current CRAN version 0.3-4), keeping the same structure but adding depth: mathematical rationale, plain-English assumptions, multiple examples (basic + advanced), scaffolding steps, and common pitfalls. Citations point to the official reference manual, vignettes, and peer-reviewed overviews of `dagitty`. [cran.r-project.org](https://cran.r-project.org/web/packages/dagitty/dagitty.pdf?utm_source=chatgpt.com)

# 1. Graph Construction & Manipulation

## 1.1 `dagitty()` — Parse DAGitty Syntax

**What it does.** Builds a graph object from a short textual spec; supports DAGs and ancestral graphs (MAGs). Use the optional `layout=TRUE` to auto-generate node coordinates. [RDocumentation](https://www.rdocumentation.org/packages/dagitty/versions/0.3-4/topics/dagitty?utm_source=chatgpt.com)

**Mathematical/causal rationale.** A DAG encodes a set of conditional independencies via **d-separation**. Edges `A -> B` represent direct causal effects (possibly with nonlinear/unknown functional form). In DAGitty syntax, bidirected edges `<->` stand for *latent* common causes (a shorthand for `A <- U -> B`). MAGs additionally allow undirected `--` edges for selection effects. [RDocumentation](https://www.rdocumentation.org/packages/dagitty/versions/0.3-4/topics/dagitty?utm_source=chatgpt.com)

**Basic example (scaffold).**

```{r}
pacman::p_load(dagitty)
```

```{r}
g <- dagitty("dag{ X -> Y; Z -> X }") # minimal confounding structure
exposures(g) <- "X"
outcomes(g) <- "Y" # mark roles for later tools
g
```

This sets up a DAG where $Z$ is a cause of $X$ (a potential confounder for $X→Y$). [Rdrr.io](https://rdrr.io/cran/dagitty/api/?utm_source=chatgpt.com)

**Advanced example (MAG with latent + selection).**s

```{r}
#| label: mag-example
g_mag <- dagitty("mag{ X <-> U <-> Y ; X -- S -- Y }")
graphType(g_mag) # "mag"
```

This MAG marks both latent confounding (`<->`) and selection (`--`). MAGs generalize DAGs for testable implications and equivalence. [RDocumentation](https://www.rdocumentation.org/packages/dagitty/versions/0.3-4/topics/graphType?utm_source=chatgpt.com)

### Pitfalls & caveats

-   Prefer a strictly acyclic structure (`isAcyclic(g)`) to use standard DAG rules; MAGs are for advanced use. [Rdrr.io](https://rdrr.io/cran/dagitty/api/?utm_source=chatgpt.com)
-   Always set `exposures()` and `outcomes()` before computing adjustment sets or IVs. [oai_citation:6‡cran.r-project.org](https://cran.r-project.org/web/packages/dagitty/refman/dagitty.html?utm_source=chatgpt.com)

## 1.2 `convert()` — Export to other Frameworks

**What it does.** Converts a DAGitty graph into formats for **lavaan**, **bnlearn**, **tikz**, or **causaleffect** packages. Useful for moving from qualitative design to model specification/estimation. [cran.r-project.org](https://cran.r-project.org/web/packages/dagitty/dagitty.pdf?utm_source=chatgpt.com)

**Example (to lavaan).**

```{r}
#| label: convert-to-lavaan
g <- dagitty("dag{ X -> Y ; Z -> X ; Z -> Y }")
model_string <- convert(g, to = "lavaan")
cat(model_string)
```

Feed `model_string` to `lavaan::sem()` if you want to fit a linear SEM (note: causal identification ≠ statistical identifiability; verify assumptions). [cran.r-project.org](https://cran.r-project.org/web/packages/dagitty/dagitty.pdf?utm_source=chatgpt.com)

## 1.3 `completeDAG()`

**What it does.** Creates a fully connected DAG over given nodes; handy for stress tests and teaching. [Rdrr.io](https://rdrr.io/cran/dagitty/man/?utm_source=chatgpt.com)

**Example.**

``` r
completeDAG(c("A", "B", "C"))
```

# 2. Graph Visualization & Layout

## 2.1 `graphLayout()` And `coordinates()`

**What they do.** `graphLayout()` assigns positions (spring layout). `coordinates()` gets/sets positions; coordinates can be stored inside the DAGitty string or assigned afterward. [RDocumentation](https://www.rdocumentation.org/packages/dagitty/versions/0.3-4/topics/graphLayout?utm_source=chatgpt.com)

**Scaffolded example.**

```{r}
#| label: graph-layout-example
#|
g <- dagitty("dag{ X->Y ; Z->X }")
g <- graphLayout(g, "spring") # auto layout
plot(g) # base plot method

coords <- coordinates(g) # retrieve
coordinates(g) <- coords # could modify and set back
```

For publication-grade plots in a tidyverse workflow, pair with **ggdag** (built on DAGitty). [cran.r-project.org](https://cran.r-project.org/web/packages/ggdag/vignettes/intro-to-ggdag.html?utm_source=chatgpt.com) and [PMC](https://pmc.ncbi.nlm.nih.gov/articles/PMC10869111/?utm_source=chatgpt.com)

**Pitfall.** The spring layout is stochastic; call repeatedly or set seeds if you need stable figures. [RDocumentation](https://www.rdocumentation.org/packages/dagitty/versions/0.3-4/topics/graphLayout?utm_source=chatgpt.com)

## 2.2 `plot.dagitty()`

**What it does.** Quick base plotting for DAGs/MAGs; respects stored coordinates. Use **ggdag** for more control and theming. [cran.r-project.org](https://cran.r-project.org/web/packages/dagitty/dagitty.pdf?utm_source=chatgpt.com)

# 3. Structural Queries & Relationships

## 3.1 Familial & Markov Relations

**What they do.** `parents()`, `children()`, `ancestors()`, `descendants()`, `spouses()`, `markovBlanket()`. These map local neighborhoods and influence sets. The Markov blanket of node $V$ (its parents, children, and the parents of its children) renders $V$ independent of all other nodes given the blanket. [oai_citation:15‡RDocumentation](https://www.rdocumentation.org/packages/dagitty/versions/0.3-4/topics/AncestralRelations?utm_source=chatgpt.com)

**Basic example.**

```{r}
g <- dagitty("dag{ Z->X->Y ; Z->Y; U->Z->Y; Y-> O ->K }")
plot(g)

cat('Ancestors of Y:\n')
ancestors(g, "Y")
cat('Descendants of Y:\n')
descendants(g, "Y")
cat('Children of Z:\n')
parents(g, "Y")
cat('Parents of Z:\n')
children(g, "Z")
cat('Markov blanket of X:\n')
markovBlanket(g, "X")
```

**Advanced example (reasoning).** For feature selection in prediction (not causal): conditioning on a node’s Markov blanket is sufficient (though not always minimal) to predict that node. Don’t confuse predictive sufficiency with causal sufficiency. [Rdrr.io](https://rdrr.io/cran/dagitty/man/AncestralRelations.html?utm_source=chatgpt.com)

**Common misconception.** “Ancestors/descendants exclude the node.” In DAGitty, `ancestors()` and `descendants()` include the node itself unless `proper=TRUE`. [RDocumentation](https://www.rdocumentation.org/packages/dagitty/versions/0.3-4/topics/AncestralRelations?utm_source=chatgpt.com)

## 3.2 `paths()`

**What it does.** Lists paths between exposure(s) and outcome(s) and whether they’re open (d-connected) or closed (d-separated) under a conditioning set $Z$. [RDocumentation](https://www.rdocumentation.org/packages/dagitty/versions/0.3-4/topics/paths?utm_source=chatgpt.com)

**Plain-English d-separation rule.** A path is **blocked** if (i) it contains a **non-collider** that is conditioned on, or (ii) it contains a **collider** (or a descendant of a collider) that is **not** conditioned on. Otherwise, it’s **open**. Identifying open backdoor paths is the core of confounding control. (DAGitty manual gives a succinct overview.) [DAGitty](https://www.dagitty.net/manual-3.x.pdf?utm_source=chatgpt.com)

**Basic example.**

```{r}
#| label: paths-example
g <- dagitty("dag{ Z->X->Y ; Z->Y }")
paths(g, from = "X", to = "Y") # backdoor X <- Z -> Y is open
paths(g, from = "X", to = "Y", Z = list("Z")) # conditioning on Z closes it
```

**Advanced example (collider bias).**

```{r}
#| label: collider-bias-example
g <- dagitty("dag{ X->Y ; X->M<-U->Y }") # M is collider on X-M-U-Y
paths(g, "X", "Y") # backdoor X <- U -> Y is closed by M
paths(g, "X", "Y", Z = list("M")) # conditioning on M OPENS it (collider bias)
```

Use `paths()` to *explain* why an adjustment set works (or doesn’t). [RDocumentation](https://www.rdocumentation.org/packages/dagitty/versions/0.3-4/topics/paths?utm_source=chatgpt.com)

# 4. Causal Analysis: Adjustment, Separation, Instruments

## 4.1 `adjustmentSets()` — Valid Covariates for Identification

**What it does.** Enumerates **valid** sets $Z$ that block all backdoor paths from $X$ to $Y$ (for the **total** effect), optionally returning minimal sets or a “canonical” superset. **Assumption:** the graph is correct (edges capture the true causal relations and measured/latent status). [search.r-project.org](https://search.r-project.org/CRAN/refmans/dagitty/html/adjustmentSets.html?utm_source=chatgpt.com)

**Backdoor identification formula (plain English).** If $Z$ satisfies the backdoor criterion, then the interventional mean is identifiable as

$$
\mathbb{E}[Y\!\mid\!\mathrm{do}(X=x)] \;=\; \sum_{z} \mathbb{E}[Y \mid X=x, Z=z]\, P(Z=z),
$$

or with integrals for continuous $Z$. In practice: regress $Y$ on $X$ and $Z$ (correct functional form) or use standardization/IPW consistent with the DAG. The *validity* of $Z$ is a **graphical** condition. [DAGitty](https://www.dagitty.net/manual-3.x.pdf?utm_source=chatgpt.com)

**Scaffolded examples.**

```{r}
g <- dagitty("dag{ Z->X->Y ; Z->Y }")
exposures(g) <- "X"
outcomes(g) <- "Y"
adjustmentSets(g, type = "minimal")
```

**Advanced (M-bias).** Adjusting for a *pre-treatment* variable can hurt when it is a collider or on a path that opens bias.

```{r}
g <- dagitty("dag{ X->Y ; X <-> M <-> Y }") # M encodes latent U via <-> shorthands
adjustmentSets(g, "X", "Y") # returns empty set; don't adjust for M
```

The reference manual shows simulation confirming bias inflation when wrongly adjusting. [oai_citation:23‡RDocumentation](https://www.rdocumentation.org/packages/dagitty/versions/0.3-4/topics/adjustmentSets?utm_source=chatgpt.com)

**Nuances & pitfalls.** - For **direct effects**, specify `effect="direct"`; valid sets then must also block **mediator** paths. Interpret carefully. [oai_citation:24‡search.r-project.org](https://search.r-project.org/CRAN/refmans/dagitty/html/adjustmentSets.html?utm_source=chatgpt.com) - Don’t auto-adjust for everything observed; that can open colliders or induce M-bias. Use the DAG. [oai_citation:25‡RDocumentation](https://www.rdocumentation.org/packages/dagitty/versions/0.3-4/topics/adjustmentSets?utm_source=chatgpt.com)

## 4.2 `instrumentalVariables()` — Detect Candidate IVs

**What it does.** Finds variables that satisfy graphical IV criteria relative to $(X,Y)$. **Assumptions (linear SEM context):** (i) relevance (IV→X), (ii) exclusion (no direct path IV→Y except via X), and (iii) independence from unmeasured causes of $Y$. Graphical discovery is heuristic and typically assumes linearity for certain algorithms. [oai_citation:26‡RDocumentation](https://www.rdocumentation.org/packages/dagitty/versions/0.3-4/topics/instrumentalVariables?utm_source=chatgpt.com)

**Basic example.**

```{r}
g <- dagitty("dag{ Z->X->Y ; Z<->U->Y }") # if Z has no path to Y except via X, it's an IV
exposures(g) <- "X"
outcomes(g) <- "Y"
instrumentalVariables(g)
```

**Advanced note.** A variable can be a *conditional* IV given covariates; you must verify no open backdoor from IV to $Y$ after conditioning. Practical estimation typically uses 2SLS; discovery here is *graph-based* only. [oai_citation:27‡RDocumentation](https://www.rdocumentation.org/packages/dagitty/versions/0.3-4/topics/instrumentalVariables?utm_source=chatgpt.com)

# 5. Testing & Model Implications

## 5.1 `impliedConditionalIndependencies()` — D-separation Claims

**What it does.** Lists CIs (e.g., $A \perp B \mid C$) that *must* hold for any distribution compatible with the graph. These implications can be confronted with data. [oai_citation:28‡RDocumentation](https://www.rdocumentation.org/packages/dagitty/versions/0.3-4/topics/impliedConditionalIndependencies?utm_source=chatgpt.com)

**Example.**

```{r}
# Use a graph with a non-trivial implied CI: Z ⟂⟂ Y | X
g <- dagitty("dag{ Z->X->Y }")
impliedConditionalIndependencies(g)
```

## 5.2 `localTests()` + `plotLocalTestResults()`

**What they do.** Derive testable implications from the graph and test them with data (partial correlations for continuous, appropriate associations for categorical); visualize effect sizes and CIs. **Caveat:** Passing tests does not *prove* the graph; failing tests falsifies it. [search.r-project.org](https://search.r-project.org/CRAN/refmans/dagitty/html/localTests.html?utm_source=chatgpt.com) [currentprotocols.onlinelibrary.wiley.com](https://currentprotocols.onlinelibrary.wiley.com/doi/pdf/10.1002/cpz1.45?utm_source=chatgpt.com)

**Scaffolded example (continuous).**

```{r}
#| label: local-tests-example
set.seed(1)
# Use a DAG with at least one implied CI (Z ⟂⟂ Y | X)
g <- dagitty("dag{ Z->X->Y }")
exposures(g) <- "X"
outcomes(g) <- "Y"
d <- simulateSEM(g, N = 1000) # quick toy data
res <- localTests(g, data = d) # tests graph-implied CIs (e.g., Z ⟂⟂ Y | X)
head(res)

# Guard against graphs with no testable implications
if (is.data.frame(res) && ncol(res) > 0 && nrow(res) > 0) {
  plotLocalTestResults(res)
} else {
  message("No testable implications for this graph; nothing to plot.")
}
```

**Advanced (categorical).** The protocol paper provides a workflow for DAG testing with categorical data (uses appropriate tests under the hood). [oai_citation:31‡currentprotocols.onlinelibrary.wiley.com](https://currentprotocols.onlinelibrary.wiley.com/doi/pdf/10.1002/cpz1.45?utm_source=chatgpt.com)

**Pitfalls.** - **Measurement error** can break expected CIs; consider latent measurement structure. [oai_citation:32‡cran.r-project.org](https://cran.r-project.org/web/packages/dagitty/dagitty.pdf?utm_source=chatgpt.com) - **Overconditioning** can induce dependencies (e.g., conditioning on a collider). Tests may “fail” because the analysis conditioned on the wrong set.

## 5.3 `vanishingTetrads()`

**What it does.** Lists **tetrad constraints** of linear SEMs implied by the graph:

$$
\operatorname{Cov}(i,j)\operatorname{Cov}(k,l) - \operatorname{Cov}(i,k)\operatorname{Cov}(j,l) \;=\; 0,
$$

useful for model equivalence checks. These hold in particular causal structures (no correlated residuals) and can be empirically checked. [oai_citation:33‡RDocumentation](https://www.rdocumentation.org/packages/dagitty/versions/0.3-4/topics/vanishingTetrads?utm_source=chatgpt.com)

**Example.**

```{r}
g <- dagitty("dag{ A->C ; B->C ; A->D ; B->D }") # classic tetrad pattern
vanishingTetrads(g)
```

# 6. Simulation Utilities

## 6.1 `simulateSEM()`

**What it does.** Treats the DAG as a *linear* SEM, draws random path coefficients (or use annotations), and simulates data; intended for validation/pedagogy rather than production simulations. [RDocumentation](https://www.rdocumentation.org/packages/dagitty/versions/0.3-4/topics/simulateSEM?utm_source=chatgpt.com) [search.r-project.org](https://search.r-project.org/CRAN/refmans/dagitty/html/simulateSEM.html?utm_source=chatgpt.com)

**Basic example.**

```{r}
g <- dagitty("dag{ Z->X->Y ; Z->Y }")
d <- simulateSEM(g, N = 1000)
```

**Advanced (annotated coefficients).**

```{r}
g <- dagitty("dag{ Z -> X [beta=0.7]; X -> Y [beta=0.5]; Z -> Y [beta=0.1] }")
d <- simulateSEM(g, N = 2000)
```

**Tip.** For tidyverse workflows, `ggdag::simulate_data()` wraps `simulateSEM()` and returns tidied outputs. [r-causal.github.io](https://r-causal.github.io/ggdag/reference/simulate_data.html?utm_source=chatgpt.com) [Rdrr.io](https://rdrr.io/cran/ggdag/man/simulate_data.html?utm_source=chatgpt.com)

## 6.2 `simulateLogistic()`

**What it does.** Simulates binary outcomes under logistic structural assumptions for quick experiments. (As with `simulateSEM()`, this is intentionally simple.) [Rdrr.io](https://rdrr.io/cran/dagitty/api/?utm_source=chatgpt.com)

------------------------------------------------------------------------

# 7. Graph Transformations & Checks

## 7.1 `canonicalize()` And `toMAG()`

**What they do.** `canonicalize()` expands ancestral graphs into DAGs by replacing bidirected and undirected edges with explicit latent/selection nodes; `toMAG()` computes an observationally equivalent MAG, respecting marginalization over latents. These tools clarify where bias can arise (latent confounding vs. selection). [cran.r-project.org](https://cran.r-project.org/web/packages/dagitty/refman/dagitty.html?utm_source=chatgpt.com) [oai_citation:40‡Rdrr.io](https://rdrr.io/cran/dagitty/api/?utm_source=chatgpt.com)

**Example.**

```{r}
#| label: fig-canonicalize-example
#| fig-cap: "DAG with a latent common causa L"
# Build a DAG with an unobserved common cause L of X and Y.
g_dag <- dagitty("dag{ X <- L -> Y ; Z -> Y ; X -> Y; L [latent] }")
plot(g_dag)
```

```{r}
#| label: canonicalize-example
# Convert DAG -> MAG by marginalizing L; result is a MAG with X <-> Y
g_mag <- toMAG(g_dag)
graphType(g_mag) # "mag"

# Try to canonicalize; some dagitty versions don't support canonicalize() for MAG
g_can <- tryCatch(
  canonicalize(g_mag),
  error = function(e) {
    message(
      "canonicalize() not supported for MAG in this dagitty version: ",
      e$message
    )
    return(g_mag)
  }
)
g_can
```

## 7.2 `isAcyclic()`, `isAdjustmentSet()`

**What they do.** Sanity checks—verify acyclicity and whether a proposed $Z$ truly satisfies the criterion (useful when you want to force a specific adjustment the algorithm didn’t return). [Rdrr.io](https://rdrr.io/cran/dagitty/api/?utm_source=chatgpt.com)

## 7.3 `backDoorGraph()`

**What it does.** Produces the graph used by the backdoor algorithm (removes arrows out of $X$, etc.)—useful for teaching/diagnostics. [cran.r-project.org](https://cran.r-project.org/web/packages/dagitty/dagitty.pdf?utm_source=chatgpt.com)

------------------------------------------------------------------------

# 8. Utility Functions & Examples

-   **`getExample()`** loads built-in classic DAGs (e.g., “Shrier”, “M-bias”) for demonstrations. [oai_citation:43‡Rdrr.io](https://rdrr.io/cran/dagitty/man/getExample.html?utm_source=chatgpt.com)\
-   **`downloadGraph()`** imports a DAG from **dagitty.net** by URL—handy when you sketch online, then continue in R. [oai_citation:44‡RDocumentation](https://www.rdocumentation.org/packages/dagitty/versions/0.3-4/topics/downloadGraph?utm_source=chatgpt.com) [oai_citation:45‡DAGitty](https://www.dagitty.net/?utm_source=chatgpt.com)\
-   **`edges()`** extracts edge lists (good for audits or exporting). [oai_citation:46‡RDocumentation](https://www.rdocumentation.org/packages/dagitty/versions/0.3-4/topics/edges?utm_source=chatgpt.com)\
-   **`graphType()`**, **`names.dagitty()`**, **`latents()`** inspect graph metadata and variable status (observed vs latent). [oai_citation:47‡RDocumentation](https://www.rdocumentation.org/packages/dagitty/versions/0.3-4/topics/graphType?utm_source=chatgpt.com) [oai_citation:48‡Rdrr.io](https://rdrr.io/cran/dagitty/api/?utm_source=chatgpt.com)\
-   **Tidy/plot ecosystem.** `ggdag` provides high-level plotting, path visualization, and wrappers for many DAGitty computations (keeps DAGitty as the analytic engine). [oai_citation:49‡cran.r-project.org](https://cran.r-project.org/web/packages/ggdag/vignettes/intro-to-ggdag.html?utm_source=chatgpt.com) [oai_citation:50‡r-causal.r-universe.dev](https://r-causal.r-universe.dev/ggdag/doc/manual.html?utm_source=chatgpt.com)

------------------------------------------------------------------------

## Summary Table (what to Reach for)

| Purpose | Functions | Key ideas & cautions |
|------------------------|------------------------|------------------------|
| Graph creation | `dagitty()`, `convert()`, `completeDAG()` | Encode causal knowledge first; set `exposures()`, `outcomes()` to prime later tools. [oai_citation:51‡RDocumentation](https://www.rdocumentation.org/packages/dagitty/versions/0.3-4/topics/dagitty?utm_source=chatgpt.com) [oai_citation:52‡cran.r-project.org](https://cran.r-project.org/web/packages/dagitty/dagitty.pdf?utm_source=chatgpt.com) |
| Plot/layout | `graphLayout()`, `coordinates()`, `plot()` | Layout is stochastic; store coordinates for reproducibility; consider `ggdag` for publication quality. [oai_citation:53‡RDocumentation](https://www.rdocumentation.org/packages/dagitty/versions/0.3-4/topics/graphLayout?utm_source=chatgpt.com) [oai_citation:54‡cran.r-project.org](https://cran.r-project.org/web/packages/ggdag/vignettes/intro-to-ggdag.html?utm_source=chatgpt.com) |
| Local structure | `parents()`, `children()`, `ancestors()`, `descendants()`, `markovBlanket()` | Markov blanket is predictive, not necessarily causal; `ancestors()`/`descendants()` include node unless `proper=TRUE`. [oai_citation:55‡RDocumentation](https://www.rdocumentation.org/packages/dagitty/versions/0.3-4/topics/AncestralRelations?utm_source=chatgpt.com) |
| Paths & bias | `paths()` | Diagnose which paths remain open under conditioning; avoid conditioning on colliders. [oai_citation:56‡RDocumentation](https://www.rdocumentation.org/packages/dagitty/versions/0.3-4/topics/paths?utm_source=chatgpt.com) [oai_citation:57‡DAGitty](https://www.dagitty.net/manual-3.x.pdf?utm_source=chatgpt.com) |
| Identification | `adjustmentSets()` | Minimal valid sets block *all* backdoors; avoid over-adjustment (M-bias). Direct vs total effects differ. [oai_citation:58‡search.r-project.org](https://search.r-project.org/CRAN/refmans/dagitty/html/adjustmentSets.html?utm_source=chatgpt.com) [oai_citation:59‡RDocumentation](https://www.rdocumentation.org/packages/dagitty/versions/0.3-4/topics/adjustmentSets?utm_source=chatgpt.com) |
| Instruments | `instrumentalVariables()` | Graphical IVs require relevance, exclusion, independence; mainly linear-SEM perspective for detection. [oai_citation:60‡RDocumentation](https://www.rdocumentation.org/packages/dagitty/versions/0.3-4/topics/instrumentalVariables?utm_source=chatgpt.com) |
| Test implications | `impliedConditionalIndependencies()`, `localTests()`, `plotLocalTestResults()` | Failing tests can falsify the DAG; passing does not prove it. Use appropriate tests for data type. [oai_citation:61‡RDocumentation](https://www.rdocumentation.org/packages/dagitty/versions/0.3-4/topics/impliedConditionalIndependencies?utm_source=chatgpt.com) [oai_citation:62‡search.r-project.org](https://search.r-project.org/CRAN/refmans/dagitty/html/localTests.html?utm_source=chatgpt.com) |
| Linear-SEM constraints | `vanishingTetrads()` | Tetrad equalities can distinguish models; assume no correlated errors unless modeled. [oai_citation:63‡RDocumentation](https://www.rdocumentation.org/packages/dagitty/versions/0.3-4/topics/vanishingTetrads?utm_source=chatgpt.com) |
| Simulation | `simulateSEM()`, `simulateLogistic()` | Quick didactic simulation; for complex studies prefer dedicated simulators. [oai_citation:64‡RDocumentation](https://www.rdocumentation.org/packages/dagitty/versions/0.3-4/topics/simulateSEM?utm_source=chatgpt.com) |
| Transform/check | `canonicalize()`, `toMAG()`, `isAcyclic()`, `isAdjustmentSet()` | Make latent/selection nodes explicit; verify assumptions programmatically. [oai_citation:65‡cran.r-project.org](https://cran.r-project.org/web/packages/dagitty/refman/dagitty.html?utm_source=chatgpt.com) |
| Import/export | `getExample()`, `downloadGraph()`, `edges()`, `graphType()` | Move between web UI and R; audit edges; keep graph type straight (DAG vs MAG). [oai_citation:66‡Rdrr.io](https://rdrr.io/cran/dagitty/man/getExample.html?utm_source=chatgpt.com) [oai_citation:67‡RDocumentation](https://www.rdocumentation.org/packages/dagitty/versions/0.3-4/topics/downloadGraph?utm_source=chatgpt.com) |

------------------------------------------------------------------------

## Closing Guidance

1)  Start by encoding domain knowledge in `dagitty()`, then **verify**: inspect paths, compute `adjustmentSets()`, and run `impliedConditionalIndependencies()`; if possible, confront implications with data via `localTests()`. 2) For teaching and quick validation, simulate with `simulateSEM()` to see bias appear/disappear as you (mis)adjust—this is a powerful scaffold for students. 3) For polished figures and tidy workflows, lean on **ggdag**, which delegates causal logic to DAGitty. [oai_citation:68‡cran.r-project.org](https://cran.r-project.org/web/packages/dagitty/dagitty.pdf?utm_source=chatgpt.com)
