---
title: "Tutorial: Using the emmeans R Package"
date: "1/6/2025"
date-modified: last-modified
date-format: long
---

# Emmeans Tutorial

## 1. Introduction

### Overview

`emmeans` (Estimated Marginal Means) is an R package designed for the analysis of linear models. It provides tools for obtaining and visualizing adjusted means for factors in statistical models, allowing users to interpret complex model outputs more easily.

### Use Cases

- **Post-hoc Comparisons:** Evaluating differences between group means after fitting a model.
- **Visualization:** Creating clear and informative plots of marginal means.
- **Model Checking:** Assessing the fit and assumptions of linear models.

### Benefits

- Simplifies the interpretation of model outputs.
- Facilitates the comparison of means across different factors.
- Integrates seamlessly with various types of models in R.

## 2. Getting Started

### Loading the Package

First, load the `emmeans` package into your R session:

```{r}
library(emmeans)
```

### Basic Usage

Here's a simple example using a linear model:

```{r}
# Sample data
data(mtcars)

# Fit a linear model
model <- lm(mpg ~ factor(cyl), data = mtcars)

# Obtain estimated marginal means
emm <- emmeans(model, ~cyl)
summary(emm)
tapply(mtcars$mpg, mtcars$cyl, mean)
```

*Annotation:* This will provide the estimated marginal means of miles per gallon (mpg) adjusted for the number of cylinders (cyl).

## 3. Key Functions and Features

### `emmeans()`

- **Description:** Computes estimated marginal means for specified factors in a model.
- **Syntax:**

```r
emmeans(object, specs, ...)
```

  - `object`: A fitted model object.
  - `specs`: Terms for which to compute marginal means.
- **Example:**

```{r}
# Example of emmeans
emm_cyl <- emmeans(model, ~cyl)
print(emm_cyl)
```

*This will display the estimated marginal means for each cylinder category.*

### `pairs()`

- **Description:** Computes pairwise comparisons among estimated marginal means.
- **Syntax:**

  ```r
  pairs(x, ...)
  ```

  - `x`: An object of class `emmeans`.
- **Example:**

```{r}
# Pairwise comparisons
pairs(emm_cyl)
```

*This gives pairwise comparisons between the cylinder levels.*

## 4. In-Depth Examples

### Title: Comparing MPG Across Cylinder Counts

#### Objective

To evaluate how the number of cylinders affects the miles per gallon (mpg) in cars.

#### Steps

1. **Data Preparation:**

```{r}
data(mtcars)
model <- lm(mpg ~ factor(cyl), data = mtcars)
```

2. **Function Application:**

```{r}
emm_cyl <- emmeans(model, ~cyl)
```

3. **Visualization:**

```{r}
plot(emm_cyl)
```

4. **Output:**
   The plot will show the estimated marginal means of mpg for each cylinder group, allowing for quick visual comparisons.

## 5. Advanced Features

### Customization

You can customize the output of `emmeans` using additional arguments:

```{r}
emm_cyl <- emmeans(model, ~cyl, at = list(cyl = c(4, 6, 8)))
```

### Integration

`emmeans` can be used with packages like `ggplot2` for enhanced visualizations:

```{r}
library(ggplot2)
emm_df <- as.data.frame(emm_cyl)
ggplot(emm_df, aes(x = cyl, y = emmean)) +
    geom_point() +
    geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL))
```

## 6. Troubleshooting and FAQs

### Common Issues

- **Error: Object not found**: Ensure the model object exists and is correctly specified.
- **Warnings about convergence**: Check the model fitting process for issues related to data or model complexity.

### FAQs

- **Q:** Can `emmeans` be used with non-linear models?
  - **A:** Yes, `emmeans` supports various model types, including generalized linear models.

## 7. Best Practices

- Use clear and descriptive variable names when fitting models.
- Regularly check model diagnostics to ensure assumptions are met.
- Document your analysis steps for reproducibility.

## 8. Conclusion

### Summary

In this tutorial, we explored the `emmeans` package, covering its key functions, in-depth examples, and advanced features.

### Next Steps

Try applying `emmeans` to your own datasets or explore more advanced statistical models.

### Encouragement

Dive into your data analysis projects with confidence, utilizing the power of `emmeans`!

## 9. References and Resources

- **Documentation:** [emmeans Documentation](https://cran.r-project.org/web/packages/emmeans/emmeans.pdf)
- **Learning Materials:** [R for Data Science](https://r4ds.had.co.nz/)
- **Support Channels:** [GitHub Issues](https://github.com/rvlenth/emmeans/issues), [Stack Overflow](https://stackoverflow.com/questions/tagged/emmeans)

```{mermaid}
graph TD;
    A[Start] --> B[Load Package];
    B --> C[Fit Model];
    C --> D[Compute EMM];
    D --> E[Visualize Results];
    E --> F[Interpret Output];
    F --> G[Explore Further];
    G --> H[End];
```

---

# Handling Interactions with `emmeans`

When working with linear models that include interaction terms, `emmeans` allows you to explore and visualize these interactions effectively. Here's how to manage interactions using the `emmeans` package.

## 1. Understanding Interactions

Interactions occur when the effect of one predictor variable on the response variable depends on the level of another predictor. For example, the relationship between treatment and outcome may vary depending on gender.

## 2. Fitting a Model with Interactions

### Example

Let's create a model with an interaction between two factors, `treatment` and `gender`, on the response variable `outcome`.

```{r}
# Sample data
data(mtcars)

# Create a categorical variable for the example
mtcars$am <- factor(mtcars$am, labels = c("Automatic", "Manual"))

# Fit a linear model with interaction
model <- lm(mpg ~ am * cyl, data = mtcars)
```

## 3. Using `emmeans` for Interactions

### Compute Estimated Marginal Means

To compute estimated marginal means for the interaction, use the `emmeans()` function with the interaction term specified.

```{r}
# Obtain estimated marginal means for the interaction
emm_interaction <- emmeans(model, ~ am * cyl)
summary(emm_interaction)
```

### Interpret the Output

The output will give you the estimated marginal means for each combination of the levels of `am` and `cyl`. This allows for examining how the means differ based on the interaction.

## 4. Visualizing Interactions

Visualizing interactions can help in understanding how the effects vary. Use the `plot()` function to create interaction plots.

```{r}
# Interaction plot
plot(emm_interaction)
```

### Customizing the Plot

You can further customize the plot using `ggplot2` for enhanced visual representation:

```{r}
library(ggplot2)

# Convert emmeans object to a data frame
emm_df <- as.data.frame(emm_interaction)

# Create a ggplot
ggplot(emm_df, aes(x = cyl, y = emmean, color = am)) +
    geom_point() +
    geom_line() +
    labs(
        title = "Interaction between Treatment and Cylinder",
        x = "Number of Cylinders",
        y = "Estimated Marginal Mean MPG"
    ) +
    theme_minimal()
```

## 5. Pairwise Comparisons

You can also perform pairwise comparisons for the interaction effects:

```{r}
# Pairwise comparisons for the interaction
pairs(emm_interaction)
```

This will give you insights into which combinations of `am` and `cyl` are significantly different from one another.

## Conclusion

Handling interactions in `emmeans` is straightforward and provides valuable insights into how different factors influence the response variable. By fitting models that include interactions, computing estimated marginal means, and visualizing the results, you can effectively interpret complex relationships in your data.

---

# Handling Interactions with More Than Two Factors in `emmeans`

When dealing with interactions involving multiple factors, `emmeans` can help you explore and visualize these complex relationships. Here's a guide on how to manage interactions with more than two factors using the `emmeans` package.

## 1. Understanding Multi-Factor Interactions

Multi-factor interactions occur when the effect of one factor on the response variable depends on the levels of two or more other factors. For example, you may want to investigate how the effect of treatment varies across different levels of gender and age group.

## 2. Fitting a Model with Multiple Interactions

### Example

Let's create a hypothetical dataset and fit a model that includes interactions among three factors: `treatment`, `gender`, and `age_group`.

```{r}
# Load necessary library
library(dplyr)

# Simulated dataset
set.seed(123)
data <- data.frame(
    treatment = factor(rep(c("A", "B"), each = 30)),
    gender = factor(rep(c("Male", "Female"), times = 30)),
    age_group = factor(rep(c("Young", "Old"), each = 15, times = 2)),
    outcome = rnorm(60, 50, 10)
)

# Fit a linear model with interactions
model <- lm(outcome ~ treatment * gender * age_group, data = data)
```

## 3. Using `emmeans` for Multi-Factor Interactions

### Compute Estimated Marginal Means

To compute estimated marginal means for the interaction among the three factors, specify the interaction in the `emmeans()` function.

```{r}
# Obtain estimated marginal means for the interaction
emm_multi <- emmeans(model, ~ treatment * gender * age_group)
summary(emm_multi)
```

### Interpret the Output

The output will display the estimated marginal means for each combination of the levels of `treatment`, `gender`, and `age_group`. This helps you understand how the outcome varies across these groups.

## 4. Visualizing Multi-Factor Interactions

Visualizing interactions with multiple factors can be complex, but it can be done effectively using `ggplot2`. Here's how to create a plot:

```{r}
library(ggplot2)

# Convert emmeans object to a data frame
emm_df <- as.data.frame(emm_multi)

# Create a ggplot
ggplot(emm_df, aes(x = treatment, y = emmean, group = interaction(gender, age_group), color = interaction(gender, age_group))) +
    geom_point(position = position_dodge(width = 0.5)) +
    geom_line(position = position_dodge(width = 0.5)) +
    labs(
        title = "Multi-Factor Interaction",
        x = "Treatment",
        y = "Estimated Marginal Mean Outcome"
    ) +
    theme_minimal()
```

### Customizing the Plot

You can adjust the aesthetics of the plot to improve clarity, such as adding `facet_wrap()` to separate plots by one of the factors:

```{r}
ggplot(emm_df, aes(x = treatment, y = emmean, color = gender)) +
    geom_point(position = position_dodge(width = 0.5)) +
    geom_line(position = position_dodge(width = 0.5), aes(group = age_group)) +
    facet_wrap(~age_group) +
    labs(
        title = "Multi-Factor Interaction by Age Group",
        x = "Treatment",
        y = "Estimated Marginal Mean Outcome"
    ) +
    theme_minimal()
```

## 5. Pairwise Comparisons

You can perform pairwise comparisons for the multi-factor interactions as well:

```{r}
# Pairwise comparisons for the multi-factor interaction
pairs(emm_multi)
```

This will provide insights into significant differences among the combinations of the three factors.

## Conclusion

Handling interactions with more than two factors in `emmeans` allows you to explore complex relationships in your data effectively. By fitting models that include multiple interactions, computing estimated marginal means, and visualizing the results, you can gain a deeper understanding of how different factors influence the response variable.

---

# Handling Interactions with More Than Two Factors in `emmeans`

When dealing with interactions involving multiple factors, `emmeans` can help you explore and visualize these complex relationships. Here's a guide on how to manage interactions with more than two factors using the `emmeans` package.

## 1. Understanding Multi-Factor Interactions

Multi-factor interactions occur when the effect of one factor on the response variable depends on the levels of two or more other factors. For example, you may want to investigate how the effect of treatment varies across different levels of gender and age group.

## 2. Fitting a Model with Multiple Interactions

### Example

Let's create a hypothetical dataset and fit a model that includes interactions among three factors: `treatment`, `gender`, and `age_group`.

```{r}
# Load necessary library
library(dplyr)

# Simulated dataset
set.seed(123)
data <- data.frame(
    treatment = factor(rep(c("A", "B"), each = 30)),
    gender = factor(rep(c("Male", "Female"), times = 30)),
    age_group = factor(rep(c("Young", "Old"), each = 15, times = 2)),
    outcome = rnorm(60, 50, 10)
)

# Fit a linear model with interactions
model <- lm(outcome ~ treatment * gender * age_group, data = data)
```

## 3. Using `emmeans` for Multi-Factor Interactions

### Compute Estimated Marginal Means

To compute estimated marginal means for the interaction among the three factors, specify the interaction in the `emmeans()` function.

```{r}
# Obtain estimated marginal means for the interaction
emm_multi <- emmeans(model, ~ treatment * gender * age_group)
summary(emm_multi)
```

### Interpret the Output

The output will display the estimated marginal means for each combination of the levels of `treatment`, `gender`, and `age_group`. This helps you understand how the outcome varies across these groups.

## 4. Visualizing Multi-Factor Interactions

Visualizing interactions with multiple factors can be complex, but it can be done effectively using `ggplot2`. Here's how to create a plot:

```{r}
library(ggplot2)

# Convert emmeans object to a data frame
emm_df <- as.data.frame(emm_multi)

# Create a ggplot
ggplot(emm_df, aes(x = treatment, y = emmean, group = interaction(gender, age_group), color = interaction(gender, age_group))) +
    geom_point(position = position_dodge(width = 0.5)) +
    geom_line(position = position_dodge(width = 0.5)) +
    labs(
        title = "Multi-Factor Interaction",
        x = "Treatment",
        y = "Estimated Marginal Mean Outcome"
    ) +
    theme_minimal()
```

### Customizing the Plot

You can adjust the aesthetics of the plot to improve clarity, such as adding `facet_wrap()` to separate plots by one of the factors:

```{r}
ggplot(emm_df, aes(x = treatment, y = emmean, color = gender)) +
    geom_point(position = position_dodge(width = 0.5)) +
    geom_line(position = position_dodge(width = 0.5), aes(group = age_group)) +
    facet_wrap(~age_group) +
    labs(
        title = "Multi-Factor Interaction by Age Group",
        x = "Treatment",
        y = "Estimated Marginal Mean Outcome"
    ) +
    theme_minimal()
```

## 5. Pairwise Comparisons

You can perform pairwise comparisons for the multi-factor interactions as well:

```{r}
# Pairwise comparisons for the multi-factor interaction
pairs(emm_multi)
```

This will provide insights into significant differences among the combinations of the three factors.

## Conclusion

Handling interactions with more than two factors in `emmeans` allows you to explore complex relationships in your data effectively. By fitting models that include multiple interactions, computing estimated marginal means, and visualizing the results, you can gain a deeper understanding of how different factors influence the response variable.