---
title: "Tutorial: Using the emmeans R Package"
date: "1/6/2025"
date-modified: last-modified
date-format: long
---

## 1. Introduction

The `emmeans` package in R simplifies post-hoc analysis and estimation of marginal means from statistical models. It provides tools to estimate, compare, and test means across levels of predictors while accounting for the model structure.

## Overview

- **Purpose**: Provides estimated marginal means (EMMs) or least-squares means.
- **Core Features**:
  - Compute EMMs for any combination of factors.
  - Perform pairwise comparisons or custom contrasts.
  - Integrate with popular visualization packages like `ggplot2`.
  - Easily handle model interactions and transformations.
- **Key Supported Models**:
  - Linear models (`lm`, `aov`).
  - Generalized linear models (`glm`).
  - Mixed-effects models (`lme4`, `nlme`).

## What Are Marginal Means?

Marginal means are predicted values from a model, adjusted to account for the distribution of covariates or other factors. They provide a clearer interpretation of group effects compared to raw means.

## 2. Getting Started

## Loading the Package

```{r}
library(emmeans)
```

## Example Dataset

We will use the `warpbreaks` dataset, which contains data on the number of breaks in yarn (response variable `breaks`) across two factors: wool type (`wool`) and tension level (`tension`).

## Basic Model

```{r}
# Fitting a linear model
lm_model <- lm(breaks ~ wool * tension, data = warpbreaks)
```

## Basic Marginal Means

```{r}
# Marginal means by wool
emmeans(lm_model, ~ wool)
```

## 3. Key Functions in `emmeans`

## 3.1 `emmeans()`

The `emmeans()` function is the foundation of the package, used to estimate marginal means.

**Usage**:

```r
emmeans(object, specs, by = NULL, options = NULL, ...)
```

**Arguments**:

- **`object`**: A fitted model (e.g., `lm`, `glm`, or `lmer`).
- **`specs`**: Formula specifying factors for which marginal means are computed. Use `|` to specify conditions (e.g., `~ wool | tension`).
- **`by`**: Variables to group results.
- **`options`**: Adjustments like type of prediction (`response`, `link`, etc.).

**Examples**:

```{r}
# Marginal means for wool
emmeans(lm_model, ~wool)

# Marginal means of tension within each wool group
emmeans(lm_model, ~ tension | wool)
```

## 3.2 `contrast()`

The `contrast()` function performs post-hoc comparisons between marginal means, such as pairwise comparisons or custom contrasts.

**Usage**:

```r
contrast(object, method = "pairwise", interaction = FALSE, by = NULL, ...)
```

**Arguments**:

- **`object`**: An `emmGrid` object (output of `emmeans()`).
- **`method`**: Type of comparison. Common methods:
	- `"pairwise"`: Pairwise differences.
	- `"poly"`: Polynomial contrasts.
	- `"eff"`: Effect contrasts (deviation from overall mean).
- **`interaction`**: Defines interaction contrasts, generating contrasts of contrasts for multi-factor interactions.
- **`by`**: Grouping variables for separate contrasts.

**Examples**:

```{r}
# Pairwise comparisons of tension levels
warp.emm <- emmeans(lm_model, ~ tension | wool)
contrast(warp.emm, method = "pairwise")

# Custom contrasts for tension levels
contrast(warp.emm, list(mid.vs.ends = c(-1, 2, -1) / 2, lo.vs.hi = c(1, 0, -1)))
```

## 3.3 `pairs()`

This function simplifies pairwise comparisons (shortcut for `contrast(…, method = "pairwise")`).

**Usage**:

```r
pairs(object, reverse = FALSE, ...)
```

**Arguments**:

- **`reverse`**: If `TRUE`, reverses the order of comparisons.

**Example**:

```{r}
# Pairwise comparisons for tension
pairs(warp.emm)
```

## 4. Advanced Functionality

## 4.1 Interaction Contrasts

Interaction contrasts allow you to compare interactions between factors. The `interaction` argument in `contrast()` enables this.

**Example**:

```{r}
# Interaction contrast for wool and tension
interaction_contrast <- contrast(warp.emm, interaction = c(wool = "eff", tension = "poly"))
interaction_contrast
coef(interaction_contrast)  # View coefficients
```

## 4.2 Scaling and Offsetting

Use `scale` and `offset` to transform results for interpretation (e.g., converting units).

**Example**:

```{r}
# Convert temperature from Celsius to Fahrenheit
mod <- lm(Water.Temp ~ poly(stack.loss, degree = 2), data = stackloss)
emm <- emmeans(mod, "stack.loss", at = list(stack.loss = 10 * (1:4)))
contrast(emm, "identity", scale = 9 / 5, offset = 32)
```

## 4.3 Adjusting for Multiple Comparisons

Use the `adjust` argument to control corrections like Bonferroni or Tukey:

```{r}
contrast(warp.emm, method = "pairwise", adjust = "tukey")
```

## 5. Visualizing Results

## Interaction Plots

The `emmip()` function creates interaction plots to visualize EMMs:

```{r}
# Interaction plot for wool and tension
library(ggplot2)
emmip(warp.emm, wool ~ tension)
```

## Pairwise Comparisons Plot

```{r}
# Pairwise comparison plot
plot(pairs(warp.emm))
```

## 6. Common Issues and Troubleshooting

## Common Issues

1. **Degrees of freedom not available**:

	- Some models (e.g., Bayesian) require manual specification of degrees of freedom.
2. **Invalid contrast method**:

	- Ensure the `method` matches the structure of your `emmGrid` object.

## Best Practices

- Always inspect the `emmGrid` object using `summary()` before applying contrasts.
- Use `confint()` to obtain confidence intervals.
- Apply `adjust` for proper corrections in multiple comparisons.

## 7. Detailed Argument Reference

## `emmeans()` Specific Arguments

- **`specs`**: Defines the terms of interest. For example:
	- `~ factor1`: Marginal means for `factor1`.
	- `~ factor1 | factor2`: Marginal means of `factor1` grouped by `factor2`.

## `contrast()` Specific Arguments

- **`interaction`**: Defines contrasts for interactions. Example:

```{r}
contrast(warp.emm, interaction = list(wool = "pairwise", tension = "eff"))
```

- **`scale` and `offset`**: For transforming estimates. Example:

```{r}
contrast(warp.emm, method = "identity", scale = 2, offset = 1)
```
## 8. Conclusion

The `emmeans` package provides a robust framework for estimating and interpreting marginal means. With extensive support for complex models and versatile post-hoc analysis tools, it is essential for data analysis workflows.

## Next Steps

- Explore advanced vignettes in the package documentation.
- Integrate results into visualizations with `ggplot2`.

## 9. References and Resources

- [emmeans CRAN Page](https://cran.r-project.org/package=emmeans)
- [Vignettes: Estimated Marginal Means](https://cran.r-project.org/web/packages/emmeans/vignettes/emmeans.html)
- [GitHub Repository](https://github.com/rvlenth/emmeans)