---
title: "ggdag: A Tutorial for Causal DAGs in R"
execute: 
  echo: true
---

# Ggdag Tutorial with Function and Argument Explanations

This revision expands each function’s description with clear argument notes and runnable examples using base pipes (`|>`). Installation steps are omitted as requested.

# 1. Introduction

ggdag is a tidy, ggplot2-based interface to dagitty for drawing and diagnosing Directed Acyclic Graphs (DAGs). You define a DAG once, then reuse it to check identification (backdoor/frontdoor, adjustment sets, d-separation) and to produce consistent figures.

Key functions covered

- Create and prepare DAGs: `dagify()`, `dagitty()`, `ggdag::tidy_dagitty()`
- Plot scaffolds and theming: `ggdag::ggdag()`, `ggdag::ggdag_classic()`, `ggdag::theme_dag()`
- Roles and node status: `ggdag::node_status()`, `ggdag::ggdag_status()`
- Paths and d-separation: `ggdag::ggdag_paths()`, `ggdag::ggdag_dseparated()`
- Identification helpers: `ggdag::dag_adjustment_sets()`, `ggdag::ggdag_adjustment_set()`
- Structure inspection: `ggdag::ggdag_adjustment_set()`, `ggdag::ggdag_collider()`, `ggdag::ggdag_equivalent_dags()`
- Low-level geoms: `ggdag::geom_dag_point()`, `ggdag::geom_dag_edges()`, `ggdag::geom_dag_text()`

Typical use cases

- Teaching causal diagrams with publication-quality plots
- Rapid checks of valid adjustment sets and d-separation
- Documenting modeling assumptions alongside code

# 2. Getting Started

```{r}
#| label: setup-packages
pacman::p_load(ggdag, dagitty, dplyr, ggplot2)
```

```{r}
#| label: basic-workflow
# Define a simple confounding DAG: Z -> X -> Y and Z -> Y
dag <- dagify(
  Y ~ X + Z,
  X ~ Z,
  exposure = "X",
  outcome = "Y"
)

td <- ggdag::tidy_dagitty(dag, layout = "nicely", seed = 123)

td |>
  ggdag::ggdag_classic(text = FALSE) +
  ggdag::geom_dag_text() +
  ggdag::theme_dag()
```

# 3. Key Functions and Features (with Arguments)

## 3.1 dagify()

Purpose

Programmatically define a DAG with formulas and set exposure/outcome/latent nodes.

Syntax (no data example)

```r
dagify(…, exposure = NULL, outcome = NULL, latent = NULL, coords = NULL)
```

Argument notes

- `…`: One or more one-sided formulas `child ~ parents`. Example: `Y ~ X + Z` means `X → Y`, `Z → Y`.
- `exposure`, `outcome`: Character scalars naming the exposure and outcome. Setting these enables identification helpers downstream.
- `latent`: Character vector of unobserved nodes. These are retained in the graph but marked latent.
- `coords`: Optional named list or data frame of node coordinates to fix positions; otherwise layouts are computed later.

Runnable example

```{r}
#| label: dagify-example
dag2 <- dagify(
  Y ~ X + Z + U,
  X ~ Z + U,
  exposure = "X",
  outcome = "Y",
  latent = "U"
)
dag2
```

## 3.2 ggdag::tidy_dagitty()

Purpose

Convert a `dagitty` DAG into tidy node/edge data for plotting and analysis.

Syntax

```r
ggdag::tidy_dagitty(x, layout = "nicely", seed = NULL, …)
```

Argument notes

- `x`: A `dagitty` object.
- `layout`: Layout algorithm name (forwarded to ggraph/igraph), e.g., `"nicely"`, `"circle"`, `"tree"`, `"kk"`, `"fr"`.
- `seed`: Optional integer to make the layout reproducible.
- `…`: Passed to the underlying layout creation; use sparingly unless you know ggraph’s layout args.

Runnable example

```{r}
#| label: tidy-example
td2 <- ggdag::tidy_dagitty(dag2, layout = "nicely", seed = 1)
dplyr::glimpse(td2$data) # nodes
dplyr::glimpse(td2$edges) # edges
```

Concept

Returns a list-like tidy DAG object with `data` (nodes tibble) and `edges` (edges tibble), plus the original `dag`.

## 3.3 ggdag::ggdag() And ggdag::ggdag_classic()

Purpose

Create a ggplot scaffold for DAGs with sensible defaults; `ggdag_classic()` uses a cleaner minimal theme.

Syntax

```r
ggdag(.data, text = TRUE, use_labels = TRUE, label = "name", …)
ggdag_classic(.data, text = TRUE, use_labels = TRUE, label = "name", …)
```

Argument notes

- `.data`: Output of `tidy_dagitty()`.
- `text`: Logical; if `TRUE`, labels are added automatically.
- `use_labels`: If `TRUE`, prefer node labels (if present) over names.
- `label`: Column used for labels; default `"name"`.
- `…`: Additional aesthetics or parameters forwarded to internal geoms/theme.

Runnable example

```{r}
#| label: ggdag-classic-example
td2 |>
  ggdag::ggdag_classic(text = FALSE) +
  ggdag::geom_dag_point(size = 4) +
  ggdag::geom_dag_text(aes(label = name)) +
  ggdag::theme_dag()
```

## 3.4 ggdag::node_status() And ggdag::ggdag_status()

Purpose

Compute and visualize node roles: exposure, outcome, adjusted, latent, and other.

Syntax

```r
node_status(.data, …)
ggdag_status(.data, …)
```

Argument notes

- `.data`: A `tidy_dagitty` or `dagitty` object.
- `…`: Currently unused, reserved for future options.

Runnable example

```{r}
#| label: node-status-example
# node_status() must be called on the tidy_dagitty object, not inside mutate().
# It returns a tidy_dagitty with a `status` column that ggdag_status() will plot.

td2 |>
  ggdag::node_status() |>
  ggdag::ggdag_status()
```

## 3.5 ggdag::ggdag_paths()

Purpose

Highlight and inspect paths between two nodes.

Syntax

```r
ggdag_paths(.data, from, to, directed = FALSE, …)
```

Argument notes

- `.data`: Tidy DAG.
- `from`, `to`: Character scalars naming the endpoints.
- `directed`: If `TRUE`, limit to directed paths; if `FALSE`, show all paths (useful for backdoor inspection).
- `…`: Plot customizations passed to geoms.

Runnable example

```{r}
#| label: paths-example
ggdag::ggdag_paths(td2, from = "X", to = "Y")
```

## 3.6 ggdag::ggdag_dseparated()

Purpose

Visualize whether two nodes are d-separated given a conditioning set.

Syntax

```
ggdag_dseparated(.data, from, to, controlling_for = NULL, …)
```

Argument notes

- `.data`: Tidy DAG.
- `from`, `to`: Endpoints as character scalars.
- `controlling_for`: Character vector of nodes to condition on.
- `…`: Aesthetic overrides.

Runnable example

```{r}
#| label: dsep-example
ggdag::ggdag_dseparated(td2, from = "X", to = "Y")
ggdag::ggdag_dseparated(td2, from = "X", to = "Y", controlling_for = "Z")
```

Interpretation

Open paths appear highlighted; conditioning may open/close paths depending on colliders.

## 3.7 ggdag::dag_adjustment_sets() and ggdag::ggdag_adjustment_set()

Purpose

Compute and plot valid covariate sets that identify the total causal effect via backdoor adjustment.

Syntax

```r
dag_adjustment_sets(.tdy_dag, exposure = NULL, outcome = NULL, ...)
ggdag_adjustment_set(
  .tdy_dag,
  exposure = NULL,
  outcome = NULL,
  ...,
  shadow = TRUE
)
```

Argument notes

- `.data`: Tidy DAG with exposure/outcome known.
- `exposure`, `outcome`: Optional overrides; if omitted, use the dag’s attributes.
- `...`: Additional arguments passed to `dagitty::adjustmentSets()`.
- `shadow`: If `TRUE`, de-emphasize non-selected nodes/edges.
- `…`: Plot tweaks.

Runnable example

```{r}
#| label: adjust-sets-example
# Use the simple confounding DAG (no latent U) from the basic workflow
# so that valid backdoor sets exist and no warnings are emitted.
ggdag::dag_adjustment_sets(td)
ggdag::ggdag_adjustment_set(td, shadow = TRUE)
```

```{r}
#| label: adjust-sets-latent-example
# If a latent confounder exists (e.g., td2 has latent U), there may be
# NO valid backdoor adjustment set. This is expected; suppress warnings
# in rendered docs and explain in text.
suppressWarnings(ggdag::dag_adjustment_sets(td2))
```

Notes

- Under the hood, uses dagitty’s `adjustmentSets()` to enumerate valid sets.
- If the graph contains unblocked latent confounding between exposure and outcome (e.g., `X <- U -> Y` with `U` latent), no adjustment set exists. ggdag will warn that backdoor paths cannot be closed; this is by design.

## 3.8 Frontdoor Visualization (manual)

Purpose

Illustrate frontdoor structure when your ggdag version does not provide a dedicated helper.

Approach

- Use `ggdag_paths()` to highlight the directed path `X → M → Y`.
- Then plot the DAG to discuss the required assumptions (e.g., unblocked `X ← U → Y` but no unblocked backdoor from `X` to `M`, and backdoors from `M` to `Y` blocked by `X`).

Runnable example

```{r}
#| label: backdoor-frontdoor-example
# Visualize a valid backdoor adjustment set (if one exists)
ggdag::ggdag_adjustment_set(td, shadow = TRUE)
```

Frontdoor illustration (with latent U):

```{r}
#| label: frontdoor-example
dagC <- dagify(
  Y ~ M + U,
  M ~ X,
  X ~ U,
  exposure = "X",
  outcome = "Y",
  latent = "U"
)
tdC <- ggdag::tidy_dagitty(dagC, seed = 2025)

# Highlight the mediator path X -> M -> Y
ggdag::ggdag_paths(tdC, from = "X", to = "Y", directed = TRUE)
```

## 3.9 ggdag::ggdag_adjustment_set() And ggdag::ggdag_collider()

Purpose

Highlight valid adjustment sets for backdoor identification and visualize colliders.

Syntax

```r
ggdag_adjustment_set(.data, exposure = NULL, outcome = NULL, …)
ggdag_collider(.data, …)
```

Argument notes

- `.data`: Tidy DAG.
- `exposure`, `outcome`: Optional overrides; otherwise use dag attributes.
- `…`: Plot tweaks.

Runnable example

```{r}
#| label: confounder-collider-example
ggdag::ggdag_adjustment_set(td, shadow = TRUE)
ggdag::ggdag_collider(td2)
```

## 3.10 ggdag::ggdag_equivalent_dags()

Purpose

Enumerate and facet Markov equivalent DAGs (same conditional independences).

Syntax

```r
ggdag_equivalent_dags(.data, …)
```

Argument notes

- `.data`: Tidy DAG.
- `…`: Facet/plot options.

Runnable example

```{r}
#| label: equivalent-dags-example
ggdag::ggdag_equivalent_dags(td2)
```

## 3.11 Low-level Geoms and Theme

Purpose

Compose custom DAG plots with fine control.

Syntax

```r
geom_dag_point(…)
geom_dag_edges(arrow_directed = FALSE, arrow_size = 0.25, edge_width = 0.5, …)
geom_dag_text(mapping = aes(label = name), …)
theme_dag()
```

Argument notes

- `geom_dag_point(…)`: Accepts standard ggplot aesthetics (`size`, `alpha`, `shape`, etc.).
- `geom_dag_edges(…)`: Commonly used parameters include
  `arrow_directed` (logical to draw arrows),
  `arrow_size` (numeric, arrow head size),
  `edge_width` (numeric line width),
  plus standard aesthetics.
- `geom_dag_text(…)`: Default label is `name`; can map other columns to `label`. Accepts `size`, `hjust`, `vjust`, etc.
- `theme_dag()`: A minimalist theme tailored for DAG readability.

Runnable example

```{r}
#| label: geoms-theme-example
td2 |>
  ggdag::ggdag(text = FALSE) +
  ggdag::geom_dag_point(size = 4) +
  ggdag::geom_dag_text(aes(label = name)) +
  ggdag::geom_dag_edges(edge_width = 0.9) +
  ggdag::theme_dag()
```

# 4. Common Workflows with Interpretation

## Basic: Validate Adjustment for a Confounder

Objective

Confirm that adjusting for Z blocks backdoor paths from X to Y.

Steps

Define confounding DAG, compute adjustment sets, compare d-separation with and without Z.

```{r}
#| label: workflow-basic-adjust
dagA <- dagify(
  Y ~ X + Z,
  X ~ Z,
  exposure = "X",
  outcome = "Y"
)
tdA <- ggdag::tidy_dagitty(dagA, seed = 42)

# List valid adjustment sets for X -> Y (tidy tibble)
setsA <- ggdag::dag_adjustment_sets(tdA)
setsA
# Visualize a valid set
ggdag::ggdag_adjustment_set(tdA, shadow = TRUE)
ggdag::ggdag_dseparated(tdA, from = "X", to = "Y") # not separated
ggdag::ggdag_dseparated(tdA, from = "X", to = "Y", controlling_for = "Z") # separated when conditioning on Z
```

Interpretation

`ggdag_adjustment_sets()` highlights Z as a valid set; the d-separation plot flips once we condition on Z.

## Intermediate: Avoid Collider Adjustment

Objective

Show that conditioning on M in `X → M ← Z` opens a path.

```{r}
#| label: workflow-collider
dagB <- dagify(
  Y ~ X + Z + M,
  M ~ X + Z,
  exposure = "X",
  outcome = "Y"
)
tdB <- ggdag::tidy_dagitty(dagB, seed = 1)

ggdag::ggdag_collider(tdB)
ggdag::ggdag_dseparated(tdB, from = "X", to = "Z") # blocked by collider M
ggdag::ggdag_dseparated(tdB, from = "X", to = "Z", controlling_for = "M") # opens when conditioning on M
```

Interpretation

Conditioning on colliders induces association between otherwise independent parents.

## Advanced: Front-door Visualization

Objective

Visualize front-door identification with observed mediator M and unobserved confounder U.

```{r}
#| label: workflow-frontdoor
dagC <- dagify(
  Y ~ M + U,
  M ~ X,
  X ~ U,
  exposure = "X",
  outcome = "Y",
  latent = "U"
)
tdC <- ggdag::tidy_dagitty(dagC, seed = 2025)

# Highlight mediator path and all paths for discussion
ggdag::ggdag_paths(tdC, from = "X", to = "Y", directed = TRUE)
ggdag::ggdag_paths(tdC, from = "X", to = "Y", directed = FALSE)
```

Interpretation

Front-door plots emphasize M’s role in identifying the effect despite unblocked confounding between X and Y.

# 5. Advanced Features and Customization

- Stable layouts: set `seed` in `tidy_dagitty()` or pass `coords` to `dagify()`.
- Aesthetics: customize via geoms (`size`, `alpha`, `edge_width`).
  Note: Older ggdag versions may not support arrow parameters; if you
  encounter an "invalid 'arrow' argument" error, omit arrow options or
  pass a proper `grid::arrow()` object directly when supported.
- Programmatic relabeling: join a node tibble on `name` and map `aes(color = group, shape = status)` after adding `ggdag::node_status()`.

```{r}
#| label: customization-example
td2 |>
  ggdag::node_status() |>
  ggdag::ggdag(text = FALSE) +
  ggdag::geom_dag_point(aes(shape = status), size = 4, na.rm = TRUE) +
  ggdag::geom_dag_text(aes(label = name), na.rm = TRUE) +
  ggdag::geom_dag_edges(edge_width = 0.8, na.rm = TRUE) +
  ggplot2::labs(title = "Custom DAG with latent node shape") +
  ggdag::theme_dag()
```

# 6. Troubleshooting and FAQs

Common issues

- Empty adjustment sets
  Ensure `exposure` and `outcome` are set in the dagitty object, or pass them explicitly to the ggdag function.
- Layout changes every run
  Provide `seed` to `tidy_dagitty()` or supply fixed `coords` in `dagify()`.
- Labels overlap
  Turn off automatic `text` in the scaffold and add `geom_dag_text()` with adjusted `size`; consider jittering coordinates slightly.

FAQs

- How do I select a specific adjustment set?
  Use `ggdag_adjustment_set(td, set_index = k)`.
- Can I facet equivalent DAGs?
  Yes, `ggdag_equivalent_dags()` returns a facetted plot of Markov equivalent structures.

# 7. Best Practices

- Always set exposure and outcome on DAG creation to activate identification helpers.
- Use `ggdag_paths()` and `ggdag_dseparated()` to inspect why a set is valid or invalid.
- Keep DAGs minimal and question-focused; extra nodes add visual noise and can obscure identification logic.
- Make plots reproducible with a deterministic layout (`seed` or explicit coordinates).

# 8. Conclusion

With ggdag, you write your DAG once and reuse it to reason about identification and to create consistent, high-quality figures. Knowing each function’s arguments helps you control both the logic (d-separation, adjustment) and the presentation (geoms, themes, and layouts).

# 9. References and Resources

- ggdag package reference and vignettes
- dagitty documentation and theory notes
- ggplot2 reference for customization and export
