---
title: "Tutorial for the ibd Package"
date: "3/15/2025"
date-modified: last-modified
date-format: long
---

# A Comprehensive Tutorial on the R Package 'ibd' for Incomplete Block Designs

The `ibd` package provides tools for creating and analyzing **Incomplete Block Designs** (IBDs) in R, facilitating efficient experimentation when it's not possible to include all treatments in each experimental block.

## Introduction

### Overview

The 'ibd' R package (version 1.5) is designed for working with Incomplete Block Designs - experimental arrangements where not all treatments appear in every block[2]. This is particularly useful in situations where testing all treatments in a single block is impractical due to resource constraints or experimental limitations.

Key functions in the package include:

- `find_design()` - For discovering efficient IBD designs
- `bibd()` - For creating Balanced Incomplete Block Designs
- `pbibd()` - For creating Partially Balanced Incomplete Block Designs
- `design_efficiency()` - For evaluating the statistical efficiency of designs
- `analyze_ibd()` - For analyzing data collected from incomplete block experiments

### Use Cases

- Agricultural field trials with limited land or resources
- Clinical trials where patient availability or grouping is constrained
- Industrial experiments where testing all treatments in each batch is unfeasible
- Plant breeding and genomic selection studies where not all lines can be tested in all locations[8]
- Any experimental scenario where resource constraints prevent complete blocking

### Benefits

- Enables efficient experimentation with a large number of treatments
- Provides statistically rigorous designs when complete blocking is impossible
- Optimizes experimental efficiency by balancing treatment comparisons
- Reduces experimental costs while maintaining statistical power
- Facilitates the analysis of data collected from incomplete designs[8]

## Getting Started

### Loading the Package

```{r}
pacman::p_load(ibd)
```

### Basic Usage

Let's create a simple balanced incomplete block design for an experiment with 7 treatments in blocks of size 3:

```{r}
# Create a balanced incomplete block design
design <- ibd::bibd(v = 7, b = 3, k = 3, r = 3, lambda = 1)

# View the design structure
print(design)

# Check the properties of the design
summary(design)
```

This creates a design where each treatment appears in 3 blocks, and each pair of treatments appears together in exactly 1 block, ensuring balance in treatment comparisons.
