---
title: "Tutorial: Using the multcomp R Package"
date: "1/6/2025"
date-modified: last-modified
date-format: long
---

## 1. Introduction

### Overview

The `multcomp` package in R is designed for simultaneous inference in linear models, specifically focused on multiple comparisons. It provides tools to test hypotheses about multiple parameters in a model while controlling for family-wise error rates.

### Use Cases

- Comparing group means in ANOVA.
- Simultaneous confidence intervals.
- Adjusting p-values for multiple testing.

### Benefits

- Handles a variety of multiple comparison tests.
- Provides flexibility with user-defined contrasts.
- Supports many types of linear models, including generalized linear models and mixed-effects models.

## 2. Getting Started

### Loading the Package

```{r}
# Load the multcomp package
pacman::p_load(multcomp)
```

### Basic Usage

Let's start with a simple example:

```{r}
# Example dataset
data(warpbreaks)

# Fit a linear model
model <- lm(breaks ~ wool + tension, data = warpbreaks)

# Perform multiple comparisons
comparison <- glht(model, linfct = mcp(tension = "Tukey"))

# Summarize the results
summary(comparison)
```

#### Explanation

1. **`mcp`**: Specifies the type of multiple comparisons (e.g., Tukey's test).
2. **`glht`**: General linear hypothesis testing function.
3. **`summary`**: Displays adjusted p-values and confidence intervals.

## 3. Key Functions and Features

### `glht()`

**Description**: Performs multiple comparisons or general linear hypothesis tests.

**Syntax**:

```r
multcomp::glht(model, linfct = mcp("factor"="method"))
```

**Arguments**:

- `model`: A fitted linear model object.
- `linfct`: Defines the hypotheses to test (e.g., Tukey or user-defined contrasts).

**Example**:

```{r}
comparison <- glht(model, linfct = mcp(tension = "Tukey"))
summary(comparison)
```

### `mcp()`

**Description**: Creates predefined contrasts for a factor.

**Syntax**:

```r
mcp(factor = "method")
```

**Example**:

```r
linfct = mcp(tension = "Tukey")
```

## 4. In-Depth Examples

### Example 1: ANOVA Multiple Comparisons

**Objective**: Compare tension levels in `warpbreaks` dataset.

**Steps**:

1. **Fit the model**:

```{r}
model <- aov(breaks ~ tension, data = warpbreaks)
```

2. **Perform Tukey's test**:

```{r}
comparison <- glht(model, linfct = mcp(tension = "Tukey"))
summary(comparison)
```

3. **Visualize the results**:

```{r}
plot(comparison)
```

**Output**: Provides adjusted p-values and confidence intervals.

## 5. Advanced Features

### Custom Contrasts

Create user-defined contrasts:

```{r}
contrast_matrix <- rbind(
    "M - L" = c(-1, 1, 0),
    "H - L" = c(-1, 0, 1),
    "H - M" = c(0, -1, 1)
)
comparison <- glht(model, linfct = contrast_matrix)
summary(comparison)
```

### Integration with Mixed-Effects Models

Supports models from `lme4`:

```{r}
library(lme4)
model <- lmer(breaks ~ tension + (1 | wool), data = warpbreaks)
comparison <- glht(model, linfct = mcp(tension = "Tukey"))
summary(comparison)
```

## 6. Troubleshooting and FAQs

### Common Issues

1. **Error: Non-conformable arguments**

	- Ensure the `linfct` matches the model's factor levels.
2. **Confusion with model types**

	- Use `glht` with models that support linear hypotheses (e.g., `lm`, `lmer`).

### FAQs

- **Can `multcomp` handle unbalanced data?**
	Yes, but ensure your model accounts for this.

## 7. Best Practices

- Always check model diagnostics before multiple comparisons.
- Use visualizations like `plot()` to complement numerical results.
- For large datasets, consider summarizing key results to avoid overloading the output.

## 8. Conclusion

### Summary

The `multcomp` package is a robust tool for conducting multiple comparisons in R, providing flexibility and accuracy.

### Next Steps

Explore interactions in models and extend functionality with packages like `emmeans`.

### Encouragement

Experiment with different datasets to build confidence in interpreting outputs!

## 9. References and Resources

- [Package Documentation](https://cran.r-project.org/web/packages/multcomp/multcomp.pdf)
- [Examples on CRAN](https://cran.r-project.org/package=multcomp)

