---
title: "Tutorial for the purrr Package"
date: "1/6/2025"
date-modified: last-modified
date-format: long
---

# Tutorial for the `purrr` Package

## Introduction

### Overview

The `purrr` package, part of the `tidyverse`, provides a set of tools for functional programming in R. It simplifies iteration, mapping, and manipulation of lists, vectors, and other data structures. The functions in `purrr` are designed to replace and enhance base R functions like `lapply()` and `sapply()` with more readable and consistent alternatives.

Key components:

- **Core Functions**: `map()`, `map2()`, `pmap()`, `reduce()`, `pluck()`, and others.
- **Helpers**: `safely()`, `possibly()`, `transpose()`, etc.

### Use Cases

- Applying a function to each element of a list or vector.
- Handling nested data structures like lists of data frames.
- Aggregating results using custom operations.
- Safely working with functions that might return errors.

### Benefits

- Consistent syntax and predictable outputs.
- Seamless integration with other tidyverse packages.
- Tools for error handling and debugging.

---

## Getting Started

### Loading the Package

Use the `pacman` package to ensure `purrr` is loaded:

```{r}
pacman::p_load(purrr)
```

### Basic Usage

A simple example of applying a function to a list using `map()`:

```{r}
# Define a list
numbers <- list(1, 2, 3, 4, 5)

# Apply a square function to each element
squared <- numbers |> map(~ .x^2)

# Print results
squared
```

Here, `~ .x^2` defines an anonymous function that squares each number.

## Key Functions and Features

### `map()`

#### Description

The `map()` family applies a function to each element of a list or vector and returns a list, vector, or another type based on the specific `map_*` function used.

#### Syntax

```r
map(.x, .f, ...)
```

- `.x`: Input list or vector.
- `.f`: Function to apply.
- `…`: Additional arguments passed to `.f`.

#### Example

```{r}
# Apply `sqrt()` to each element
roots <- list(4, 9, 16) |> map(sqrt)

# Print results
roots
```

### `map2()`

#### Description

Applies a function to two inputs simultaneously, element by element.

#### Syntax

```r
map2(.x, .y, .f, ...)
```

#### Example

```{r}
# Combine two vectors by addition
vector1 <- c(1, 2, 3)
vector2 <- c(4, 5, 6)

result <- map2(vector1, vector2, ~ .x + .y)

result
```

### `pmap()`

#### Description

Applies a function to multiple inputs, where inputs are supplied as a list of named arguments.

#### Syntax

```r
pmap(.l, .f, ...)
```

- `.l`: A list of arguments.
- `.f`: Function to apply.

#### Example

```{r}
# Define inputs
inputs <- list(a = c(1, 2, 3), b = c(4, 5, 6), c = c(7, 8, 9))

# Apply a custom function
result <- pmap(inputs, \(a, b, c) a + b + c)

result
```

## In-Depth Examples

### Title: Aggregating Results Using `reduce()`

#### Objective

Sum a list of numbers iteratively using `reduce()`.

#### Steps

1. **Define the data**:

```{r}
numbers <- list(1, 2, 3, 4, 5)
```

2. **Apply `reduce()`**:

```{r}
total <- reduce(numbers, \(x, y) x + y)
total
```

3. **Interpret the result**: The sum of the numbers is computed iteratively.

### Title: Handling Errors with `safely()`

#### Objective

Use `safely()` to apply a function that might fail.

#### Steps

1. **Define the data**:

```{r}
inputs <- list(10, "a", 20)
```

2. **Use `safely()`**:

```{r}
safe_sqrt <- safely(sqrt)
results <- map(inputs, safe_sqrt)
```

3. **Inspect results**:

```{r}
results
```

The output includes successful results and captured errors.

### Title: Extracting Nested Elements with `pluck()`

#### Objective

Retrieve nested elements from a list.

#### Steps

1. **Define a nested list**:

```{r}
nested_list <- list(a = list(b = list(c = 42)))
```

2. **Use `pluck()`**:

```{r}
value <- pluck(nested_list, "a", "b", "c")
value
```

## Advanced Features

### Customization

Customize outputs with `map_*` variants:

```{r}
# Return a vector instead of a list
numbers <- list(1, 2, 3)
squared <- map_dbl(numbers, ~ .x^2)
squared
```

### Integration

Combine `purrr` with `dplyr` for tidy workflows:

```{r}
library(dplyr)
data <- tibble(x = list(1:5, 6:10, 11:15))

# Apply a function to each list element
data |> mutate(mean_x = map_dbl(x, mean))
```

## Troubleshooting and FAQs

### Common Issues

1. **Error: "Object not found"**: Ensure all variables are correctly defined and scoped.
2. **Unexpected list outputs**: Use `map_*` variants (`map_dbl`, `map_chr`) for specific return types.

### FAQs

- **Can I use custom functions with `purrr`?** Yes, pass any custom or anonymous function to `map()`.

## Best Practices

- Use `map_*` variants for predictable output types.
- Combine `purrr` with other tidyverse packages for efficient workflows.
- Leverage error-handling tools like `safely()` for robust code.

## Conclusion

### Summary

The `purrr` package simplifies iteration, functional programming, and working with nested data. Its consistent syntax and integration with the tidyverse make it an essential tool for R users.

### Next Steps

- Explore `purrr`'s advanced tools like `transpose()` and `possibly()`.
- Integrate with visualization packages like `ggplot2` for applied analyses.

## References and Resources

- [Official Documentation](https://purrr.tidyverse.org/)
- [Tidyverse Website](https://www.tidyverse.org/)
- [R for Data Science](https://r4ds.had.co.nz/)

```{mermaid}
%%| label: fig-mermaid
%%| fig-cap: "Data Transformation Workflow"
%%| echo: false
graph TD
    A["Data Input"] --> B["Apply map()"]
    B --> C["Iterate over elements"]
    C --> D["Transform Results"]
    D --> E["Output List/Vector"]
```