---
title: "sjPlot: Visualizing Regression Models in R"
---

## Introduction

### Overview

The `sjPlot` package provides tools for creating publication-ready visualizations and tables for statistical models, survey data, and cross-tabulations. Key functions include:

- `sjPlot::plot_model()`: Visualizes regression model effects
- `sjPlot::tab_model()`: Creates HTML/Markdown tables for model summaries
- `sjPlot::tab_xtab()`: Generates contingency tables with statistics

### Use Cases

- Visualizing interactions in mixed-effects models
- Reporting odds ratios from logistic regression
- Creating APA-style frequency tables

### Benefits

- Unified syntax for diverse model types (lm, glm, lme4)
- Automatic effect size calculations
- Seamless integration with `ggplot2` for customization

## Getting Started

### Loading the Package

```{r}
pacman::p_load(sjPlot, lme4, ggplot2, ggpubr, effects)
theme_set(theme_pubr())
```

### Basic Usage

```{r}
# Fit a linear model
model <- lm(mpg ~ wt + cyl + gear, data = mtcars)

# Create effect plot
plot_model(model, type = "eff", terms = "wt") |>
  ggplot2::ggtitle("Effect of Weight on MPG")
```

## Key Functions and Features

### plot_model()

**Description:** Visualizes regression coefficients or marginal effects
**Syntax:**

```r
sjPlot::plot_model(
  model, 
  type = c("est", "eff", "pred"), 
  terms = NULL,
  show.values = TRUE
)
```

**Example:**

```{r}
# Fit mixed-effects model
mixed_model <- lme4::lmer(Reaction ~ Days + (1 | Subject), data = sleepstudy)

# Plot coefficients
plot_model(mixed_model, type = "est", title = "Sleep Study Coefficients") +
  ggplot2::theme_minimal()
```

### tab_model()

**Description:** Creates formatted model summary tables
**Syntax:**

```r
sjPlot::tab_model(
  ...,
  show.ci = TRUE,
  show.aic = TRUE,
  string.pred = "Predictor"
)
```

**Example:**

```{r}
glm_model <- glm(vs ~ wt + mpg, data = mtcars, family = binomial)

tab_model(
  glm_model,
  show.r2 = TRUE,
  pred.labels = c("Intercept", "Weight", "MPG")
)
```

## In-Depth Examples

### Example 1: Prediction plot for Interaction Effects Visualization

**Objective:** Demonstrate temperature × insulation interaction in housing data

```{r}

data(housing, package = "MASS")
model_interaction <- glm(
  Freq ~ Type * Cont + Sat + Infl,
  family = poisson,
  data = housing
)

anova(model_interaction)
```

```{r}
#| label: fig-interaction-plot
#| fig-cap: "Prediction Plot of Interaction Effects"
plot_model(
  model_interaction,
  type = "pred",
  terms = c("Type", "Cont"),
  axis.title = "Type of rental accommodation"
) +
  ggplot2::scale_color_viridis_d()
```

## Advanced Features

### Customization

Modify table appearances using built-in themes:

```{r}
tab_model(
  glm_model,
  CSS = list(css.table = "color: blue;"),
  title = "Customized Model Table"
)
```

### Integration with Ggplot2

```{r}
plot_model(mixed_model, type = "re") +
  ggplot2::scale_y_discrete(labels = c("Subject 1", "Subject 2")) +
  ggplot2::labs(caption = "Random Effects Plot")
```

## Troubleshooting and FAQs

### Common Issues

**Problem:** "Model class not supported" error
**Solution:** Convert models using `sjPlot::convert_to_sjmodel()`

**Problem:** Missing confidence intervals
**Solution:** Specify `ci.lvl` in plot_model:

```{r}
plot_model(model, ci.lvl = 0.99)
```

### FAQs

**Q:** How to export tables to Word?
**A:** Use `flextable::save_as_docx()`:

```{r}
tab_model(model) |> flextable::save_as_docx(path = "table.docx")
```

## Best Practices

1. Use model wrappers for complex objects:

```r
sjPlot::tab_model(lme4::lmer(...))
```

2. Maintain consistent theming:

```{r}
sjPlot::set_theme(
  theme = "539",
  axis.title.size = .9,
  geom.label.size = 3
)
```

## Conclusion

### Summary

- `sjPlot` streamlines statistical reporting through unified visualization syntax
- Key functions support >30 model types from common R packages

### Next Steps

- Explore `plot_stackfrq()` for Likert scale visualization
- Experiment with `tab_mixed()` for mixed-model comparisons

## References and Resources

- [Official Documentation](https://strengejacke.github.io/sjPlot/)
- [CRAN Vignettes](https://CRAN.R-project.org/package=sjPlot)
- [GitHub Issues](https://github.com/strengejacke/sjPlot/issues)