---
title: "Tutorial for the stringr Package"
date: '1/6/2025'
date-modified: last-modified
date-format: long
---

# Tutorial for the `stringr` Package

## Introduction

### Overview

The `stringr` package, part of the `tidyverse`, simplifies string manipulation in R by providing a consistent and human-readable interface for common operations like pattern matching, extraction, replacement, and string metrics. It builds on the base R string functions but offers better integration with modern workflows and tools.

Key functions include:

- **Basic Manipulation**: `str_c()`, `str_length()`, `str_trim()`, `str_pad()`.
- **Pattern Matching**: `str_detect()`, `str_extract()`, `str_match()`.
- **Substitution**: `str_replace()`, `str_replace_all()`.
- **Splitting and Joining**: `str_split()`, `str_flatten()`.
- **Other Utilities**: `str_to_upper()`, `str_to_title()`, `str_sort()`.

### Use Cases

- Cleaning and formatting text data for analysis.
- Extracting structured information from unstructured text.
- Detecting and replacing patterns in strings.
- Comparing and sorting textual data.

### Benefits

- Consistent and simple syntax.
- Handles regular expressions for powerful pattern matching.
- Fully vectorized for efficient processing of large datasets.
- Easily integrates with other `tidyverse` tools.

---

## Getting Started

### Loading the Package

Ensure `stringr` is loaded using the `pacman` package:

```{r}
pacman::p_load(stringr)
```

### Basic Usage

A simple example of string concatenation using `str_c()`:

```{r}
# Concatenate strings
str_c("Hello", "World", sep = " ")
```

This outputs "Hello World", showcasing the basic functionality of combining strings.

## Key Functions and Features

### `str_c()`

#### Description

Combines strings into a single string or vector of strings.

#### Syntax

```r
str_c(..., sep = "", collapse = NULL)
```

- `…`: Strings to combine.
- `sep`: Separator between strings.
- `collapse`: Collapses output into a single string.

#### Example

```{r}
# Combine names with separators
names <- c("John", "Jane", "Doe")
str_c(names, collapse = ", ")
```

### `str_length()`

#### Description

Returns the number of characters in each string.

#### Syntax

```r
str_length(string)
```

#### Example

```{r}
# Get string lengths
strings <- c("apple", "banana", "cherry")
str_length(strings)
```

### `str_detect()`

#### Description

Detects if a pattern exists in a string.

#### Syntax

```r
str_detect(string, pattern)
```

#### Example

```{r}
# Detect strings containing "an"
strings <- c("apple", "banana", "cherry")
str_detect(strings, "an")
```

### `str_extract()`

#### Description

Extracts the first match of a pattern from a string.

#### Syntax

```r
str_extract(string, pattern)
```

#### Example

```{r}
# Extract numbers from strings
strings <- c("Item 1", "Price 20", "Count 100")
str_extract(strings, "\\d+")
```

### `str_replace_all()`

#### Description

Replaces all occurrences of a pattern in a string.

#### Syntax

```r
str_replace_all(string, pattern, replacement)
```

#### Example

```{r}
# Replace spaces with underscores
strings <- c("apple pie", "banana split", "cherry tart")
str_replace_all(strings, " ", "_")
```

## In-Depth Examples

### Title: Extracting Phone Numbers from Text

#### Objective

Identify and extract phone numbers from text data.

#### Steps

1. **Prepare data**:

```{r}
text <- c("Call me at 555-123-4567", "My number is 987-654-3210", "No phone here")
```

2. **Define a regex pattern**:

```{r}
pattern <- "\\d{3}-\\d{3}-\\d{4}"
```

3. **Extract phone numbers**:

```{r}
phone_numbers <- str_extract(text, pattern)
phone_numbers
```

4. **Output interpretation**: Non-matching strings return `NA`.

### Title: Detecting Email Addresses

#### Objective

Check if a dataset contains email addresses.

#### Steps

1. **Input data**:

```{r}
strings <- c("Contact: john.doe@example.com", "Visit us", "Email: support@site.org")
```

2. **Detect emails**:

```{r}
email_pattern <- "[a-zA-Z0-9._%+-]+@[a-zA-Z0-9.-]+\\.[a-zA-Z]{2,}"
email_detected <- str_detect(strings, email_pattern)
email_detected
```

### Title: Splitting Strings into Words

#### Objective

Split sentences into individual words.

#### Steps

1. **Define sentences**:

```{r}
sentences <- c("This is an example", "R is great for text")
```

2. **Split into words**:

```{r}
words <- str_split(sentences, " ")
words
```

3. **Visualize results**:

```{r}
str_flatten(words[[1]], ", ")
```

## Advanced Features

### Customization

Use additional parameters for flexible pattern matching, such as `fixed()` for fixed strings or `regex()` for advanced options:

```{r}
# Case-insensitive matching
str_detect(c("Apple", "banana"), regex("apple", ignore_case = TRUE))
```

### Integration with Other Packages

Combine with `dplyr` to process text in data frames:

```{r}
library(dplyr)

data <- tibble(text = c("One 1", "Two 2", "Three 3"))
data |> mutate(numbers = str_extract(text, "\\d+"))
```

## Troubleshooting and FAQs

### Common Issues

1. **Regex not matching**: Ensure patterns are correctly escaped. Use `\\` for special characters like `\d`.
2. **Unexpected results**: Verify input data type and length. `NA` values in input may propagate.

### FAQs

- **Can `stringr` handle non-ASCII text?** Yes, it supports UTF-8 encoded text.
- **How do I extract multiple matches?** Use `str_extract_all()`.

## Best Practices

- Leverage vectorization for efficient processing of large datasets.
- Use clear and descriptive regex patterns.
- Combine with tidyverse tools for seamless workflows.

## Conclusion

### Summary

The `stringr` package offers powerful and intuitive tools for string manipulation in R. Its consistent syntax and comprehensive feature set make it a go-to package for text processing tasks.

### Next Steps

- Explore advanced regex patterns.
- Learn about `glue` for string interpolation.

## References and Resources

- [Official Documentation](https://stringr.tidyverse.org/)
- [Regex Cheatsheet](https://www.rstudio.com/resources/cheatsheets/)
- [R for Data Science: String Manipulation](https://r4ds.had.co.nz/strings.html)

```{mermaid}
%%| label: fig-mermaid
%%| fig-cap: "Text Data Workflow"
%%| echo: false
graph TD
    A[Raw Text Data] --> B[Pattern Matching]
    B --> C[Extraction]
    B --> D[Substitution]
    C --> E[Cleaned Data]
    D --> E
```