---
title: "R Package Tutorial: tipr"
---

# Introduction

- **tipr** performs *tipping-point sensitivity analyses* for unmeasured confounding: it solves for the strength of a hypothetical confounder that would move an observed estimate to the null (or any chosen reference value). [R Causal](https://r-causal.github.io/tipr/?utm_source=chatgpt.com)
- **Key functions covered here**
  - `tip()`, `tip_with_binary()` – generic continuous / binary confounder engines.
  - Scale-specific wrappers: `tip_or()`, `tip_rr()`, `tip_hr()`, `tip_coef()` and their `_with_binary`, `_with_r2` variants.
- **Typical use-cases**
  - Post-hoc robustness checks for observational studies.
  - Sensitivity reporting in clinical-trial subgroup or per-protocol analyses.
  - Pre-analysis planning (how strong must lurking confounding be to matter?).
- **Benefits vs. alternatives**
  - Can target *any* effect scale (OR, RR, HR, β).
  - Integrates cleanly with `broom` output pipes and plotting tools (unlike **EValue**, which returns scalars, or **sensemakr**, which focuses on partial $R^2$).
  - Offers quick grid evaluations useful for interactive dashboards. [Journal of Open Source Software](https://joss.theoj.org/papers/10.21105/joss.04495?utm_source=chatgpt.com)

## Getting Started

```{r}
#| label: setup
#| message: false
pacman::p_load(tipr, tidyverse, ggplot2, dagitty, ggdag)
```

## A First “Hello tipr”
```{r}
#| label: hello-tipr
#| message: false
# Suppose we observed a hazard ratio (HR) of 1.50 from a Cox model.
tipr::tip_hr(
  effect_observed = 1.50,
  exposure_confounder_effect = 0.40 # HR linking exposure and confounder
)
```

Interpretation: **If** an unmeasured *continuous* confounder raised the outcome hazard by the returned HR (≈ 2.8), *and* was associated with exposure via HR = 0.40, the observed HR would shrink to 1 (no effect).

### Scaffold-Yourself Exercise

1. Change the `exposure_confounder_effect` to 0.10, 0.20, … 1.00.
2. Record the required confounder–outcome HR each time.
3. Plot and describe the trend.

# Conceptual Foundations

### What Is a “Tipping Point”?

A tipping point is the set of confounder parameters $(\gamma_{UX},\ \gamma_{UY})$ that exactly *cancels* the bias in an observed effect $\hat\theta$. Algebraically we solve

$$
\theta_\text{obs} = \theta_\text{true}\times\operatorname{Bias}(\gamma_{UX},\gamma_{UY}),
$$

then set $\theta_\text{true}=1$ (null) and invert. For linear models the bias is additive; for ratio scales it is multiplicative.

### DAG Perspective (drawn with `dagitty` + `ggdag`)


```{r}
#| label: dag-basic
#| fig-cap: "Minimal DAG for tipping-point logic: U confounds the X→Y relation."
dag <- dagitty(
  "dag{
  U -> X
  U -> Y
  X -> Y
}"
)
ggdag(dag) +
  theme_dag_blank()
```

**Nuance:** *tipr* assumes *single* unmeasured confounder $U$. Multiple moderate confounders are equivalent to one composite confounder with aggregated effects—a useful but sometimes overlooked approximation.

# Key Functions and Features

Below each function appears twice: a **basic** and an **advanced** example, followed by tips and contrasts.

## 1. `tip()` – Continuous $U$ on Any Linear-Coefficient Scale

### Detailed Explanation

`tip()` targets estimates on *additive* scales (e.g., linear-regression coefficients). It requires the observed coefficient and *either* the exposure–confounder effect or the confounder–outcome effect; it returns the unknown partner.

**Syntax**

```r
tip(effect_observed, exposure_confounder_effect = NULL,
    confounder_outcome_effect = NULL)
```

### Basic Example – Body-Mass vs. Blood Pressure


```{r}
#| label: tip-basic
#| message: false
coef_observed <- 2.0 # systolic mmHg per kg
tip(effect_observed = coef_observed, exposure_confounder_effect = 0.3)
```

### Advanced Example – High-Dimensional Simulation


```{r}
#| label: tip-adv
#| message: false
set.seed(579)
n <- 2e4
U <- rnorm(n)
X <- 0.5 * U + rnorm(n)
Y <- 1.2 * X + 0.8 * U + rnorm(n)
lm_raw <- lm(Y ~ X)
beta <- coef(lm_raw)[2]

tip(effect_observed = beta, confounder_outcome_effect = 0.8) # solve for γ_UX
```

### Tips, Nuances, Misconceptions

- **Linearity**: bias formula derives from linear expectation; use with log-transformed outcomes for skewed data.
- **Common misconception**: A large required γ_UY means “safe.” Not necessarily—plausibility depends on domain knowledge.
- **Contrast:** **sensemakr** computes *partial $R^2$* needed; *tipr* computes effect size directly, often easier to interpret in applied biomedical terms.

## 2. `tip_with_binary()` – Binary $U$

### Detailed Explanation

Adapts Cornfield conditions: specify two of

`exposed_confounder_prev`, `unexposed_confounder_prev`, `confounder_outcome_effect`.

### Basic Example – Smoking (U) in Diet-Heart Study


```{r}
#| label: tip-bin-basic
#| message: false
tip_with_binary(
  effect_observed = 1.4, # observed RR
  exposed_confounder_prev = 0.25,
  unexposed_confounder_prev = 0.05
)
```

### Advanced Example – Sensitivity Grid

```{r}
#| label: tbl-tip-bin-adv
#| tbl-cap: "Grid of required confounder–outcome effects over varying prevalences."
prev_grid <- expand.grid(
  exp_prev = seq(0.1, 0.5, 0.1),
  unexp_prev = seq(0.01, 0.2, 0.03)
)

# Create a "safe" version of tip_with_binary() that returns NA on error
safe_tip <- possibly(
  .f = function(exp_p, unexp_p) {
    tip_with_binary(
      effect_observed = 1.4,
      exposed_confounder_prev = exp_p,
      unexposed_confounder_prev = unexp_p
    )$confounder_outcome_effect
  },
  otherwise = NA_real_
)

prev_grid <- prev_grid |>
  mutate(
    gamma_UY = map2_dbl(
      exp_prev,
      unexp_prev,
      safe_tip
    )
  )

prev_grid
```

### What happens here

1. **`possibly()`**  
   ```r
   safe_tip <- possibly(.f, otherwise = NA_real_)
   ```  
   Wraps your original anonymous function so that if it ever errors (i.e. the prevalence difference is too small), you quietly get `NA_real_` instead of stopping.

2. **`map2_dbl()`**  
   Applies `safe_tip(exp_prev[i], unexp_prev[i])` for each row and always returns a **numeric** vector (with `NA`s where the underlying call failed).

3. **Result**  
   Your `prev_grid$gamma_UY` will now be a length-15 numeric vector. Whenever `tip_gamma()` can’t find a valid confounder (as in your index 16 error), you’ll see an `NA` there rather than a hard stop.


Visualize:

```{r}
#| label: tip-bin-heat
ggplot(prev_grid, aes(exp_prev, unexp_prev, fill = gamma_UY)) +
  geom_tile() +
  labs(fill = "Required U→Y RR")
```

### Nuances & Compare

- **Rare U assumption**: When `unexposed_confounder_prev≃0`, results mimic classical Cornfield inequality.
- **Contrast**: **EValue** provides a single minimal RR needed under *equal* prevalence—*tipr* lets prevalences differ, giving richer scenarios.


## 3. Scale-Specific Wrappers (`tip_or()`, `tip_rr()`, `tip_hr()`)

### Detailed Explanation

These thin wrappers set default link transformers and interpretation messages. `tip_hr()` offers an **`hr_correction`** argument for common outcomes (>15 %) to adjust the risk-ratio approximation [RDocumentation](https://www.rdocumentation.org/packages/tipr/versions/1.0.2/topics/tip_hr?utm_source=chatgpt.com).

### Basic Example – Odds Ratio


```{r}
#| label: tip-or-basic
or_obs <- 2.1
tip_or(effect_observed = or_obs, exposure_confounder_effect = 0.3)
```

### Advanced Example – Hazard Ratio with Correction


```{r}
#| label: tip-hr-adv
tip_hr(
  effect_observed = 1.6,
  exposure_confounder_effect = 0.4,
  hr_correction = TRUE
)
```

### Nuances

- **HR Correction**: For cumulative incidences >15 %, hazard ratios diverge from risk ratios—enable `hr_correction`.
- **Misconception**: Setting *both* `exposure_confounder_effect` *and* `confounder_outcome_effect` will *always* error; but if both are supplied, the function reports how many such confounders would be needed to tip (nice diagnostic!).

## 4. Partial $R^2$ Variants (`*_with_r2`)

### Detailed Explanation

Instead of effect sizes, specify the *partial $R^2$* that $U$ explains in *X* and *Y*. Helpful when domain experts think in %-variance terms.

### Example – Robustness Value

```{r}
#| label: tip-r2
tip_coef_with_r2(
  effect_observed = 0.25,
  se = 0.05,
  df = 100,
  # Assume U explains 4% of the variance in the exposure (X).
  # The function then solves for the required variance U must explain in the outcome (Y).
  exposure_r2 = 0.04,
  confounder_outcome_r2 = .1, # optional: if known, plug it in to see how many such confounders would be needed to tip
)
```

### Compare vs. **sensemakr**

`sensemakr` computes “robustness value” via $R^2$; *tipr* adds direct solver so you can toggle between effect-size and variance descriptions seamlessly.

# Common Workflows & In-Depth Examples

## Basic Workflow (Logistic Regression)


```{r}
#| label: wf-logistic
model <- glm(am ~ hp + wt, data = mtcars, family = binomial)
or_hp <- broom::tidy(model, exponentiate = TRUE) |>
  filter(term == "hp") |>
  pull(estimate)

tip_or(effect_observed = or_hp, exposure_confounder_effect = 0.2)
```

## Intermediate Workflow – Continuous vs. Binary $U$ Side-by-Side

1. Run `tip()` with realistic continuous $γ_{UX} = 0.3$.
2. Run `tip_with_binary()` under plausible prevalences.
3. Plot both required $γ_{UY}$ values to compare robustness.

## Advanced Workflow – Interactive Grid + DAG Overlay

```{r}
#| label: wf-adv
# Build grid of γ_UX and plot required γ_UY
ux <- seq(0.05, 1, 0.05)
# tip_or is vectorized, so we can pass the entire `ux` vector directly.
# This is more efficient and readable than using map_dbl.
grid <- tibble(
  ux = ux,
  # We need to extract the numeric vector from the tip_or() output
  uy = tip_or(
    effect_observed = 1.8,
    exposure_confounder_effect = ux
  )$confounder_outcome_effect
)

ggplot(grid, aes(ux, uy)) +
  geom_line() +
  labs(x = "γ_UX (Exposure–Confounder OR)", y = "γ_UY needed") +
  geom_vline(xintercept = 0.2, linetype = 2) # domain-expert plausible
```

Interpret the intersection where plausible $γ_{UX}$ meets required γ_UY. If intersection lies above typical U effects, the estimate is robust.

## Advanced Features & Customization

- **Vectorized calls**: pass vectors to `effect_observed` for batch analyses.
- **Integration**: because outputs are numeric, you can feed them into **shiny** dashboards for real-time robustness sliders.
- **Large datasets**: no raw data are needed—only summary estimates—making *tipr* trivially memory-safe.

## Troubleshooting and FAQs

| Symptom | Diagnosis | Fix |
|---------|-----------|-----|
| “unused argument” | Called `tip_with_binary` without parentheses | Add `()` or ensure all required args |
| Infinite output | Requested impossible combination (e.g., $γ_{UX} = 0$) | Use realistic non-zero $γ_{UX}$ or check input order |
| Negative required effect | Swapped sign on `exposure_confounder_effect` | Remember: log-ratio <0 means protective confounder |

## Best Practices

- **Triangulate**: report *tipr* alongside **EValue** and **sensemakr** for convergent credibility.
- **Visualize**: always pair numerical tipping points with ggplot curves for interpretability.
- **Document assumptions**: specify whether $γ_{UX}$ refers to log-OR, log-HR, etc.

## Conclusion

*tipr* translates the abstract worry of “unmeasured confounding” into tangible numbers researchers can debate. Use the scaffolded examples above to build from single-scenario checks to fully automated robustness dashboards.

## References and Resources

- tipr website and docs [R Causal](https://r-causal.github.io/tipr/?utm_source=chatgpt.com)
- McGowan LD (2022). *tipr: An R package for sensitivity analyses for unmeasured confounders*. *JOSS*, 7 (77), 4495. DOI 10.21105/joss.04495 [Journal of Open Source Software](https://joss.theoj.org/papers/10.21105/joss.04495?utm_source=chatgpt.com)
- CRAN manual (v1.0.2) [CRAN](https://cran.r-project.org/web/packages/tipr/tipr.pdf?utm_source=chatgpt.com)
- `tip_hr` documentation for HR correction details [RDocumentation](https://www.rdocumentation.org/packages/tipr/versions/1.0.2/topics/tip_hr?utm_source=chatgpt.com)