---
title: "Tutorial on the Games-Howell Test Using PMCMRplus"
---

## Introduction to the Games-Howell Test

The **Games-Howell test** is a post-hoc statistical procedure used to perform **all-pairs comparisons** between group means after a significant ANOVA result. It is particularly useful when the assumption of **equal variances across groups** is violated. Unlike the Tukey HSD test, which assumes homogeneity of variances, the Games-Howell test uses Welch's degrees of freedom to adjust for unequal variances, making it robust for such scenarios.

This tutorial will guide you through using the `gamesHowellTest` function from the `PMCMRplus` package in R.

## Installation

First, install and load the `PMCMRplus` package if not already installed:

```{r}
pacman::p_load("PMCMRplus")
```

## Function Syntax

The `gamesHowellTest` function has three primary methods:

1. **Default method**:
   `gamesHowellTest(x, g, …)`
   - `x`: Numeric vector of data or a list of numeric vectors.
   - `g`: Grouping vector/factor (ignored if `x` is a list).

2. **Formula method**:
   `gamesHowellTest(formula, data, …)`
   - `formula`: Formula of the form `response ~ group`.
   - `data`: Data frame containing variables in the formula.

3. **`aov` method**:
   `gamesHowellTest(x, …)`
   - `x`: A fitted model object of class `aov`.

## Step-by-Step Examples

### Example 1: Using an `aov` Object

We'll use the built-in `chickwts` dataset to compare chicken weights across feed types.

#### Step 1: Check Assumptions

1. **Normality**: Shapiro-Wilk test on residuals.
2. **Equal Variances**: Bartlett test.

```{r}
# Fit ANOVA model
fit <- aov(weight ~ feed, data = chickwts)

# Check normality of residuals
shapiro.test(residuals(fit))  # p > 0.05 → normality assumed

# Check homogeneity of variances
bartlett.test(weight ~ feed, data = chickwts)  # small p → unequal variances
```

#### Step 2: Run ANOVA

```{r}
anova(fit) 
```

#### Step 3: Perform Games-Howell Test

```{r}
res <- gamesHowellTest(fit)
summary(res)  # Pairwise comparisons with p-values
```

#### Step 4: Interpret Results

- **`feed`**: A significant effect of feed type on weight is observed. The Games-Howell test suggests that the means of weight are different between the different feed types.
- **`Residuals`**: The remaining variance is due to random error.

### Example 2: Using the Formula Method

Simulate data with three groups (A, B, C) with unequal variances.

```{r}
# Create data
set.seed(123)
group <- factor(rep(c("A", "B", "C"), each = 20))
values <- c(rnorm(20, mean=10, sd=2),  # Group A
            rnorm(20, mean=15, sd=4),  # Group B
            rnorm(20, mean=12, sd=3)) # Group C

# Perform Games-Howell Test
result <- gamesHowellTest(values ~ group)
summary(result)
```

## Key Output Interpretation

- **`statistic`**: Test statistic for each pair (t-value).
- **`p.value`**: Adjusted p-values using the Tukey distribution.
- Values are arranged in a lower-triangle matrix. A p-value <- 0.05 indicates a significant difference between groups.

## Assumptions and Considerations

1. **Normality**: The test assumes normality. Check using Shapiro-Wilk or QQ-plots.
2. **Unequal Variances**: Use Games-Howell when variances are unequal (check via Bartlett/Levene test).
3. **Post-Hoc Use**: Conduct only after a significant ANOVA result.

## When to Use Games-Howell vs. Other Tests

- **Tukey HSD**: Use when variances are equal.
- **Dunnett's Test**: Compare all groups against a control.
- **Bonferroni**: Conservative adjustment for any pairwise test.

## Conclusion

The `gamesHowellTest` is ideal for comparing group means when variances are unequal. Always validate assumptions before applying the test. Use the formula or `aov` method for cleaner syntax with data frames or model objects.