---
title: "Testing Linear and Quadratic Trends with Unequal Variance in ANOVA"
---

## Intuitive Understanding

**Concept**:
In ANOVA, when groups have a natural order (e.g., drug doses: low, medium, high), we often test if the outcome follows a **linear** or **quadratic** trend. A linear trend implies a steady increase/decrease across groups, while a quadratic trend suggests a U-shaped or inverted U-shaped pattern.

**Problem with Unequal Variance**:
Standard ANOVA assumes equal variance across groups. When variances differ (heteroscedasticity), the usual F-test becomes unreliable. For example, groups with larger variances may dominate the analysis, leading to incorrect conclusions.

**Solution (Satterthwaite's Approximation)**:
Satterthwaite's method adjusts the degrees of freedom (DF) for the t-test used to evaluate contrasts (e.g., linear/quadratic trends). This adjustment accounts for unequal variances, producing a more accurate p-value. Think of it as "balancing" the influence of each group based on its variability.

---

## Mathematical Foundations

### Linear And Quadratic Contrasts

For $k$ groups with means $\bar{y}_1, \dots, \bar{y}_k$, define contrasts using coefficients $c_i$:

- **Linear**: $c_i = \text{linear scaling (e.g., -1, 0, 1)}$
- **Quadratic**: $c_i = \text{quadratic scaling (e.g., 1, -2, 1)}$

**Contrast Estimate**:

$$
\psi = \sum_{i=1}^k c_i \bar{y}_i
$$

**Standard Error (Unequal Variances)**:

$$
SE(\psi) = \sqrt{\sum_{i=1}^k \frac{c_i^2 s_i^2}{n_i}}
$$

where $s_i^2$ is the variance of group $i$, and $n_i$ is its sample size.

**Satterthwaite's Degrees of Freedom**:

$$
\nu = \frac{\left( \sum_{i=1}^k \frac{c_i^2 s_i^2}{n_i} \right)^2}{\sum_{i=1}^k \frac{c_i^4 s_i^4}{n_i^2 (n_i - 1)}}
$$

This formula approximates the effective DF by combining group-specific variances, derived from Welch's t-test extension.

**Test Statistic**:

$$
t = \frac{\psi}{SE(\psi)} \sim t_{\nu}
$$

**Key Theorems**:

- **Welch-Satterthwaite Equation**: Approximates the distribution of a weighted sum of variances as a scaled chi-square distribution.
- **Central Limit Theorem**: Justifies using the t-distribution for large samples.

---

## Numerical Example in R

**Data Generation**:
Simulate three groups with unequal variances:

```{r}
set.seed(123)
group <- factor(rep(1:3, each = 5))
y <- c(
    rnorm(5, 10, 1), # Group 1: μ=10, σ=1
    rnorm(5, 12, 2), # Group 2: μ=12, σ=2
    rnorm(5, 15, 3)
) # Group 3: μ=15, σ=3
```

**Manual Calculation**:

```{r}
# Group statistics
means <- tapply(y, group, mean) # 10.19357 11.91136 15.92371
vars <- tapply(y, group, var) # 0.6577564 5.4148322 3.6802184
ns <- tapply(y, group, length) # 5 5 5

# Linear contrast coefficients (from contr.poly)
c_linear <- contr.poly(3)[, 1] # [-0.707, 0, 0.707]

# Contrast estimate
contrast_est <- sum(c_linear * means) # ≈ 4.051817

# Standard error
se <- sqrt(sum((c_linear^2 * vars) / ns)) # 0.658633

# Satterthwaite DF
df_num <- (sum((c_linear^2 * vars) / ns))^2
df_den <- sum((c_linear^4 * vars^2) / (ns^2 * (ns - 1)))
df_satt <- df_num / df_den # ≈ 5.6

# t-test
t_ratio <- contrast_est / se # ≈ 6.15
p_value <- 2 * pt(abs(t_ratio), df_satt, lower.tail = FALSE)

# output create a table
data.frame(contrast_est, se, df_satt, t_ratio, p_value)
```

<!-- **Using R Packages**: -->

```{r}
#| label: tbl-conduct-trend-test-unequal-var
#| tbl-cap: "Conducting a trend test with unequal variance"
pacman::p_load(emmeans)
library(nlme)

# Fit a model allowing unequal variances
model_gls <- gls(y ~ group, weights = varIdent(form = ~ 1 | group))
emm <- emmeans(model_gls, ~group)

# Test linear trend
contrast(emm, method = "poly", adjust = "none")
```

---

## Comprehension Questions (Appendix)

**Conceptual**:

1. Why does unequal variance violate ANOVA assumptions?
2. How does Satterthwaite's approximation improve trend testing?

**Computational**:
3. Calculate the standard error for a quadratic contrast with coefficients $(1, -2, 1)$, variances $(4, 1, 9)$, and $n_i = 5$.

**Application**:
4. Interpret the R output: $t = 2.5, df = 8.3, p = 0.03$.

---

## Detailed Solutions

**1. Answer**:
Unequal variances distort the pooled variance estimate, inflating Type I/II errors. Groups with larger variances dominate the error term, reducing test reliability.

**2. Answer**:
Satterthwaite's method adjusts the DF to account for heterogeneous variances, ensuring the t-test better approximates the true sampling distribution.

**3. Solution**:

$$
SE = \sqrt{\frac{1^2 \cdot 4}{5} + \frac{(-2)^2 \cdot 1}{5} + \frac{1^2 \cdot 9}{5}} = \sqrt{\frac{4 + 4 + 9}{5}} = \sqrt{3.4} \approx 1.84
$$

**4. Answer**:
The linear trend is significant ($t(8.3) = 2.5, p = 0.03$), suggesting a steady increase/decrease across groups despite unequal variances.