---
title: "The Causal Effect of [Treatment] on [Outcome]: A Target Trial Emulation"
subtitle: "STAT 579 Final Project"
author:
  - name: Your Name
    email: your.email@unm.edu
    affiliations: 
      - id: unm
        name: University of New Mexico
        department: Department of Mathematics & Statistics
        address: 1 University of New Mexico
        city: Albuquerque
        state: NM
        postal-code: 87131
    attributes:
      corresponding: true
    note: This manuscript was generated as a final project for STAT 579.
abstract: |
  # TODO: Rewrite this abstract last.
  This study investigates the causal effect of [Treatment X] on [Outcome Y] using observational data from [Data Source]. We adopt a Target Trial Emulation framework to minimize selection bias and clearly define the causal contrast. Using [Method: e.g., Propensity Score Matching], we adjusted for [Number] baseline confounders to satisfy the exchangeability assumption. Our primary analysis yields a [Risk Difference/Ratio] of [Value] (95% CI: [Lower], [Upper]). Sensitivity analysis using the E-value suggests that an unmeasured confounder would need a relative risk of at least [E-value] to explain away these results. These findings suggest [Conclusion].
keywords: 
  - Causal Inference
  - Target Trial Emulation
  - Propensity Scores
  - Sensitivity Analysis
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Applied Causal Inference
      formatting: preprint
      cite-style: authoryear
---

```{r}
#| label: setup
#| include: false
# Load necessary packages
library(tidyverse)
library(ggdag)
library(dagitty)
library(MatchIt)
library(cobalt)
library(EValue)
library(knitr)
library(broom) # Added for tidying model outputs

# REVIEWER NOTE: Set echo = FALSE for the paper body.
# Only show code in the Appendix unless the assignment asks otherwise.

# --- Dummy Data Generation for Template ---
# This code creates a sample dataset to make the template runnable.
# Replace this with your actual data loading and cleaning steps.
set.seed(123)
n <- 500
sim_data <- tibble(
  Age = rnorm(n, mean = 50, sd = 10),
  SES = rnorm(n, mean = 100, sd = 20),
  Health = rnorm(n, mean = 75, sd = 5)
) |>
  mutate(
    # Treatment assignment depends on confounders
    A_prob = plogis(-1 + 0.02 * Age + 0.01 * SES - 0.01 * Health),
    A = rbinom(n, 1, A_prob),
    # Outcome depends on treatment and confounders
    Y = 10 +
      1.5 * A +
      0.5 * Age +
      0.2 * SES +
      0.1 * Health +
      rnorm(n, sd = 5)
  )

# 1. Propensity Score Estimation & Matching
m_out <- matchit(
  A ~ Age + SES + Health,
  data = sim_data,
  method = "nearest",
  distance = "glm"
)

# 2. Get matched data
matched_data <- match.data(m_out)

# 3. Estimate Treatment Effect
fit <- lm(Y ~ A, data = matched_data, weights = weights)
```

# Introduction

The central objective of this study is to estimate the causal effect of **[Treatment A]** on **[Outcome Y]** among **[Population P]**. Specifically, we ask: *"Does [Intervention] cause a change in [Outcome] compared to [Control]?"*

Previous literature has established associations between X and Y [@citation], yet causal claims remain difficult due to confounding. This paper contributes to the literature by applying rigorous causal identification strategies, specifically the **Target Trial Framework** described by Hernán and Robins [-@hernan2020causalwhatif], to observational data.

# Methods

## The Target Trial Protocol

To avoid common biases such as immortal time bias or ill-defined interventions, we explicitly map our observational analysis to a hypothetical target trial. @tbl-protocol details this correspondence.

::: {.callout-important}

## REVIEWER ACTION REQUIRED

The table below contains generic placeholders. You must update the "Observational Emulation" column to match the **exact** filters and definitions you used in your R data cleaning step.
:::

```{r}
#| label: tbl-protocol
#| tbl-cap: Correspondence between the Target Trial and Observational Emulation

protocol <- tibble(
  Component = c(
    "Eligibility",
    "Treatment Strategies",
    "Assignment",
    "Follow-up",
    "Outcome",
    "Causal Contrast"
  ),
  `Target Trial (Ideal)` = c(
    "Adults >18 with [Condition]",
    "Randomize to [A=1] or [A=0]",
    "Random Assignment",
    "From randomization until event",
    "Diagnosed [Outcome]",
    "Intention-to-Treat (ITT)"
  ),
  `Observational Emulation` = c(
    "Filter data: Age >18, clean records",
    "Classify based on prescription at t=0",
    "Adjust for confounders via Matching",
    "From index date (t=0) until censoring",
    "ICD-10 codes in EHR",
    "ITT effect of initiation"
  )
)

knitr::kable(protocol)
```

## Identification Strategy

We rely on the assumption of conditional exchangeability ($Y^a \perp A | L$). The assumed causal structure is visualized in @fig-dag.

```{r}
#| label: fig-dag
#| fig-cap: Directed Acyclic Graph (DAG) representing the assumed causal structure.
#| fig-width: 6
#| fig-height: 4
#| fig-align: center
#| out-width: 80%

# TODO: Update this DAG to match your actual variables
dag <- dagify(
  Y ~ A + Age + SES + Health + U, # Added U for unmeasured
  A ~ Age + SES + Health + U,
  exposure = "A",
  outcome = "Y",
  latent = "U", # Mark U as latent (unmeasured)
  labels = c(
    Y = "Outcome",
    A = "Treatment",
    Age = "Age",
    SES = "Socioecon",
    Health = "Health",
    U = "Unmeasured"
  )
)

# Plot
ggdag(dag, text = FALSE, use_labels = "label") +
  theme_dag() +
  theme(legend.position = "none")
```

Our statistical analysis plan proceeds as follows:

1. **Propensity Score Estimation:** We estimated the probability of treatment using logistic regression conditional on baseline covariates.
2. **Adjustment Method:** We utilized [Method, e.g., 1:1 Nearest Neighbor Matching] to create exchangeable groups.
3. **Estimation:** The treatment effect was estimated using a [Linear/Logistic] regression on the matched data.
4. **Sensitivity Analysis:** We calculated the E-value to assess robustness to unmeasured confounding (represented by $U$ in the DAG).

# Results

## Study Population and Balance

The initial cohort included $N = [Total]$ individuals. Before adjustment, the treated group was significantly different from the control group.

As shown in @fig-loveplot, the matching procedure successfully reduced the standardized mean differences (SMD) for all covariates below the 0.1 threshold.

```{r}
#| label: fig-loveplot
#| fig-cap: Covariate Balance (Love Plot). Red dots represent unadjusted imbalance; blue dots represent balance after matching.
#| fig-width: 6
#| fig-height: 4
#| fig-align: center
#| out-width: 80%

# This plot shows the balance of covariates before and after matching.
# We are aiming for standardized mean differences (SMDs) below 0.1.
love.plot(
  m_out,
  stats = c("mean.diffs"),
  threshold = .1,
  var.order = "unadjusted",
  binary = "std",
  title = "Covariate Balance Before and After Matching"
)
```

## Causal Effect Estimate

@tbl-results presents the primary findings.

::: {.callout-important}

## REVIEWER ACTION REQUIRED

I have updated the code chunk below. Do not manually type your results. Use `broom::tidy()` on your model object (`fit`) so the table updates automatically if you change your data cleaning.
:::

```{r}
#| label: tbl-results
#| tbl-cap: Estimated Causal Effect of Treatment on Outcome

# The table below is generated automatically from our regression model `fit`.
# We use broom::tidy() to create a clean summary.
broom::tidy(fit, conf.int = TRUE) |>
  filter(term == "A") |>
  select(term, estimate, std.error, p.value, conf.low, conf.high) |>
  knitr::kable(
    digits = 3,
    col.names = c(
      "Term",
      "Estimate",
      "Std. Error",
      "P-value",
      "95% CI Low",
      "95% CI High"
    )
  )
```

The adjusted analysis yields a causal effect estimate of **1.20** (p = 0.04). This suggests that [Treatment] increases [Outcome] by 1.2 units compared to the control group, holding baseline characteristics constant.

# Sensitivity Analysis

Since this is an observational study, we cannot rule out unmeasured confounding. We calculated the **E-value** [@vanderweele2017sensitivity] for our point estimate and confidence interval.

```{r}
#| label: calc-evalue
#| echo: true

# Extract model results for E-value calculation
model_summary <- summary(fit)
est <- model_summary$coefficients["A", "Estimate"]
se <- model_summary$coefficients["A", "Std. Error"]
sd_y <- sd(matched_data$Y)

# Calculate E-value
e_value_results <- EValue::evalues.OLS(
  est = est,
  se = se,
  sd = sd_y
)

# Print the E-value for the point estimate
e_value_results["E-values", "point"]
```

The E-value for the point estimate is **`r knitr::inline_expr('round(e_value_results["E-values", "point"], 2)')`**. This implies that an unmeasured confounder would need to be associated with both the treatment and the outcome by a risk ratio of at least **`r knitr::inline_expr('round(e_value_results["E-values", "point"], 2)')`** to explain away the observed effect. Given that known confounders like Smoking usually have RRs around [Value], it is [plausible/implausible] that an unmeasured confounder of this magnitude exists.

# Discussion

We successfully emulated a target trial to estimate the effect of X on Y. Our findings support the hypothesis that [Conclusion].

**Limitations:**

1. **Unmeasured Confounding:** Despite our sensitivity analysis, unmeasured factors such as [Factor Z] could bias results.
2. **Positivity:** We excluded patients with [Characteristic] due to lack of overlap, limiting generalizability.

**Conclusion:**

These results provide evidence for [Policy/Clinical implication]. Future research should focus on [Next Step].


\newpage

# Appendix: R Code {-}

The following R code was used to generate the analysis and results presented in this report.

```{r}
#| label: appendix-code
#| ref.label: !expr knitr::all_labels()
#| echo: true
#| eval: false

# This Chunk Acts as a Placeholder.
# The 'ref.label' Option above Automatically Pulls in Code from All other Chunks.
```

\newpage

# References {-}