---
title: "Creating Data Frames in R"
date: "1/6/2025"
date-modified: last-modified
date-format: long
---

# Creating Data Frames in R

## Introduction

### Overview

Creating data frames is fundamental in R for managing and analyzing tabular data. This tutorial explores:

- **Base R Functions:** `data.frame()` and `as.data.frame()`.
- **`tidyverse` Functions:** `tibble::tibble()` and `dplyr::data_frame()`.

Data frames allow structured data storage, enabling row and column manipulations akin to spreadsheets.

### Use Cases

- Cleaning and structuring raw data.
- Aggregating results from computations.
- Preparing data for visualization or modeling.

### Benefits

- **Base R:** Straightforward and widely supported.
- **`tidyverse`:** Intuitive syntax, compatibility with the pipe operator `|>`, and enhanced features like printing.

## Getting Started

### Loading the Package

```{r}
# Use pacman to load packages
pacman::p_load(tidyverse)
```

### Basic Usage in Base R

```{r}
# Create a simple data frame
data_base <- data.frame(
    Name = c("Alice", "Bob", "Charlie"),
    Age = c(25, 30, 35),
    Score = c(85.5, 90.3, 78.9)
)
data_base
```

### Basic Usage in `tidyverse`

```{r}
# Create a tibble
data_tidy <- tibble::tibble(
    Name = c("Alice", "Bob", "Charlie"),
    Age = c(25, 30, 35),
    Score = c(85.5, 90.3, 78.9)
)
data_tidy
```

## Key Functions and Features

### Base R Functions

#### Description

`data.frame()` creates data frames from vectors or lists.

#### Syntax

```r
#| echo: fenced
data.frame(x, ..., row.names = NULL, check.rows = FALSE, check.names = TRUE)
```

#### Example

```{r}
data_example <- data.frame(
    ID = 1:3,
    Category = c("A", "B", "A"),
    Value = c(100, 200, 150)
)
data_example
```

### `tidyverse` Functions

#### Description

`tibble::tibble()` constructs modern data frames with automatic column type detection.

#### Syntax

```r
tibble::tibble(...)
```

#### Example

```{r}
data_tidy_example <- tibble::tibble(
    ID = 1:3,
    Category = c("A", "B", "A"),
    Value = c(100, 200, 150)
)
data_tidy_example
```

## In-Depth Examples

### Example 1: Joining Data Frames

#### Objective

Combine two data frames by a common key.

#### Steps

```{r}
df1 <- tibble::tibble(ID = 1:3, Value1 = c(10, 20, 30))
df2 <- tibble::tibble(ID = 2:4, Value2 = c(40, 50, 60))
merged <- dplyr::left_join(df1, df2, by = "ID")
merged
```

#### Output

Observe how `NA` fills for unmatched rows.

### Example 2: Filtering and Selecting

#### Objective

Filter rows where `Value1 > 15` and select specific columns.

#### Steps

```{r}
filtered <- merged |>
    dplyr::filter(Value1 > 15) |>
    dplyr::select(ID, Value1)
filtered
```

## Advanced Features

### Customization

#### Adjust Column Names

```{r}
colnames(data_base) <- c("FullName", "Years", "ExamScore")
data_base
```

#### Add Calculated Columns

```{r}
data_tidy <- data_tidy |>
    dplyr::mutate(Grade = ifelse(Score > 80, "Pass", "Fail"))
data_tidy
```

### Integration with Visualization

Generate a plot from the data.

```{r}
ggplot2::ggplot(data_tidy, ggplot2::aes(x = Age, y = Score)) +
    ggplot2::geom_point() +
    ggplot2::labs(title = "Age vs. Score", x = "Age", y = "Score")
```

### Optimization Tips

- Use `data.table` for large data sets.
- Predefine column types for better memory usage.

## Troubleshooting and FAQs

### Common Issues

1. **Data Mismatch:** Ensure column lengths match when creating data frames.

```r
# Error example
# data.frame(a = 1:3, b = 1:4)
```

2. **Type Conversion Warnings:** Avoid unintended type coercion by specifying `stringsAsFactors = FALSE`.

### FAQs

- **Why use `tibble`?** `tibble` provides cleaner, more readable output, and better handling of column names and types.

## Best Practices

- Use consistent column names and data types.
- Validate data integrity before further processing.
- Leverage `tidyverse` pipes for readability and maintainability.

## Conclusion

### Summary

- Base R is great for quick setups and legacy code.
- `tidyverse` excels for scalable, readable, and modern workflows.

### Next Steps

Explore advanced techniques like `data.table` or `R6` for complex data operations.

## References and Resources

- [Base R Documentation](https://www.rdocumentation.org/)
- [Tidyverse Guide](https://www.tidyverse.org/)
- [R for Data Science](https://r4ds.had.co.nz/)