---
title: "Data Transformation and Wrangling in R"
date: "1/6/2025"
date-modified: last-modified
date-format: long
---

# Data Transformation and Wrangling in R

## 1. Introduction

### Overview

Data transformation and wrangling involve cleaning, reshaping, and modifying data to prepare it for analysis. This tutorial introduces:

- **Base R Functions**: `subset()`, `transform()`, `merge()`.
- **tidyverse Functions**: `dplyr::filter()`, `dplyr::mutate()`, `tidyr::pivot_longer()`, and others.

### Use Cases

- Filtering rows based on conditions.
- Adding or modifying columns.
- Reshaping data between wide and long formats.
- Combining or joining datasets.

### Benefits

- **Base R**: Built into R, no additional installations required.
- **tidyverse**: Consistent syntax and powerful chaining with the pipe operator `|>`.

## 2. Getting Started

### Loading the Package

```{r}
# Base R requires no additional loading.

# Load tidyverse package using pacman
pacman::p_load(tidyverse)
```

## 3. Key Functions and Features

### Using Base R

#### Filtering Rows

**Description**: Extract rows based on a condition.

**Syntax**:

```r
subset(x, subset, select = NULL)
```

**Arguments**:

- `x`: A data frame or similar object.
- `subset`: Logical expression indicating rows to keep.
- `select`: Columns to retain, or `NULL` to keep all.

**Example**:

```{r}
data(iris)
# Filter rows where Species is "setosa"
subset(iris, Species == "setosa")
```

#### Modifying Columns

**Description**: Add or modify columns in a data frame.

**Syntax**:

```r
transform(data, ...)
```

**Arguments**:

- `data`: A data frame.
- `…`: Expressions specifying transformations.

**Example**:

```{r}
# Add a new column for Sepal Area
iris <- transform(iris, Sepal.Area = Sepal.Length * Sepal.Width)
head(iris)
```

#### Joining Data

**Description**: Combine datasets based on common columns (keys).

**Syntax**:

```r
merge(x, y, by, all)
```

**Arguments**:

- `x`, `y`: Data frames to merge.
- `by`: Column names to use as keys. Defaults to all common columns.
- `all`: Logical. `TRUE` for full outer join; `FALSE` for inner join.

**Example**:

```{r}
# Create example datasets
df1 <- data.frame(ID = 1:3, Value = c(10, 20, 30))
df2 <- data.frame(ID = 2:4, Score = c(15, 25, 35))
# Merge on ID
merge(df1, df2, by = "ID")
```

### Using Tidyverse

#### Filtering Rows

**Description**: Select rows using conditions with `dplyr::filter()`.

**Syntax**:

```r
dplyr::filter(.data, ...)
```

**Arguments**:

- `.data`: A data frame or tibble.
- `…`: Logical conditions. Rows where the condition is `TRUE` are retained.

**Example**:

```{r}
# Filter rows where Species is "setosa"
iris |>
  dplyr::filter(Species == "setosa")
```

#### Adding or Modifying Columns

**Description**: Create or modify columns using `dplyr::mutate()`.

**Syntax**:

```r
dplyr::mutate(.data, ...)
```

**Arguments**:

- `.data`: A data frame or tibble.
- `…`: Name-value pairs where names are new or existing column names, and values are expressions.

**Example**:

```{r}
# Add Sepal.Area column
iris |>
  dplyr::mutate(Sepal.Area = Sepal.Length * Sepal.Width)
```

#### Reshaping Data

**Description**: Convert data between wide and long formats with `tidyr::pivot_longer()`.

**Syntax**:

```r
tidyr::pivot_longer(data, cols, names_to, values_to)
```

**Arguments**:

- `data`: A data frame or tibble.
- `cols`: Columns to pivot into rows.
- `names_to`: Name of the new column containing the names of the pivoted variables.
- `values_to`: Name of the new column containing the values of the pivoted variables.

**Example**:

```{r}
# Convert iris to long format
iris_long <- iris |>
  tidyr::pivot_longer(
    cols = starts_with("Sepal"),
    names_to = "Measurement",
    values_to = "Value"
  )
head(iris_long)
```

#### Joining Data

**Description**: Combine datasets using `dplyr::inner_join()`.

**Syntax**:

```r
dplyr::inner_join(x, y, by)
```

**Arguments**:

- `x`, `y`: Data frames or tibbles to join.
- `by`: Column names to join by. Defaults to common column names.

**Example**:

```{r}
df1 <- tibble(ID = 1:3, Value = c(10, 20, 30))
df2 <- tibble(ID = 2:4, Score = c(15, 25, 35))
df1 |>
  dplyr::inner_join(df2, by = "ID")
```

## 4. In-Depth Examples

### Example 1: Cleaning and Transforming Data

**Objective**: Prepare `mtcars` for analysis by filtering rows, creating a new column, and summarizing data.

**Steps**:

1. Filter cars with `mpg > 20`.
2. Add a column `Weight_kg` converting `wt` to kilograms.
3. Summarize average MPG by the number of cylinders.

**Code**:

```{r}
# Using Base R
mtcars_filtered <- subset(mtcars, mpg > 20)
mtcars_filtered <- transform(mtcars_filtered, Weight_kg = wt * 453.6)
aggregate(mpg ~ cyl, data = mtcars_filtered, mean)

# Using tidyverse
mtcars |>
  dplyr::filter(mpg > 20) |>
  dplyr::mutate(Weight_kg = wt * 453.6) |>
  dplyr::group_by(cyl) |>
  dplyr::summarize(Average_MPG = mean(mpg))
```

## 5. Advanced Features

### Customization

**Base R**: Use `tapply()` for custom aggregations. **tidyverse**: Leverage `dplyr::group_by()` for grouped operations.

**Example**:

```{r}
# Base R
tapply(mtcars$mpg, mtcars$cyl, mean)

# tidyverse
mtcars |>
  dplyr::group_by(cyl) |>
  dplyr::summarize(Average_MPG = mean(mpg))
```

## 6. Troubleshooting and FAQs

### Common Issues

1. **`tidyr::pivot_longer()` errors**:

	- Ensure column names in `cols` exist.
	- Use `names_pattern` for extracting variable parts.
2. **Merge Mismatches in Base R**:

	- Verify columns in `by` are consistent between datasets.

## 7. Best Practices

1. Use **descriptive column names**.
2. Keep data in **tidy format** for easier analysis.
3. Prefer **tidyverse** for a consistent and readable workflow.

## 8. Conclusion

### Summary

- **Base R**: Ideal for simple tasks.
- **tidyverse**: Superior for complex workflows with a consistent syntax.

### Next Steps

Explore additional `tidyverse` packages like `lubridate` (dates) or `stringr` (text manipulation).

## 9. References and Resources

- [Base R Documentation](https://stat.ethz.ch/R-manual/R-devel/library/base/html/00Index.html)
- [tidyverse Documentation](https://www.tidyverse.org/)
- [R for Data Science](https://r4ds.had.co.nz/)

