Great! Let’s walk through a **real data example** of **fixed-effects nested ANOVA** in R, step-by-step. I'll explain the design, show the R syntax, how to interpret it, and—most importantly—**how to choose the correct denominator** for your F-tests.

---

## 🔷 Scenario: A Study on Machine Parts

Suppose you're testing **3 machines** (factor A), and within each machine, **3 operators** (factor B) are nested. You measure **output precision** (continuous variable) from **2 trials per operator**.

### 🧱 Hierarchical design:

- **Machine**: Fixed effect (Factor A)
- **Operator(Machine)**: Fixed effect nested in machine (Factor B nested in A)
- **Trial (replicates)**: Observational units

| Machine | Operator | Trial 1 | Trial 2 |
|---------|----------|---------|---------|
| A1      | O1       | 94      | 95      |
| A1      | O2       | 89      | 91      |
| A1      | O3       | 90      | 88      |
| A2      | O4       | 86      | 87      |
| A2      | O5       | 85      | 84      |
| A2      | O6       | 83      | 82      |
| A3      | O7       | 80      | 81      |
| A3      | O8       | 79      | 78      |
| A3      | O9       | 77      | 76      |

Let’s turn this into an R data frame and perform the analysis.

---

## 🔷 R Code: Creating the Data

```{r}
# Create the data frame
machine <- factor(rep(c("A1", "A2", "A3"), each = 6))
operator <- factor(rep(
    c("O1", "O2", "O3", "O4", "O5", "O6", "O7", "O8", "O9"),
    each = 2
))
response <- c(
    94,
    95,
    89,
    91,
    90,
    88,
    86,
    87,
    85,
    84,
    83,
    82,
    80,
    81,
    79,
    78,
    77,
    76
)

data <- data.frame(machine, operator, response)
```

---

## 🔷 R Code: Fixed-Effect Nested ANOVA

To do **nested ANOVA with fixed effects**, you must:

- Treat both factors as **fixed**.
- Use the nesting structure explicitly.

```{r}
# Nested ANOVA model: fixed-effects, operator nested in machine
model <- aov(response ~ machine + Error(operator), data = data)
summary(model)
```

BUT — if you want to **test Machine using Operator(Machine) as the denominator**, you **must specify nesting explicitly** using:

```{r}
model_nested <- aov(response ~ machine + machine:operator)
summary(model_nested)
```

---

## 🔶 Understanding the Output

From this model:

- `machine` is the main effect (A)
- `machine:operator` represents Operator nested in Machine (B nested in A)
- Residuals represent trial-to-trial variation (replicates)

The ANOVA table will look like:

| Source           | Df | Sum Sq | Mean Sq | F-value | Pr(>F) |
|------------------|----|--------|---------|---------|--------|
| machine          | 2  | ...    | ...     | F = MS_machine / MS_machine:operator | ... |
| machine:operator | 6  | ...    | ...     | F = MS_machine:operator / MS_residual | ... |
| Residuals        | 9  | ...    | ...     |         |        |

---

## ✅ Ensuring Correct Denominator (Key Step)

In R:
- The model `aov(response ~ machine + machine:operator)` assumes **all effects are fixed**.
- `machine:operator` provides the **correct denominator** for testing the `machine` effect.
- The **residual** is used for testing the `operator nested in machine`.

If instead you use:

```{r}
summary(aov(response ~ machine + operator))
```

Then R **won't** treat operator as nested, and you’ll get **wrong F-tests** because it’ll assume **crossed design**.

To emphasize nesting, always do:

```{r}
aov(response ~ machine + Error(operator)) # For mixed-model view
aov(response ~ machine + machine:operator) # For fixed effect ANOVA
```

---


## ✅ Summary: How to Get the Correct Denominator

| Effect Tested             | Numerator Term       | Denominator Term         | Syntax in R |
|--------------------------|----------------------|--------------------------|-------------|
| Machine (A)              | `machine`            | `machine:operator`       | `machine + machine:operator` |
| Operator nested in Machine (B(A)) | `machine:operator` | Residuals (trial-to-trial) | Implicit in `aov(...)` |
