---
title: "Figures and Tables in Quarto: Advanced Features"
---

# Introduction

Quarto is a powerful scientific and technical publishing system that extends R Markdown, offering enhanced functionality for rendering figures, tables, captions, and cross-references. With Quarto, you can produce high-quality, reproducible documents that integrate both code and text seamlessly. In this tutorial, we will explore how to create and reference figures and tables, include captions, and leverage various Quarto features.

Packages and functions we will focus on:

- Quarto chunk options (e.g.,`fig-cap`, `table-cap`, `label`, etc.)
- Base R functions for data wrangling and plotting (e.g., `plot(), summary()`)
- `ggplot2::ggplot()` for advanced plotting
- `dplyr::mutate()` and other `dplyr` functions for data manipulation
- `knitr::kable()` for basic table creation
- `gt::gt()` for more advanced table formatting

Use Cases:

- Creating reproducible reports with figure and table references.
- Writing scientific articles where cross-referencing figures and tables is essential.
- Producing dynamic documents that blend analysis, commentary, and visuals.

Benefits:

- Automatically numbered and labeled figures and tables.
- Easily referenced in text using Quarto's cross-reference syntax.
- Flexible customization of captions and layout.
- Consistent appearance across multiple output formats (HTML, PDF, Word).

# Getting Started

## Loading the Package

While you do not need a specific package to enable figure and table references in Quarto (beyond Quarto itself), you will typically rely on a suite of R packages for data manipulation and visualization. Below is an example of loading packages with pacman:

```{r}
pacman::p_load(
    dplyr,
    ggplot2,
    knitr,
    gt
)

ggplot2::theme_set(theme_minimal())
```

## Basic Usage

Quarto uses specially marked code chunks to control figure and table output. A simple code chunk with a figure could look like this:

```{r}
#| label: fig-basicplot
#| fig-cap: "A basic scatter plot with base R"
#| echo: fenced
plot(
    cars$speed,
    cars$dist,
    main = "Speed vs. Stopping Distance",
    xlab = "Speed (mph)",
    ylab = "Distance (ft)"
)
```

Annotations:

- `#| label: fig-basicplot` assigns a label to this code chunk, used for cross-referencing.
- `#| fig-cap: "A basic scatter plot with base R"` adds a figure caption.

To reference the figure in your text, you can write: "As shown in @fig-basicplot …".

# Key Functions and Features

## Quarto Chunk Options for Figures

Description: Quarto recognizes options like `fig-cap` to set a figure caption, `label` to set a reference label, and `fig-width`/`fig-height` to control figure size.

Syntax (not executed, just shown):

```r
#| label: chunk_label
#| fig-cap: "Figure caption here"
#| fig-width: 6
#| fig-height: 4
```

Explanation of arguments:

- `label` (character): Identifies the figure so it can be cross-referenced.
- `fig-cap` (character): The caption text for the figure.
- `fig-width`, `fig-height` (numeric): Dimensions (in inches) of the figure.

Example:

```{r}
#| label: fig-ggplot-example
#| fig-cap: "Iris Data: Sepal Width vs. Sepal Length"
#| fig-width: 6
#| fig-height: 4
#| echo: fenced
iris |>
    ggplot2::ggplot(ggplot2::aes(x = Sepal.Length, y = Sepal.Width)) +
    ggplot2::geom_point(color = "blue") +
    ggplot2::labs(title = "Iris Scatter Plot")
```

## Quarto Chunk Options for Tables

Description: Similar to figures, Quarto also provides `table-cap` for a table caption and `label` for referencing.

Syntax (not executed, just shown):

```r
#| label: tbl-label
#| tbl-cap: "Table caption here"
```

Example:

```{r}
#| label: tbl-kable-example
#| tbl-cap: "Summary of the cars dataset"
#| echo: fenced
summary_table <- knitr::kable(summary(cars))
summary_table
```

# In-Depth Examples

## Example 1: Referencing a Figure Generated by Ggplot2

Title: Iris Dataset Visualization

Objective: Create a scatter plot and reference it in the text.

Steps:

1. Load required packages.
2. Prepare the data (in this case, the iris dataset is already built in).
3. Create a figure chunk with a label and caption.
4. Reference the figure in the text.

```{r}
#| label: fig-iris
#| fig-cap: "Sepal Length vs. Petal Length in Iris dataset"
#| echo: fenced
iris |>
    ggplot2::ggplot(ggplot2::aes(
        x = Sepal.Length,
        y = Petal.Length,
        color = Species
    )) +
    ggplot2::geom_point() +
    ggplot2::labs(title = "Iris: Sepal vs. Petal Length")
```

Output:

- A scatter plot labeled as "fig-iris," automatically numbered by Quarto.
- To reference this figure in your text, use `Figure @fig-iris shows …`.

## Example 2: Creating and Referencing a Table with Gt

- Title: Summarizing mtcars with gt
- Objective: Produce a well-formatted table, then reference it in text.

Steps:

1. Use the mtcars dataset.
2. Summarize or transform if needed (e.g., select columns).
3. Create a table chunk with a label and caption.
4. Cross-reference within text.

```{r}
#| label: tbl-gtexample
#| tbl-cap: "Average Horsepower by Number of Cylinders"
#| echo: fenced
mtcars |>
    dplyr::group_by(cyl) |>
    dplyr::summarise(avg_hp = mean(hp)) |>
    gt::gt()
```

Output:

- A formatted table with "tbl-gt-example" as the label.
- Reference it with `Table @tbl-gt-example displays …` in your text. This produces a clickable link in the output as follows: Table @tbl-gtexample displays …".

# Advanced Features

## Customization

- Adjust figure size with `fig-width` and `fig-height` to control appearance.
- Change plot themes in `ggplot2::theme()`.
- Use chunk options like `eval = FALSE` or `echo = FALSE` to control whether code is run or shown.

Example of advanced customization for a figure:

```{r}
#| label: fig-custom-plot
#| fig-cap: "Customized Plot with Theme"
#| fig-width: 7
#| fig-height: 5
#| echo: fenced
iris |>
    ggplot2::ggplot(ggplot2::aes(
        x = Species,
        y = Sepal.Length,
        fill = Species
    )) +
    ggplot2::geom_boxplot() +
    ggplot2::theme_minimal() +
    ggplot2::labs(title = "Iris Boxplot")
```

## Integration

- Combine `dplyr::mutate()` and `ggplot2::ggplot()` to produce elaborate figures.
- Use `tidyr::pivot_longer()` or `pivot_wider()` to reshape data before tabulation.
- Pipe operator `|>` ensures code readability.

Example:

```{r}
#| label: tbl-integrated-example
#| table-cap: "Iris Species Mean Measurements"
#| echo: fenced
iris |>
    dplyr::group_by(Species) |>
    dplyr::summarise(across(
        .cols = c(Sepal.Length, Sepal.Width, Petal.Length, Petal.Width),
        .fns = ~ mean(.x, na.rm = TRUE)
    )) |>
    gt::gt()
```

## Optimization

- For large datasets, consider sampling or summarizing before creating tables or plots.
- Cache expensive operations using `#| cache: true` in your code chunk if your workflow supports it.

# Troubleshooting and FAQs

## Common Issues

1. **Figure or table labels not found**
   - Make sure you set `#| label: <label_name>` in the correct code chunk.
   - Check that you are referencing it with `@fig-<label_name>` or `@tbl-<label_name>` in the text.

2. **Caption not displayed**
   - Confirm you used `fig-cap` for figures or `tbl-cap` for tables.
   - Ensure no conflicting chunk options are overriding your settings.
å
3. **Figures and tables appear in the wrong order**
   - Quarto processes code chunks in the order they appear. Check the chunk's position in the document.

## FAQs

4. *How do I hide the code but show the figure/table?*
   - Use `echo = FALSE` in your chunk options.

5. *Can I reference a table as a figure or vice versa?*
   - Technically yes, but it's best to use `fig-` for figures and `tbl-` for tables for clarity.

6. *Why is my code block not producing a figure?*
   - Make sure you have a plotting function in the chunk and that `eval = TRUE` is not set to FALSE.

# Best Practices

- Always label your chunks clearly (e.g., `fig-mycars`, `tbl-irisstats`) for easy referencing.
- Use descriptive captions that help readers understand the figure or table at a glance.
- Keep your R code organized and modular (avoid overly large code chunks).
- For reproducibility, set a seed before random processes or sampling.
- Avoid copying large data directly into the chunk; load data from external files or scripts where possible.

# Conclusion

Summary:

In this tutorial, we explored how to produce, label, and reference figures and tables in Quarto using chunk options like `fig-cap` and `tbl-cap`. We demonstrated how ggplot2 and gt can create attractive visuals and tables, and how Quarto automatically handles cross-referencing.

Next Steps:

- Explore more advanced formatting with packages like kableExtra or flextable.
- Learn about Quarto's extended features such as bookdown-like multi-chapter documents.
- Consider adding interactive visuals with HTML widgets (e.g., `plotly`).

# References and Resources

Documentation:

- [Quarto Official Documentation](https://quarto.org/docs/)
- [Quarto Figures and Layout](https://quarto.org/docs/authoring/figures.html)
- [Quarto Tables](https://quarto.org/docs/authoring/tables.html)

Learning Materials:

- [R for Data Science](https://r4ds.had.co.nz/) (for dplyr, ggplot2 usage)
- [RStudio Blog on Quarto](https://www.rstudio.com/blog/) for updates and tutorials

Support Channels:

- [Quarto GitHub Issues](https://github.com/quarto-dev/quarto-cli/issues)
- [RStudio Community](https://community.rstudio.com/) for discussion and help
- [Stack Overflow](https://stackoverflow.com/questions/tagged/quarto) for Q&A

By following these guidelines and best practices, you can produce clean, professional figures and tables with clear references and captions, all within a single Quarto document. Happy reporting!