---
title: "Figures, Tables, Captions, and Referencing in Quarto: A Comprehensive Tutorial"
---

## 1. Introduction

### Overview

Quarto is a publishing system that integrates R code, text, and output (figures, tables) into dynamic documents. This tutorial focuses on creating figures/tables with captions, cross-referencing them, and generating reproducible reports.

**Key Features**:

- Add captions using `#| fig-cap` or `#| tbl-cap` in code chunks.
- Cross-reference figures/tables with `@label` syntax.
- Customize outputs using `ggplot2`, `knitr`, `kableExtra`, and `gt`.

### Use Cases

- Academic papers with numbered figures/tables.
- Technical reports with automated cross-references.
- Dashboards integrating dynamic visualizations and summaries.

### Benefits

- **Automation**: Update figures/tables and references dynamically.
- **Consistency**: Ensure captions and labels follow document structure.
- **Flexibility**: Output to HTML, PDF, Word, or slides.



## 2. Getting Started

### Loading the Package

```{r}  
# Load required packages
pacman::p_load(ggplot2, dplyr, kableExtra, gt, plotly)
ggplot2::theme_set(theme_minimal())
```  

### Basic Usage

**Figure with Caption**:

```{r}  
#| label: fig-mpg
#| fig-cap: "Scatterplot of car weight vs. MPG."

ggplot2::ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point()
```  

Reference the figure in text: `See Figure @fig-mpg`.

**Table with Caption**:

```{r}  
#| label: tbl-summary
#| tbl-cap: "Summary of MPG by cylinders."

mtcars |>
  group_by(cyl) |>
  summarize(avg_mpg = mean(mpg)) |>
  knitr::kable()
```  

Reference the table: `Table @tbl-summary shows…` produces "@tbl-summary shows…".

## 3. Key Functions and Features

### Quarto Chunk Options

**Description**: Control figure/table captions, labels, and layout.

**Syntax**:

```r  
#| label: fig-example  
#| fig-cap: "Caption text"  
#| tbl-cap: "Table caption"  
#| layout: "l-body-outset"  
```  

**Example**:

```{r}
#| label: fig-density
#| fig-cap: "Density plot of MPG."
#| echo: fenced

ggplot(mtcars, aes(mpg)) +
  geom_density(fill = "skyblue")
```  

Refer to the figure: `Figure @fig-density shows…` produces "@fig-density shows…".

### kableExtra::kbl()

**Description**: Create customizable tables with captions.

**Syntax**:

```r  
data |>  
  kableExtra::kbl(  
    caption = "Table title",  
    align = "c",  
    format = "html"  
  ) |>  
  kableExtra::kable_styling()  
```  

**Example**:

```{r}  
#| label: tbl-styled
#| tbl-cap: "Styled table with kableExtra."

mtcars |>
  head(3) |>
  kableExtra::kbl(caption = "Sample Cars") |>
  kableExtra::kable_styling(full_width = FALSE)
```  

## 4. In-Depth Examples

### Example 1: Dynamic Figure Captions

**Objective**: Generate a figure caption with variables.

**Steps**:

```{r}  
#| label: fig-dynamic
#| fig-cap: "MPG distribution for {{nrow(mtcars)}} cars."

ggplot(mtcars, aes(mpg)) +
  geom_histogram(bins = 10, fill = "steelblue")
```  

**Output**: Caption dynamically updates if data changes.

### Example 2: Multi-Column Table

**Objective**: Create a summary table grouped by cylinders.

**Steps**:

```{r}  
#| label: tbl-grouped
#| tbl-cap: "Detailed summary by cylinders."

mtcars |>
  group_by(cyl) |>
  summarize(
    Avg_MPG = mean(mpg),
    SD_MPG = sd(mpg)
  ) |>
  gt::gt() |>
  gt::tab_header(title = "Cylinder Performance")
```  

**Output**: Professionally formatted table with grouped data.

## 5. Advanced Features

### Customization

**Conditional Table Formatting**:

```{r}  
#| label: tbl-color
#| tbl-cap: "Table with conditional formatting."

mtcars |>
  head() |>
  mutate(
    high_mpg = ifelse(mpg > 20, "green", "red")
  ) |>
  kableExtra::kbl() |>
  kableExtra::column_spec(6, background = mtcars$high_mpg)
```  

**Interactive Figures with Plotly**:

```{r}  
#| label: fig-interactive
#| fig-cap: "Interactive MPG vs. Weight plot."

plotly::ggplotly(
  ggplot(mtcars, aes(wt, mpg)) +
    geom_point()
)
```  

### Integration

**Combine with dplyr**:

```{r}  
#| label: tbl-filtered
#| tbl-cap: "Filtered data table."

mtcars |>
  filter(cyl == 6) |>
  gt::gt()
```  

### Optimization

**Caching Large Datasets**:

```r  
#| cache: true  
```  

## 6. Troubleshooting and FAQs

### Common Issues

1. **Missing Captions**: Ensure `#| fig-cap` or `#| tbl-cap` is placed *inside* the code chunk.
2. **Broken Cross-References**: Verify that labels match exactly (e.g., `@fig-label` vs. `fig-label`).

### FAQs

**Q: How to reference figures in PDF output?**
A: Use the same `@label` syntax; Quarto handles format-specific numbering.

**Q: Can I use LaTeX in captions?**
A: Yes, for PDF output: `#| fig-cap: "\\textbf{Bold} caption text"`.


## 7. Best Practices

- Use **descriptive labels** (e.g., `fig-mpg-distribution` instead of `fig1`).
- Store reusable components (e.g., themes, table styles) in separate scripts.
- Prefer `gt` for complex tables and `kableExtra` for quick HTML formatting.

## 8. Conclusion

### Summary

Quarto streamlines figure/table creation, captioning, and cross-referencing. Key tools include:

- Chunk options (`#| label`, `#| fig-cap`).
- `ggplot2` for figures, `gt`/`kableExtra` for tables.

### Next Steps

Explore Quarto's [cross-referencing syntax](https://quarto.org/docs/authoring/figures.html) for equations, sections, and theorems.

## 9. References and Resources

- **Quarto Documentation**: <https://quarto.org>
- **ggplot2 Cheat Sheet**: <https://ggplot2.tidyverse.org>
- **Community Support**: RStudio Community, Stack Overflow (tag: `quarto`).
