---
title: "Simulate Random Interaction in Crossed Random Effects Models"
---

Okay, let's break down `lmer` interpretation, data simulation, and formula specification for crossed random effects models.

## Model Interpretation in `lmer()`

When you use `lmer()` with multiple grouping factors specified with `(1|grouping_factor)`, `lmer` interprets these as independent sources of random variability associated with the levels of each `grouping_factor`.

- **Automatically detecting factor levels and assigning unique random intercepts/slopes:** Yes, `lmer` automatically identifies the unique levels within the factor column specified after the `|`. For a term like `(1|FactorA)`, it will assign a unique random intercept deviation to each distinct level found in the `FactorA` column of your data. Similarly, for `(slope_var|FactorA)`, it would assign a unique random intercept *and* a unique random slope for `slope_var` to each level of `FactorA`. When factors are crossed, like `(1|A) + (1|B)`, `lmer` recognizes the levels of `A` and the levels of `B` independently.
- **Distinguishing between fixed effects vs. random effects:** This distinction is made *entirely* by how you specify the terms in the `lmer` formula.
  - Terms *outside* the parentheses `(…)` are treated as **fixed effects**. `lmer` estimates a single coefficient (or set of coefficients for factors) that represent the average effect across the entire population (or the specific levels included if the factor is treated as fixed).
  - Terms *inside* the parentheses `(…)` are treated as **random effects**. `lmer` assumes the levels of the grouping factor associated with this term are a random sample from a larger population of possible levels. It estimates the *variance* of these random effects in that population, rather than estimating a specific coefficient for each level. The individual level deviations are then "shrunk" towards the mean (usually 0) based on the estimated variance and the data for that level.
- **Handling the *crossed* nature of Factor A and Factor B:** When you specify `(1|A) + (1|B)`, `lmer` assumes that any level of `A` can potentially appear with any level of `B` in your data. The term `(1|A)` contributes a random deviation specific to the level of A for an observation, and the term `(1|B)` contributes a random deviation specific to the level of B for the *same* observation. These two random deviations are assumed to be *independent*. `lmer` does *not* assume that all possible A-B combinations must exist in the data; it can handle unbalanced designs where some combinations are missing or have unequal numbers of observations. The "crossed" structure is captured by the *simultaneous inclusion* of independent random effects for A and B in the model equation for a given observation $i$ at level $a$ of A and level $b$ of B: $y_i = \mu + \text{fixed effects}_i + u_a + v_b + \epsilon_i$, where $u_a \sim N(0, \sigma_A^2)$ and $v_b \sim N(0, \sigma_B^2)$ are the random effects for level $a$ of A and level $b$ of B, respectively, and $u_a$ and $v_b$ are independent.

## Data Simulation for Crossed Designs

To simulate data for a crossed random effects ANOVA with two random factors (A and B) and their interaction, you need to create a data structure where levels of A are combined with levels of B.

- **Minimal code structure for crossed:** The most straightforward way to ensure a *balanced, fully crossed* design is to use `expand.grid()`. This function creates a data frame from all combinations of the supplied vectors or factors. 
- **How to avoid conflating crossed vs. nested:** The structure created by `expand.grid()` is inherently crossed. In a nested design (e.g., B nested within A), the levels of B would be unique within each level of A (e.g., B1_A1, B2_A1, B3_A2, B4_A2). Your factor levels or the data frame structure itself would reflect this. By creating `FactorA` and `FactorB` as independent sets of levels and combining them with `expand.grid`, you ensure that level 1 of FactorA is associated with level 1, 2, 3, etc., of FactorB, and similarly for all other levels of FactorA. 
- **Parameters controlling variance components:** When simulating, you define the standard deviation (or variance) for the random effects associated with each factor and their interaction.
  - **Factor A variance ($\sigma_A^2$):** You'll generate random deviates for each level of A, drawn from $N(0, \sigma_A^2)$.
  - **Factor B variance ($\sigma_B^2$):** You'll generate random deviates for each level of B, drawn from $N(0, \sigma_B^2)$.
  - **Interaction A×B variance ($\sigma_{AB}^2$):** You'll generate random deviates for *each unique combination* of A and B levels (each cell), drawn from $N(0, \sigma_{AB}^2)$.
  - **Residual variance ($\sigma_e^2$):** You'll generate random deviates for each observation, drawn from $N(0, \sigma_e^2)$.

## Random Interaction Generation

If both Factor A and Factor B are random, their interaction (A×B) represents variability specific to the unique pairing of a level from A and a level from B, *beyond* what the main effects of A and B explain.

- **Is the interaction effect modeled as a separate random term?** Yes, in a mixed-effects model, a random interaction term `(1|A:B)` is treated as a separate source of random variation. It estimates a variance component ($\sigma_{AB}^2$) associated with the specific combinations of levels from A and B. For each unique A-B pair (e.g., A=1, B=1; A=1, B=2; …; A=5, B=4), there is a random effect drawn from $N(0, \sigma_{AB}^2)$.
- **How do we ensure the interaction effects are orthogonal to the main effects?** In the *model specification* and *estimation* process by `lmer`, the main effects of A and B are estimated first (conceptually, though estimation is iterative), and the random interaction term `(1|A:B)` captures the remaining random variation associated with the cells after accounting for the main effects. The model structure itself, $y_{ijk} = \mu + u_i + v_j + (uv)_{ij} + e_{ijk}$, where $u_i$, $v_j$, and $(uv)_{ij}$ are independent random draws from $N(0, \sigma_A^2)$, $N(0, \sigma_B^2)$, and $N(0, \sigma_{AB}^2)$ respectively, ensures that their contributions to the variance are partitioned appropriately during estimation. You don't need to manually enforce orthogonality (like sum-to-zero constraints) on the simulated random draws *before* adding them; the estimation method handles this partitioning.
- **Distributional assumptions and constraints:**
  - **Distributional Assumptions:** The standard assumption for random effects (main effects and interactions) is that they are independently drawn from a Gaussian distribution with a mean of zero and a specific variance ($\sigma_A^2$, $\sigma_B^2$, $\sigma_{AB}^2$). Residual errors are also assumed to be independently drawn from $N(0, \sigma_e^2)$.
  - **Constraints:** Unlike fixed effects ANOVA where sum-to-zero constraints are often applied to make effects estimable and interpretable, for *random* effects, the assumption is simply that they are sampled from a distribution (usually Gaussian) with mean zero. The model estimates the *variance* of this distribution. `lmer` uses parameterizations that do not rely on sum-to-zero constraints on the individual random effect values but rather model the variance-covariance structure of the random terms.

## `lmer()` Formula Specification

For a crossed random effects model with random intercepts for Factor A, Factor B, and their interaction, the formula syntax is:

```r
y ~ (1|FactorA) + (1|FactorB) + (1|FactorA:FactorB)
```

- `(1|FactorA)`: Specifies random intercepts grouped by levels of `FactorA`. Estimates $\sigma_A^2$.
- `(1|FactorB)`: Specifies random intercepts grouped by levels of `FactorB`. Estimates $\sigma_B^2$.
- `(1|FactorA:FactorB)`: Specifies random intercepts grouped by the unique combinations (cells) of `FactorA` and `FactorB`. Estimates $\sigma_{AB}^2$. Note that `FactorA:FactorB` represents the interaction grouping factor.
- **Optional fixed effects:** You add fixed effects terms before the random effects.
  - To include a fixed intercept (usually included by default): `y ~ 1 + (1|A) + (1|B) + (1|A:B)` (though `1` is often implicit).
  - To include a fixed covariate `X` and a fixed factor `C`: `y ~ X + FactorC + (1|FactorA) + (1|FactorB) + (1|FactorA:FactorB)
`
**Important Note:** The term `(1|FactorA*FactorB)` is *not* equivalent to `(1|FactorA) + (1|FactorB) + (1|FactorA:FactorB)`. `(1|FactorA*FactorB)` creates *one* random effect term grouped by the interaction factor `FactorA:FactorB`. It estimates a single variance component for the A:B cells. It implies that the variance associated with Factor A and Factor B main effects are somehow implicitly captured within the A:B cell variance, which is usually not the desired interpretation for crossed main effects and interaction. Use `(1|A) + (1|B) + (1|A:B)` to model separate variance components for the A main effect, the B main effect, and the A-B interaction.

## Code Example Validation

Here is a reproducible R example simulating data for a balanced crossed design with random A, B, and A:B interaction, fitting the `lmer` model, and verifying the estimated variances.

```{r}
#| label: tbl-simulate-random-interaction
#| tbl-cap: "Simulate Random Interaction in Crossed Random Effects Models"

pacman::p_load(
    lme4,
    dplyr
)

# --- Simulation Parameters ---
set.seed(42) # for reproducibility

nA <- 20 # Number of levels for Factor A
nB <- 20 # Number of levels for Factor B
nAB <- nA * nB # Total number of A:B combinations
n_per_cell <- 15 # Observations per A:B combination

mu <- 10 # Overall mean (fixed intercept)

# True variance components (as specified in the request)
sigma_A_sq <- 2.0
sigma_B_sq <- 1.0
sigma_AB_sq <- 0.5
sigma_error_sq <- 1.0

# Standard deviations
sigma_A <- sqrt(sigma_A_sq)
sigma_B <- sqrt(sigma_B_sq)
sigma_AB <- sqrt(sigma_AB_sq)
sigma_error <- sqrt(sigma_error_sq)

# --- Data Generation ---

# 1. Create balanced design grid
design_grid <- expand.grid(
    FactorA = factor(1:nA),
    FactorB = factor(1:nB)
) |>
    mutate(
        # Create a unique identifier for each A:B combination
        AB_interaction = interaction(FactorA, FactorB, sep = ":"),
        A_id = as.numeric(FactorA),
        B_id = as.numeric(FactorB),
        AB_id = 1:n()
    )

# Repeat for observations per cell
sim_data <- design_grid[rep(1:nrow(design_grid), each = n_per_cell), ]

# Check the number of observations
cat("Total observations:", nrow(sim_data), "\n")
# Check the number of levels for each factor
cat("Levels of Factor A:", nlevels(sim_data$FactorA), "\n")
cat("Levels of Factor B:", nlevels(sim_data$FactorB), "\n")
cat("Levels of A:B interaction:", nlevels(sim_data$AB_interaction), "\n")

# 2. Generate random effects
# Random effects for Factor A levels
re_A <- rnorm(nA, mean = 0, sd = sigma_A)
# names(re_A) <- levels(sim_data$FactorA)

# Random effects for Factor B levels
re_B <- rnorm(nB, mean = 0, sd = sigma_B)
# names(re_B) <- levels(sim_data$FactorB)

# Random effects for A:B interaction cells
re_AB <- rnorm(nAB, mean = 0, sd = sigma_AB)
# names(re_AB) <- levels(sim_data$AB_interaction)

# 3. Generate residual error
error <- rnorm(nrow(sim_data), mean = 0, sd = sigma_error)

# 4. Calculate the simulated response variable 'y'
sim_data <- sim_data |>
    mutate(
        # Overall mean
        y = mu +
            # Add random effect for FactorA level of this row
            re_A[A_id] +
            # Add random effect for FactorB level of this row
            re_B[B_id] +  # Fixed the bracket here
            # Add random effect for A:B interaction cell of this row
            re_AB[AB_id] + # Updated to use AB_id for indexing
            # Add residual error
            error
    )
# Display the first few rows of the simulated data
head(sim_data)
```

```{r}
#| label: tbl-model-fitting
#| tbl-cap: "Model Fitting and Variance Component Extraction"
#| echo: false

# Fit the lmer model with crossed random effects for A, B, and A:B
# y ~ 1 + (1|FactorA) + (1|FactorB) + (1|FactorA:FactorB)
# The '1 +' for the intercept is implicit
model_crossed <- lmer(y ~ (1 | FactorA) + (1 | FactorB) + (1 | FactorA:FactorB), data = sim_data)

# --- Verification ---

# Print model summary
summary(model_crossed)

# Extract variance components
vc <- as.data.frame(VarCorr(model_crossed))

# Print estimated variances (square of StdDev)
cat("--- Estimated Variance Components ---\n")
print(vc |>
    select(grp, vcov) |>
    rename(Variance = vcov))

# Compare with true variances
cat("\n--- True Variance Components ---\n")
cat("FactorA:", sigma_A_sq, "\n")
cat("FactorB:", sigma_B_sq, "\n")
cat("FactorA:FactorB:", sigma_AB_sq, "\n")
cat("Residual:", sigma_error_sq, "\n")

# --- Common Error Explanation ---

cat("\n--- Common Error with Random Interaction Terms ---\n")
cat("If you encounter this error with your own data:\n")
cat("'number of levels of each grouping factor must be < number of observations'\n\n")

cat("This occurs when:\n")
cat("1. You have exactly one observation per cell (e.g., one observation per A:B combination)\n")
cat("2. The interaction term (A:B) has as many levels as there are observations\n")
cat("3. This creates a saturated model where the interaction perfectly predicts each observation\n\n")

cat("Solutions:\n")
cat("1. Ensure multiple observations per cell: n_per_cell > 1\n")
cat("2. Use an incomplete design where not every factor combination exists\n")
cat("3. If you truly have only one observation per cell, you cannot include the interaction term\n")
cat("   as random - consider using only the main random effects: (1|A) + (1|B)\n")
```


```{r}
#| label: tbl-error-demonstration
#| tbl-cap: "Demonstration of Error with One Observation per Cell"
#| echo: false
# For demonstration, let's create a problematic dataset with only one obs per cell
problematic_data <- design_grid # Just one obs per A:B cell
problematic_data$y <- rnorm(nrow(problematic_data))

cat("\nDemonstration of error - uncomment to run:\n")
cat("# This will produce the error because n_per_cell = 1:\n")
cat("# lmer(y ~ (1|FactorA) + (1|FactorB) + (1|FactorA:FactorB), data = problematic_data)\n")

# You can also calculate the sample variance of the generated random effects
# to see how they match the true variances before adding error
cat("\n--- Sample Variance of Generated Effects ---\n")
cat("FactorA (Sample Variance):", var(re_A), "\n")
cat("FactorB (Sample Variance):", var(re_B), "\n")
cat("FactorA:FactorB (Sample Variance):", var(re_AB), "\n")
```

When you run this code, you will see the `Estimated Variance Components` printed by `lmer` (extracted from `VarCorr()`) should be reasonably close to the `True Variance Components` defined at the start of the simulation. They won't match exactly due to the sampling variability in generating the random effects and the residual error, especially with a limited number of levels and observations. Increasing `nA`, `nB`, and `n_per_cell` would generally lead to estimates closer to the true values.

The explanation section shows why the common error occurs when attempting to model random interaction terms with too few observations per cell, which is a frequent issue in practice.